---
title: "Interrupted Time Series"
---

## The ITS idea

Interrupted time series (ITS) drops the comparison unit entirely. The counterfactual is built from the **treated unit's own pre-period dynamics**: fit a model on 1970–1988 California, extrapolate it into 1989–2000, and call the gap between the extrapolation and the observed data the effect. Where the naive pre-post estimate of chapter 1 assumes "no change", ITS allows a non-zero pre-trend. If California was already declining, the ITS counterfactual continues that decline; only the *extra* drop after 1989 gets attributed to the policy.

This chapter fits two ITS variants on the same Proposition 99 data:

1. A linear **growth-curve** model — the simplest pre-trend extrapolation possible.
2. An **AICc-selected ARIMA** model — a more flexible time-series alternative.

The two estimates disagree dramatically, and the disagreement is the lesson.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(fpp3)   # loads tsibble, fable, feasts

set.seed(42)
```

We reuse the cached Proposition 99 dataset and build a California-only `tsibble` indexed by year. A centred year index (`year0 = year - 1989`) is convenient for the post-period extrapolation.

```{r}
#| label: data-load
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()

# California-only time series with a Pre/Post factor and a centred year
# index (year0 = 0 at the first post-period year). The tsibble class is
# required by the fpp3 forecasting tools used later.
prop99_ts <- prop99 |>
  filter(state == "California") |>
  select(year, cigsale) |>
  mutate(prepost = factor(year > 1988, labels = c("Pre", "Post"))) |>
  as_tsibble(index = year) |>
  mutate(year0 = year - 1989)
```

## Linear growth-curve ITS

**The idea.** Fit a single straight line on California's pre-period cigarette sales, then extrapolate it forward as the counterfactual.

**The equation.** Fit a linear trend on the pre-period only,

$$Y_{1t} = \alpha + \beta\, t + \varepsilon_t, \qquad t \le t^* = 1988,$$

then extrapolate the fitted line into the post-period as the counterfactual,

$$\widehat{Y_{1t}(0)} = \hat\alpha + \hat\beta\, t, \qquad t > t^*,$$

and finally average the per-year gap between observed and counterfactual,

$$\widehat{\text{ATT}}_{\text{ITS-growth}} = \frac{1}{T_{\text{post}}} \sum_{t > t^*} \Big[Y_{1t} - (\hat\alpha + \hat\beta\, t)\Big].$$

In words: a single straight line, fit on 1970–1988 cigarette sales in California, becomes the counterfactual for 1989–2000. The policy effect is the *average* of the per-year residuals between what was actually observed and what the extrapolated line predicted. The slope $\hat\beta$ captures whatever secular trend California was already on; only deviations *from* that trend after 1989 are attributed to Proposition 99.

```{r}
#| label: fit-growth
# Fit a linear pre-period trend (cigsale on year, 1970-1988 only).
fit_growth <- lm(cigsale ~ year, data = prop99_ts |> filter(prepost == "Pre"))

summary(fit_growth)$coefficients
```

The pre-period linear trend is about $-1.78$ packs per capita per year ($p < 10^{-5}$, $R^2 \approx 0.73$) — so California was already declining about 1.8 packs per year before Proposition 99. To estimate the policy effect we extrapolate that line forward to 2000 and average the gap.

```{r}
#| label: its-growth-estimate
# Subset to the 1989-2000 rows and extrapolate the fitted line forward.
post_df <- prop99_ts |> filter(prepost == "Post")
pred_growth <- predict(fit_growth, newdata = as_tibble(post_df))

# ATT estimate = average per-year gap between observed and extrapolation.
its_growth_estimate <- mean(post_df$cigsale - pred_growth)
its_growth_estimate
```

```{r}
#| label: fig-its-growth
#| fig-cap: "ITS counterfactual from a linear pre-period growth curve (1970–1988), extrapolated to 2000."
#| fig-width: 8
#| fig-height: 5
its_growth_plot <- prop99_ts |>
  as_tibble() |>
  mutate(counterfactual = predict(fit_growth, newdata = prop99_ts))

ggplot(its_growth_plot, aes(x = year)) +
  geom_line(aes(y = cigsale, color = "Observed"), linewidth = 1.1) +
  geom_line(aes(y = counterfactual, color = "Counterfactual"),
            linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 1988.5, color = "#d97757",
             linetype = "dotted", linewidth = 0.7) +
  scale_color_manual(values = c("Observed" = "#d97757",
                                "Counterfactual" = "#6a9bcc")) +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL) +
  theme_minimal()
```

**Reading the output.** The ITS-growth-curve estimate is about $-28.3$ packs per capita per year. That is essentially identical to the naive pre-post $-27.0$ from chapter 1. Why? Because both methods only use within-California information. Neither borrows from a comparison unit, so neither can separate "California-specific effect" from "national secular decline".

The coincidence is suggestive but not reassuring. Both methods can be biased the same way if California's pre-trend was *understating* the speed of the secular decline.

**Common pitfall.** Assuming the linear pre-trend is the right *shape*. If the true secular decline is accelerating or saturating, a linear extrapolation either understates or overstates what would have happened — and the policy effect inherits the bias.

## ARIMA-based ITS

**The idea.** Replace the straight line with a flexible time-series model. Let the data decide the model's complexity through an information criterion (AICc). Forecast forward as the counterfactual.

**The equation.** A general ARIMA$(p, d, q)$ model writes the $d$-th differenced series as an autoregressive-moving-average process. Using the lag operator $L$ (so $L\, Y_t = Y_{t-1}$):

$$\Phi(L)\, (1 - L)^d\, Y_{1t} \, = \, \Theta(L)\, \varepsilon_t, \qquad \varepsilon_t \sim \mathcal{N}(0, \sigma^2),$$

where $\Phi(L) = 1 - \phi_1 L - \cdots - \phi_p L^p$ collects the $p$ autoregressive coefficients and $\Theta(L) = 1 + \theta_1 L + \cdots + \theta_q L^q$ collects the $q$ moving-average coefficients. The `fable::ARIMA(..., ic = "aicc")` call searches over $(p, d, q)$ and picks the combination that minimises the corrected Akaike Information Criterion on the pre-period.

Once the model is fit, the post-period counterfactual is the model's $h$-step forecast and the ATT is the average gap, just as in the growth-curve version:

$$\widehat{Y_{1t}(0)} = \hat Y_{1t \mid t^*}, \qquad \widehat{\text{ATT}}_{\text{ITS-ARIMA}} = \frac{1}{T_{\text{post}}} \sum_{t > t^*} \Big[Y_{1t} - \hat Y_{1t \mid t^*}\Big].$$

In words: same recipe as the growth-curve version — fit on pre-period, project forward, average the gap — but the "fit on pre-period" step now uses an autoregressive-integrated-moving-average model instead of a straight line.

**What ARIMA$(p, d, q)$ means in plain English.** $p$ is the number of past values the model uses (autoregression). $d$ is the number of times the series is differenced before fitting (to handle trends). $q$ is the number of past forecast errors used (moving average). Lower AICc = "better fit traded off against complexity".

```{r}
#| label: fit-arima
#| message: false
#| warning: false
# Fit an ARIMA model on the 1970-1988 California series. ic = "aicc"
# tells fable to search over (p, d, q) and pick the AICc minimiser.
fit_arima <- prop99_ts |>
  filter(prepost == "Pre") |>
  model(timeseries = ARIMA(cigsale, ic = "aicc"))

report(fit_arima)
```

For this series AICc typically selects `ARIMA(1, 2, 0)`: one autoregressive lag and *two* rounds of differencing. The double-differencing means the model is tracking the *acceleration* of California's late-1980s drop, not just its level or slope. We then forecast 12 years out and average the gap.

```{r}
#| label: its-arima-estimate
# Project the fitted ARIMA 12 years forward as the post-period counterfactual.
fcasts <- forecast(fit_arima, h = "12 years")

# ATT estimate = average per-year gap between observed and ARIMA forecast.
ce_arima <- post_df$cigsale - fcasts$.mean
mean(ce_arima)
```

```{r}
#| label: fig-its-arima
#| fig-cap: "ITS counterfactual from an AICc-selected ARIMA model fit on 1970–1988, forecast to 2000."
#| fig-width: 8
#| fig-height: 5
arima_cf <- tibble(year = post_df$year, counterfactual = fcasts$.mean)
plot_df  <- prop99_ts |>
  as_tibble() |>
  select(year, cigsale) |>
  left_join(arima_cf, by = "year")

ggplot(plot_df, aes(x = year)) +
  geom_line(aes(y = cigsale, color = "Observed"), linewidth = 1.1) +
  geom_line(aes(y = counterfactual, color = "ARIMA counterfactual"),
            linetype = "dashed", linewidth = 1, na.rm = TRUE) +
  geom_vline(xintercept = 1988.5, color = "#d97757",
             linetype = "dotted", linewidth = 0.7) +
  scale_color_manual(values = c("Observed" = "#d97757",
                                "ARIMA counterfactual" = "#6a9bcc")) +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL) +
  theme_minimal()
```

**Reading the output.** The ARIMA-based ITS estimate comes out around $+4.5$ packs. That is *positive* — it would imply Proposition 99 *increased* California's smoking. That is plainly the wrong answer.

The visual diagnostic shows why. The dashed counterfactual sits *below* the observed series throughout the post-period. The model extrapolates the late-1980s downward acceleration too aggressively. It predicts California should have hit roughly 50 packs by 2000 if the pre-period momentum had continued. Since California actually only hit 60 packs, the model concludes Proposition 99 "raised" smoking by about 5 packs relative to that doomsday counterfactual.

**The pitfall in one sentence.** AICc minimises *in-sample* fit, but in-sample fit can come from features — here, second-order momentum — that do not persist *out-of-sample*.

**Common pitfall.** Trusting an information-criterion-selected model on a short pre-period. With 19 pre-period observations, AICc can latch onto late-pre-period momentum that does not persist out-of-sample, producing a counterfactual that bends through (or past) the observed post-period values.

## What the two ITS estimates tell us

The growth-curve and ARIMA variants share every step of the recipe — fit on pre-period, project forward, average the gap — but disagree by more than 30 packs because they extrapolate different features of the same data. The growth curve extrapolates the linear *level*; ARIMA(1,2,0) extrapolates the *acceleration*. There is no purely-within-California way to decide which is right, because the choice is identified by an assumption about the missing $Y_{1t}(0)$ that the data itself cannot verify.

The lesson is not "ARIMA is bad". The lesson is that **single-model ITS is fragile**. Always pair an ITS estimate against a comparison-unit method (the DiD of chapter 4, the synthetic control of chapter 5, or the Bayesian structural TS of chapter 6) before drawing conclusions.

## Further reading

- @bernal2017interrupted — practitioner's tutorial on ITS regression for public-health interventions.
- @hyndman2021forecasting — the canonical reference for the `fpp3` ecosystem and AICc-selected ARIMA modelling.
