---
title: "Basic Differences-in-Differences"
---

## Learning objectives

1. Construct the 2×2 DiD grid (treated vs. one control, pre vs. post) and recover the ATT as the difference of differences. This is the simplest causal contrast in the book and the conceptual anchor every later DiD generalization builds on.
2. Estimate the 2×2 DiD as a single OLS interaction (`cigsale ~ state * prepost`) and read the ATT off the interaction coefficient. The regression form is what scales to covariates and time-varying treatments in later chapters.
3. Test the parallel-trends assumption formally by fitting `cigsale ~ state * year` on the pre-period and checking whether the slope-difference interaction is zero. A reader who only "eyeballs" parallel trends has no defence against a referee who fits the regression.
4. Compute HAC-robust standard errors with `sandwich::vcovHAC` and explain why cluster-robust inference collapses at $G = 2$. This is the practical inference fix every 2-unit DiD needs, and naming the collapse keeps the reader from defaulting to a broken `cluster = state` call.

## The DiD idea

Chapter 2 showed that any within-unit method — growth curve, ARIMA — is fragile because the counterfactual is identified only by an assumption the data cannot verify. Difference-in-Differences borrows strength from *outside* California by adding one control unit, replacing an extrapolation assumption with a comparability assumption.

DiD picks one control state — Nevada, for this chapter — and treats its pre-to-post change as the counterfactual change California would have experienced absent the policy. Subtract Nevada's change from California's change. Whatever is left over is "what the policy did". The canonical applied example is @cardkrueger1994minimum, who compared fast-food employment in New Jersey and Pennsylvania around New Jersey's 1992 minimum-wage hike.

The identifying assumption is **parallel trends**: California and Nevada would have moved on parallel paths without the policy. Differences in *levels* are fine; differences in *trends* are not. The estimand is a proper **Average Treatment effect on the Treated (ATT)** for California.

## The change-of-changes identity

The formal DiD identity is

$$\hat{\tau}_{\text{DiD}} = \big(\bar{Y}_{\text{CA, post}} - \bar{Y}_{\text{CA, pre}}\big) - \big(\bar{Y}_{\text{NV, post}} - \bar{Y}_{\text{NV, pre}}\big).$$

In the chapter 1 notation, this is the imputation $\widehat{Y_{1t}(0)} = \overline{Y}_{1,\text{pre}} + (\overline{Y}_{0,\text{post}} - \overline{Y}_{0,\text{pre}})$ — California's pre-period mean plus Nevada's pre-to-post change.

The matching **population regression** is

$$Y_{it} = \alpha + \beta_1 \, \mathrm{Post}_t + \beta_2 \, \mathrm{Treat}_i + \tau \, (\mathrm{Post}_t \times \mathrm{Treat}_i) + u_{it},$$

where $\mathrm{Treat}_i = 1$ for California and $0$ for Nevada, and $\mathrm{Post}_t = 1$ for $t \ge 1989$. The interaction coefficient $\tau$ is the DiD ATT — algebraically identical to the change-of-changes identity above, but in a form that gives us a standard error.

The four ingredients of the DiD calculation are easier to see as a 2×2 grid. Each cell holds a group mean; the two within-state changes are the row differences; the DiD estimate is the difference *of* those differences.

```{mermaid}
%%| label: fig-did-22-grid
%%| fig-cap: "The DiD 2×2 grid: California vs Nevada × Pre vs Post. The interaction term recovers the difference of differences."
flowchart TB
    subgraph "California"
        CA_pre["Pre (1984–88) mean<br/>= 99.0"] --> CA_d["Δ California =<br/>72.0 − 99.0 = −27.0"]
        CA_post["Post (1989–93) mean<br/>= 72.0"] --> CA_d
    end
    subgraph "Nevada (control)"
        NV_pre["Pre (1984–88) mean<br/>= 143.1"] --> NV_d["Δ Nevada =<br/>121.8 − 143.1 = −21.3"]
        NV_post["Post (1989–93) mean<br/>= 121.8"] --> NV_d
    end
    CA_d --> DD["DiD ATT =<br/>(−27.0) − (−21.3) = −5.7"]
    NV_d --> DD

    style CA_pre fill:#d97757,stroke:#cbd5e0,color:#fff
    style CA_post fill:#d97757,stroke:#cbd5e0,color:#fff
    style NV_pre fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style NV_post fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style DD fill:#00d4c8,stroke:#cbd5e0,color:#141413
```

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `sandwich` and `lmtest` supply the heteroskedasticity-and-autocorrelation-consistent (HAC) standard errors used in the regression below. Inference on a 2×2 panel is genuinely awkward — see the "Fit and HAC inference" section for an honest discussion of why HAC is a compromise rather than a textbook fix [@bertrand2004how]. The `R/table_helpers.R` helper provides `ms_pretty()`, the `modelsummary` wrapper that styles the fitted regression table.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the transparent ggplot theme."
library(tidyverse)
library(sandwich)
library(lmtest)
source("R/table_helpers.R")

# set.seed not strictly needed here — OLS and means are deterministic —
# but kept for consistency with the rest of the book.
set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** DiD trades the comparison-unit problem for a smaller, sharper subset of the panel. From the full 39-state × 31-year dataset we keep only California and Nevada and only the 1984–1993 window — five pre-period years (1984–1988) and five post-period years (1989–1993). Nevada is the hand-picked control, chosen as a geographically and demographically adjacent state. (Reno and Las Vegas both sit inside California's media market and share its retail-price corridor; Nevada was also one of the donor states with non-zero weight in Abadie, Diamond & Hainmueller's synthetic California — see chapter 4.) The `state` factor is releveled with Nevada as the reference so that the `stateCalifornia` and `stateCalifornia:prepostPost` coefficients land on California's main effect and *extra* change relative to Nevada — the DiD coefficient is the latter.

```{r}
#| label: data-load
#| code-summary: "Code: Load Prop 99 data and subset to California and Nevada in 1984-1993 with a Pre/Post factor."
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()

# Keep only California and Nevada in the 1984-1993 window and add the
# Pre/Post factor. Make Nevada the reference level so that the
# stateCalifornia interaction lands on California's extra change.
prop99_did <- prop99 |>
  filter(state %in% c("California", "Nevada"),
         year > 1983, year < 1994) |>
  mutate(prepost = factor(year > 1988, labels = c("Pre", "Post")),
         state   = factor(state, levels = c("Nevada", "California")))
```

The resulting `prop99_did` tibble has 20 rows (2 states × 10 years) and four columns: `state`, `year`, `cigsale`, and `prepost`. Those are the only inputs the DiD regression needs.

## Fit and HAC inference

**The model.** The two-way interacted OLS `cigsale ~ state * prepost` is the *regression form* of the 2×2 grid above — the arithmetic in the grid is literally what this regression computes. The four estimated coefficients map one-to-one onto the grid cells: the intercept is the Nevada–Pre cell, `stateCalifornia` shifts up or down to land on the California–Pre cell, `prepostPost` shifts to the Nevada–Post cell, and the interaction `stateCalifornia:prepostPost` is exactly the DiD ATT $\hat\tau$ — the *extra* California change beyond what happened to Nevada over the same window.

**Standard errors on a 2×2 panel are tricky.** Textbook OLS SEs assume i.i.d. errors, which fails here because within-state residuals are serially correlated over the ten-year window. The standard fix — clustering by state [@bertrand2004how] — is **degenerate with only two clusters**: the cluster-robust variance estimator's asymptotic theory requires the number of clusters $G$ to grow, and at $G = 2$ it collapses to a single 2×2 outer product. Software will compute a number; one cannot interpret it. We report HAC-robust SEs via `sandwich::vcovHAC` as a transparent compromise: they correct for within-unit autocorrelation by treating the stacked panel as a long time series, but they ignore the cross-state row boundary in the stacked design. The qualitative reading — that the DiD point estimate is statistically indistinguishable from zero — survives every reasonable choice; readers should treat the $p$-value as a rough heuristic, not a textbook test.

```{r}
#| label: tbl-fit-did
#| tbl-cap: "Basic DiD fit (HAC-robust SEs)."
#| code-summary: "Code: Fit the two-way interacted DiD regression and report HAC-robust standard errors."
# Two-way interacted regression: state main effect + Pre/Post main effect
# + (state x Pre/Post) interaction. The interaction is the DiD estimate.
fit_did <- lm(cigsale ~ state * prepost, data = prop99_did)

# HAC-robust standard errors for the four coefficients.
ms_pretty(list("Two-way interacted OLS, HAC SEs" = fit_did),
          vcov     = sandwich::vcovHAC,
          coef_map = c("(Intercept)"                  = "Nevada · Pre (baseline)",
                       "stateCalifornia"              = "California main",
                       "prepostPost"                  = "Post main",
                       "stateCalifornia:prepostPost"  = "DiD interaction"),
          notes    = "Standard errors HAC-robust via sandwich::vcovHAC.")
```

**Reading the output.** The interaction coefficient $\hat\tau$ (R name: `stateCalifornia:prepostPost`) is roughly $-5.68$ packs (HAC SE ≈ 5.39, $p \approx 0.31$). That is *dramatically* smaller than the naive $-27.02$ from chapter 1, and statistically indistinguishable from zero. Why? Because the `prepostPost` main effect is also large: $-21.34$ packs. Nevada's own cigarette sales fell by 21.3 packs between 1984–1988 and 1989–1993. When DiD subtracts that Nevada change from California's change, almost all of California's drop is absorbed.

## Visual diagnostic

**What to look for.** DiD rests entirely on the **parallel-trends** assumption: that California and Nevada would have moved in lockstep absent the policy. The data themselves cannot prove this counterfactual, but we can at least *inspect the pre-period* by plotting both states on the same axes over the full 1970–2000 window. Three features matter visually: (1) do the two pre-1989 lines move roughly in parallel? (2) does Nevada also bend downward after 1989, and if so, by how much? (3) is the post-1989 *gap* between California and Nevada much bigger than the pre-period gap? If Nevada is itself trending down post-1989, the DiD contrast collapses — the Nevada change subtracts away most of California's drop.

```{r}
#| label: fig-did-parallel-trends
#| fig-cap: "California vs Nevada, 1970–2000 cigarette sales per capita. Pre-policy years are to the left of the dashed line."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot California vs Nevada cigarette sales over 1970-2000 with a 1988.5 policy cutoff."
two_states <- prop99 |>
  filter(state %in% c("California", "Nevada"))

ggplot(two_states, aes(x = year, y = cigsale, color = state)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 1988.5, color = "#94a3b8",
             linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = c("California" = "#d97757",
                                "Nevada"     = "#6a9bcc")) +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL)
```

**Formal pre-trends test.** The visual is suggestive, but parallel trends is a claim about *slopes* in the pre-period, and we can test it directly. Fit `cigsale ~ state * year` on the 1984–1988 pre-window: if parallel trends holds, the `stateCalifornia:year` interaction — the *slope* difference between the two states' pre-period paths — should be statistically indistinguishable from zero.

```{r}
#| label: tbl-pretrend
#| tbl-cap: "Formal pre-trends test: California × year slope difference on the 1984–1988 pre-window (HAC SEs). Under parallel trends the slope difference should be ≈ 0."
#| code-summary: "Code: Test parallel pre-trends by regressing cigsale on state-by-year over 1984-1988."
pre_window <- prop99_did |> filter(prepost == "Pre")
fit_pre <- lm(cigsale ~ state * year, data = pre_window)

ms_pretty(list("Pre-trend slope test (1984-1988)" = fit_pre),
          vcov     = sandwich::vcovHAC,
          coef_map = c("(Intercept)"          = "Nevada baseline",
                       "stateCalifornia"      = "California level shift",
                       "year"                 = "Nevada slope",
                       "stateCalifornia:year" = "Slope difference (CA - NV)"),
          notes    = "If parallel trends held, the slope difference should be ~0.")
```

The slope-difference estimate is roughly $-4.63$ packs per year with HAC $p \approx 0.024$. Even **before** Proposition 99, California was declining roughly 4.6 packs/year *faster* than Nevada — i.e. the two series were already diverging in the direction the policy would later push, so the parallel-trends assumption is *rejected* by the pre-period data themselves. This is load-bearing: it turns the chapter's diagnosis from "Nevada has its own decline (visible in the plot)" into "California was already separating from Nevada (testable, rejected)".

This is the textbook DiD pitfall. A single control unit that itself is shifting in the same direction makes the contrast collapse. Nevada is geographically and culturally adjacent to California. It inherits many of the same secular forces: rising health awareness, federal tobacco settlements, retail-price spillovers. So it is a poor "what would California have done?" control.

**Common pitfall.** Picking the *one* "most similar" control by hand. If your single control is subject to the same secular forces as the treated unit — geographic neighbours, policy spillovers, regional macro shocks — the contrast collapses and DiD silently reports zero.

**A second pitfall.** Clustering standard errors *by state* on a two-state panel. The cluster-robust variance estimator [@bertrand2004how] requires the number of clusters to grow; at $G = 2$ it collapses to a single 2×2 outer product and is uninterpretable. Software will compute a number; do not trust it. With only two units, accept that classical inference on the DiD coefficient is unavailable and let the point estimate carry the argument.

**A third pitfall.** Treating Nevada as causally exogenous. Nevada borders California on a 600-km stretch, sits inside California's TV-advertising and retail-price corridor, and inherits its anti-smoking media spillovers. The "control" is not policy-untreated in the deep sense — only in the narrow sense of not having passed Proposition 99 itself.

**Recap.** DiD against Nevada says $-5.7$ packs and we cannot reject zero. The lesson is *not* that DiD is broken — it is that DiD with a single similar control unit is fragile. **Synthetic Control (chapter 4) is the principled response**: instead of one neighbour, blend *many Nevadas* into a *weighted* synthetic California — the optimisation picks the weights so the analyst no longer has to hand-pick the control.

## Key takeaways

- **Method:** Difference-in-Differences (DiD) is a 2×2 contrast that estimates the Average Treatment effect on the Treated (ATT) as $\hat\tau = (\bar Y_{\text{CA,post}} - \bar Y_{\text{CA,pre}}) - (\bar Y_{\text{NV,post}} - \bar Y_{\text{NV,pre}})$ — California's pre-to-post change minus Nevada's, with the control's change imputing California's missing $Y_{1t}(0)$.
- **Method:** The two-way interacted regression `cigsale ~ state * prepost` is the regression form of that 2×2 grid; with one treated unit and one control, the `state:prepost` interaction is algebraically identical to the change-of-changes ATT and is equivalent to the unit-and-period two-way fixed-effects specification on this balanced panel.
- **Method:** Identification rests on **parallel trends** — California and Nevada would have evolved on parallel paths absent Proposition 99. Differences in *levels* between the two states are absorbed by the unit fixed effect; only differences in *trends* bias the DiD estimand.
- **Lesson:** Adding even one outside comparison unit dramatically shrinks the estimated effect. The naive within-California drop of $-27$ packs collapses to roughly $-5.7$ packs once Nevada's own $-21.3$ pack decline is subtracted off, and the resulting ATT is statistically indistinguishable from zero.
- **Lesson:** Inspect parallel trends both visually *and* formally. Regressing `cigsale ~ state * year` on the 1984–1988 pre-window yields a slope difference of about $-4.63$ packs/year ($p \approx 0.024$): California was already declining faster than Nevada *before* the policy, so parallel trends is rejected by the pre-period data themselves.
- **Lesson:** A failed pre-trends test reframes the chapter's headline. The shrinkage from $-27$ to $-5.7$ is not just "Nevada has its own decline" but "California was already separating from Nevada in the policy's direction" — meaning DiD against Nevada absorbs genuine treatment signal into a non-parallel counterfactual.
- **Caveat:** Hand-picking a single "most similar" control is the textbook DiD pitfall. Geographic and cultural neighbours like Nevada inherit the same secular forces as the treated unit — rising health awareness, federal tobacco settlements, California-media spillovers — so the contrast can silently collapse to zero. Synthetic control (chapter 4) is the principled response: blend many donors into a weighted synthetic California rather than choose one neighbour by hand.
- **Caveat:** Inference on a 2×2 panel is genuinely awkward. Clustering by state is degenerate at $G = 2$ clusters; HAC-robust SEs via `sandwich::vcovHAC` are a transparent compromise that ignore the cross-state row boundary, so the reported $p$-value should be read as a rough heuristic rather than a textbook test.
- **Caveat:** This single-shock, single-treated-unit DiD is a deliberately simple base case. With many units adopting at staggered dates, two-way fixed-effects DiD can be biased by treatment-effect heterogeneity; chapter 8 revisits the staggered-adoption setting with the Callaway–Sant'Anna group-time ATT and Rambachan–Roth honest-DiD sensitivity bounds.

## Further reading

This chapter is the **canonical 2×2 / single-shock DiD**: one treated unit (California), one control unit (Nevada), one pre-period, one post-period. **Chapter 8 (Part II)** revisits DiD in the *staggered-adoption* setting — many treated units that switch at different dates — using the Callaway–Sant'Anna group-time ATT estimator [@callaway2021difference], the Sun–Abraham interaction-weighted estimator [@sun2021estimating], and the Rambachan–Roth honest-DiD sensitivity bounds [@rambachan2023more] on a Callaway-Sant'Anna minimum-wage county panel. The two chapters together cover the modern DiD toolkit.

**Classical references.** @cardkrueger1994minimum is the textbook applied 2×2 DiD (New Jersey vs Pennsylvania minimum wage and fast-food employment). @bertrand2004how is the seminal warning that DiD standard errors are biased toward zero under residual autocorrelation — the reason this chapter's inference choice is non-trivial. @bernal2017interrupted is the practitioner reference for parallel-trends *visual* diagnostics. For modern multi-period DiD with staggered adoption, `did` (Callaway–Sant'Anna) and `fixest::feols` (two-way fixed effects with cluster-robust SEs — but **only when there are enough clusters**) are the standard R workhorses.
