---
title: "Basic Differences-in-Differences"
---

## The DiD idea

Difference-in-Differences picks one control state — Nevada, for this chapter — and treats its pre-to-post change as the counterfactual change California would have experienced absent the policy. Subtract Nevada's change from California's change. Whatever is left over is "what the policy did".

The identifying assumption is **parallel trends**: California and Nevada would have moved on parallel paths without the policy. Differences in *levels* are fine; differences in *trends* are not. The estimand is a proper **Average Treatment effect on the Treated (ATT)** for California.

## The change-of-changes identity

The formal DiD identity is

$$\hat{\tau}_{\text{DiD}} = \big(\bar{Y}_{\text{CA, post}} - \bar{Y}_{\text{CA, pre}}\big) - \big(\bar{Y}_{\text{NV, post}} - \bar{Y}_{\text{NV, pre}}\big).$$

The four ingredients of the DiD calculation are easier to see as a 2×2 grid. Each cell holds a group mean; the two within-state changes are the row differences; the DiD estimate is the difference *of* those differences.

```{mermaid}
%%| label: fig-did-22-grid
%%| fig-cap: "The DiD 2×2 grid: California vs Nevada × Pre vs Post. The interaction term recovers the difference of differences."
flowchart TB
    subgraph "California"
        CA_pre["Pre (1984–88) mean<br/>= 99.0"] --> CA_d["Δ California =<br/>72.0 − 99.0 = −27.0"]
        CA_post["Post (1989–93) mean<br/>= 72.0"] --> CA_d
    end
    subgraph "Nevada (control)"
        NV_pre["Pre (1984–88) mean<br/>= 143.1"] --> NV_d["Δ Nevada =<br/>121.8 − 143.1 = −21.3"]
        NV_post["Post (1989–93) mean<br/>= 121.8"] --> NV_d
    end
    CA_d --> DD["DiD ATT =<br/>(−27.0) − (−21.3) = −5.7"]
    NV_d --> DD

    style CA_pre fill:#d97757,stroke:#cbd5e0,color:#fff
    style CA_post fill:#d97757,stroke:#cbd5e0,color:#fff
    style NV_pre fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style NV_post fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style DD fill:#00d4c8,stroke:#cbd5e0,color:#141413
```

The arithmetic is literally what the regression below computes. In `cigsale ~ state * prepost`, the interaction coefficient `stateCalifornia:prepostPost` *is* the DiD estimate.

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `sandwich` and `lmtest` provide the HAC-robust standard errors used in the regression below: with only ten years on each unit, the residuals are heavily autocorrelated and the textbook OLS SEs would understate the uncertainty around the DiD interaction by a wide margin. The `R/table_helpers.R` helper provides `ms_pretty()`, the `modelsummary` wrapper that styles the fitted regression table.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(sandwich)
library(lmtest)
source("R/table_helpers.R")

set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** DiD trades the comparison-unit problem for a smaller, sharper subset of the panel. From the full 39-state × 31-year dataset we keep only California and Nevada and only the 1984–1993 window — five pre-period years (1984–1988) and five post-period years (1989–1993). Nevada is the hand-picked control, chosen as a geographically and demographically adjacent state. The `state` factor is releveled with Nevada as the reference so that the `stateCalifornia` and `stateCalifornia:prepostPost` coefficients land on California's main effect and *extra* change relative to Nevada — the DiD coefficient is the latter.

```{r}
#| label: data-load
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()

# Keep only California and Nevada in the 1984-1993 window and add the
# Pre/Post factor. Make Nevada the reference level so that the
# stateCalifornia interaction lands on California's extra change.
prop99_did <- prop99 |>
  filter(state %in% c("California", "Nevada"),
         year > 1983, year < 1994) |>
  mutate(prepost = factor(year > 1988, labels = c("Pre", "Post")),
         state   = factor(state, levels = c("Nevada", "California")))
```

The resulting `prop99_did` tibble has 20 rows (2 states × 10 years) and four columns: `state`, `year`, `cigsale`, and `prepost`. Those are the only inputs the DiD regression needs.

## Fit and HAC inference

**The model.** The two-way interacted OLS `cigsale ~ state * prepost` is the *regression form* of the 2×2 grid above. The four estimated coefficients map one-to-one onto the grid cells: the intercept is the Nevada–Pre cell, `stateCalifornia` shifts up or down to land on the California–Pre cell, `prepostPost` shifts to the Nevada–Post cell, and the interaction `stateCalifornia:prepostPost` is exactly the DiD ATT — the *extra* California change beyond what happened to Nevada over the same window. We pair the OLS coefficients with HAC-robust standard errors via `sandwich::vcovHAC` because we have ten years on each unit and the residuals are serially correlated.

```{r}
#| label: tbl-fit-did
#| tbl-cap: "Basic DiD fit (HAC-robust SEs)."
# Two-way interacted regression: state main effect + Pre/Post main effect
# + (state x Pre/Post) interaction. The interaction is the DiD estimate.
fit_did <- lm(cigsale ~ state * prepost, data = prop99_did)

# HAC-robust standard errors for the four coefficients.
ms_pretty(list("Two-way interacted OLS, HAC SEs" = fit_did),
          vcov     = sandwich::vcovHAC,
          coef_map = c("(Intercept)"                  = "Nevada · Pre (baseline)",
                       "stateCalifornia"              = "California main",
                       "prepostPost"                  = "Post main",
                       "stateCalifornia:prepostPost"  = "DiD interaction"),
          notes    = "Standard errors HAC-robust via sandwich::vcovHAC.")
```

**Reading the output.** The interaction coefficient `stateCalifornia:prepostPost` is roughly $-5.68$ packs (HAC SE ≈ 5.39, $p \approx 0.31$). That is *dramatically* smaller than the naive $-27.02$ from chapter 1, and statistically indistinguishable from zero. Why? Because the `prepostPost` main effect is also large: $-21.34$ packs. Nevada's own cigarette sales fell by 21.3 packs between 1984–1988 and 1989–1993. When DiD subtracts that Nevada change from California's change, almost all of California's drop is absorbed.

## Visual diagnostic

**What to look for.** DiD rests entirely on the **parallel-trends** assumption: that California and Nevada would have moved in lockstep absent the policy. The data themselves cannot prove this counterfactual, but we can at least *inspect the pre-period* by plotting both states on the same axes over the full 1970–2000 window. Three features matter visually: (1) do the two pre-1989 lines move roughly in parallel? (2) does Nevada also bend downward after 1989, and if so, by how much? (3) is the post-1989 *gap* between California and Nevada much bigger than the pre-period gap? If Nevada is itself trending down post-1989, the DiD contrast collapses — the Nevada change subtracts away most of California's drop.

```{r}
#| label: fig-did-parallel-trends
#| fig-cap: "California vs Nevada, 1970–2000. Nevada is also declining post-1988, so the DiD contrast collapses."
#| fig-width: 8
#| fig-height: 5
two_states <- prop99 |>
  filter(state %in% c("California", "Nevada"))

ggplot(two_states, aes(x = year, y = cigsale, color = state)) +
  geom_line(linewidth = 1.1) +
  geom_vline(xintercept = 1988.5, color = "#94a3b8",
             linetype = "dashed", linewidth = 0.7) +
  scale_color_manual(values = c("California" = "#d97757",
                                "Nevada"     = "#6a9bcc")) +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL) +
  theme_minimal()
```

This is the textbook DiD pitfall. A single control unit that itself is shifting in the same direction makes the contrast collapse. Nevada is geographically and culturally adjacent to California. It inherits many of the same secular forces: rising health awareness, federal tobacco settlements, retail-price spillovers. So it is a poor "what would California have done?" control.

**Common pitfall.** Picking the *one* "most similar" control by hand. If your single control is subject to the same secular forces as the treated unit — geographic neighbours, policy spillovers, regional macro shocks — the contrast collapses and DiD silently reports zero.

**Recap.** DiD against Nevada says $-5.7$ packs and we cannot reject zero. The lesson is *not* that DiD is broken — it is that DiD with a single similar control unit is fragile. **Synthetic Control (chapter 5) is the principled response**: instead of one control state, blend many states into a weighted "synthetic California".

## Further reading

For modern multi-period Difference-in-Differences with staggered adoption, the `did` package (Callaway & Sant'Anna group-time ATTs) and `fixest::feols` (two-way fixed effects with cluster-robust SEs) are the workhorses. See @bernal2017interrupted for parallel-trends diagnostics in policy evaluation.
