---
title: "Classical Synthetic Control"
---

## Learning objectives

1. Construct the four-stage `tidysynth` pipeline that optimizes donor weights to match the treated unit's pre-period predictor profile. This is the workhorse of the single-treated-unit synthetic-control literature and recurs in chapters 6 and 7.
2. Extract donor unit weights and the V-matrix predictor weights and inspect their concentration. Heavy concentration on one or two donors is the chapter's main fragility diagnostic and is what later prediction-interval methods try to fix.
3. Compute the post-period ATT as the mean gap between the treated unit and its synthetic counterpart. This single quantity is the headline estimate every subsequent SC variant in the book reports.
4. Conduct placebo permutation inference by refitting the pipeline on each donor and ranking California's MSPE ratio against the placebo distribution. Permutation tests are the proper SC uncertainty quantification — *p*-values from regression won't help here.
5. Run leave-one-out and in-time placebo robustness checks and report when the estimate survives them. Without these checks the headline ATT is one donor away from being an artifact.

## The SCM idea

Synthetic Control stops using one control state. Instead, it builds a *weighted combination* of donor states that matches the treated unit's pre-period as closely as possible on a chosen set of predictors. The weighted combination is "synthetic California". The gap between observed California and synthetic California is the estimated effect.

**Why it works where DiD failed.** Difference-in-Differences against Nevada (chapter 3) needed parallel pre-trends with *one* neighbour. Synthetic Control needs parallel pre-trends with a *data-driven blend* of many neighbours. The optimisation does the matching, so the analyst no longer has to pick "the right" control state by hand.

## The four-stage tidysynth pipeline

```{mermaid}
%%| label: fig-tidysynth-pipeline
%%| fig-cap: "The four-stage tidysynth pipeline: declare → predictors → weights → control, with a fifth stage of plot/grab helpers."
flowchart LR
    A["Stage 1<br/>synthetic_control()<br/>declare treated unit<br/>and intervention time"] --> B["Stage 2<br/>generate_predictor()<br/>define matching variables<br/>(one call per time window)"]
    B --> C["Stage 3<br/>generate_weights()<br/>optimise donor weights<br/>(quadratic programming)"]
    C --> D["Stage 4<br/>generate_control()<br/>build synthetic California<br/>and post-period gap series"]
    D --> E["Stage 5<br/>plot_/grab_ helpers<br/>trends, weights,<br/>placebos, MSPE ratio,<br/>Fisher exact p-value"]

    style A fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style B fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style C fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style D fill:#d97757,stroke:#cbd5e0,color:#fff
    style E fill:#00d4c8,stroke:#cbd5e0,color:#141413
```

Stages 1–4 produce the estimate. Stage 5 is a battery of inspection helpers — `plot_trends()`, `plot_differences()`, `plot_weights()`, `plot_placebos()`, `plot_mspe_ratio()`, `grab_unit_weights()`, `grab_predictor_weights()`, `grab_balance_table()`, `grab_significance()` — that turn the fitted object into figures and tables. We use all of them below.

## The equation

Let $X_1$ be the vector of $k$ pre-period predictors for the treated unit (California), and let $X_0$ be the $k \times J$ matrix holding the same predictors for the $J = 38$ donor states. The Synthetic Control estimator chooses donor weights $w$ to minimise the (V-weighted) discrepancy between treated and synthetic on the predictors:

$$w^* \, = \, \arg\min_{w \in \mathcal{W}} \, \big(X_1 - X_0 w\big)^\top V \big(X_1 - X_0 w\big),$$

subject to

$$\mathcal{W} = \big\{w \in \mathbb{R}^J \,:\, w_j \ge 0 \,\, \forall j, \,\, \textstyle\sum_{j=1}^J w_j = 1\big\}.$$

The diagonal matrix $V$ holds the **predictor** importance weights — the optimiser can care more about pre-period cigarette sales than about, say, beer consumption (we inspect $V$ below). Once $w^*$ is solved, the synthetic California outcome at any year $t$ is

$$\widehat{Y_{1t}(0)} = \sum_{j=1}^J w_j^* \, Y_{jt},$$

and the ATT over 1989–2000 is the mean post-period gap between observed California and that synthetic counterfactual.

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `tidysynth` is the workhorse for this chapter: it wraps the classical Abadie-Diamond-Hainmueller synthetic-control optimiser behind a tidy pipeline of `synthetic_control() |> generate_predictor() |> generate_weights() |> generate_control()` and ships a battery of `plot_*()` / `grab_*()` helpers for inspection. The `R/table_helpers.R` helper provides `gt_pretty()` for the donor-weight and balance tables.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the ggplot theme."
library(tidyverse)
library(tidysynth)
source("R/table_helpers.R")

# The tidysynth pipeline used in this chapter is deterministic (the IPOP
# quadratic-program solver and the placebo refits do not consume RNG), so
# this seed is set for consistency with other chapters rather than because
# any chunk below depends on it.
set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** Unlike the ITS and DiD chapters — which used only California or only California-plus-Nevada — Synthetic Control uses the **full 39-state × 31-year panel**. The donor pool *is* the point: the other 38 states are the raw material the optimiser will blend into "synthetic California". We don't pre-filter or restrict to a window here; the chapter-2 outcome `cigsale` and the covariates `lnincome`, `retprice`, `age15to24`, `beer` are all in `prop99` and are passed straight into `synthetic_control()` below.

```{r}
#| label: data-load
#| code-summary: "Code: Load the Proposition 99 panel from the RDS file."
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()
```

The loaded `prop99` is a 1,209-row × 7-column tibble (39 states × 31 years per row, columns `state`, `year`, `cigsale`, plus the four covariates). The pipeline in the next section consumes it as-is.

## Fit the synthetic-control pipeline

**The pipeline.** The chunk below runs the four core `tidysynth` stages back-to-back: stage 1 declares California as the treated unit and 1988 as the last pre-period year (with `generate_placebos = TRUE` to also refit the model treating *each donor state* as if it had been treated — those placebo fits are what powers the permutation test later); stage 2 declares the matching predictors (three pre-period covariate averages, one shorter-window covariate, and three *lagged outcomes* — cigarette sales themselves at 1975, 1980, 1988); stage 3 solves the constrained quadratic program for donor weights; stage 4 multiplies those weights through the donor outcomes to construct the synthetic-California series. The three `*_ipop` arguments are tuning knobs for the interior-point optimiser and only matter if convergence is slow.

```{r}
#| label: fit-syn
#| message: false
#| warning: false
#| code-summary: "Code: Run the four-stage tidysynth pipeline to build synthetic California."
prop99_syn <- prop99 |>
  # 1. Declare the panel structure: outcome, unit, time, treated unit
  #    ("California"), and the last full pre-period year (1988).
  #    generate_placebos = TRUE also fits the model treating each donor
  #    state as treated, for the permutation test below.
  synthetic_control(
    outcome  = cigsale, unit = state, time = year,
    i_unit   = "California", i_time = 1988,
    generate_placebos = TRUE
  ) |>
  # 2. Predictors averaged over the full pre-period (1980-1988).
  generate_predictor(
    time_window = 1980:1988,
    lnincome    = mean(lnincome, na.rm = TRUE),
    retprice    = mean(retprice, na.rm = TRUE),
    age15to24   = mean(age15to24, na.rm = TRUE)
  ) |>
  # 2b. beer is sparser, so use a narrower window where data is densest.
  generate_predictor(time_window = 1984:1988,
                     beer = mean(beer, na.rm = TRUE)) |>
  # 2c. Three "lagged outcomes" - cigsale at three pre-period dates.
  #     These pin synthetic California's pre-period trajectory.
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = 1988, cigsale_1988 = cigsale) |>
  # 3. Solve the constrained QP for donor weights w*. The three IPOP
  #    parameters are tuning knobs for the interior-point optimiser.
  generate_weights(optimization_window = 1970:1988,
                   margin_ipop = .02,
                   sigf_ipop   = 7,
                   bound_ipop  = 6) |>
  # 4. Compute the synthetic California series from w* and donor outcomes.
  generate_control()
```

**Predictor choices.** Seven predictors are passed in. Three are pre-period covariate averages over the full pre-period (`lnincome`, `retprice`, `age15to24` over 1980–1988). One uses a narrower window where data is densest (`beer` over 1984–1988). Three are *lagged outcomes* — cigarette sales themselves at 1975, 1980, and 1988. The lagged outcomes are the most important trick: anchoring the synthetic control on the treated unit's own pre-period *outcome levels* at multiple time points forces the synthetic series to track California's pre-1988 trajectory closely.

## Donor weights and predictor weights

The optimisation produces two weight vectors that drive the entire fit. Both are extractable as tidy tables.

```{r}
#| label: tbl-donor-weights
#| tbl-cap: "Donor unit weights (top 8 states; only the top 5 carry meaningful weight)."
#| code-summary: "Code: Extract the top 8 donor unit weights into a gt table."
# We display 8 rows for full transparency, but only the top 5 (Utah,
# Nevada, Montana, Colorado, Connecticut) absorb essentially all the
# donor-weight mass - the bottom three rows are effectively zero.
grab_unit_weights(prop99_syn) |>
  arrange(desc(weight)) |>
  head(8) |>
  gt_pretty(decimals = 3) |>
  cols_label(unit = "Donor state", weight = "Weight")
```

Donor weights answer the question *which states mimic California*. The companion table — the **predictor (V-matrix) weights** — answers *which variables* the optimiser used to decide what "mimics" means. The two weight vectors are produced by the same `generate_weights()` call and together describe the entire fit.

```{r}
#| label: tbl-predictor-weights
#| tbl-cap: "Predictor (V-matrix) weights."
#| code-summary: "Code: Extract the V-matrix predictor weights into a gt table."
grab_predictor_weights(prop99_syn) |>
  arrange(desc(weight)) |>
  gt_pretty(decimals = 3) |>
  cols_label(variable = "Predictor", weight = "V-matrix weight")
```

Two things to notice.

1. **Five states absorb essentially 100% of the donor weight.** Utah, Nevada, Montana, Colorado, Connecticut. Every other state gets effectively zero. California is matched mostly to other Mountain-West states with similar age structure and cigarette price levels, plus Connecticut as a smoking-rate counterweight from the east.
2. **The two earliest cigsale levels dominate the V matrix.** `cigsale_1975` and `cigsale_1980` together get roughly 88% of the predictor weight. The four non-lagged covariates (`lnincome`, `retprice`, `age15to24`, `beer`) get less than 9% combined. The optimiser has effectively decided: "the best way to predict California's cigarette sales is using *other states' cigarette sales*."

For a one-line visual of both weight vectors, tidysynth ships a `plot_weights()` helper:

```{r}
#| label: fig-sc-weights
#| fig-cap: "Donor unit weights (left) and predictor V-matrix weights (right) from tidysynth."
#| fig-width: 9
#| fig-height: 5
#| code-summary: "Code: Plot donor and predictor weights via tidysynth helper."
plot_weights(prop99_syn)
```

### A closer look at the V matrix

The combined `plot_weights()` view is convenient, but the V matrix deserves a stand-alone chart because it answers a different question than the donor weights. Donor weights say *which states* mimic California; the V matrix says *which variables* the optimiser used to decide what "mimics" means.

```{r}
#| label: fig-sc-predictor-weights
#| fig-cap: "V matrix: cigsale_1975 and cigsale_1980 dominate; behavioural covariates get nearly zero weight."
#| fig-width: 8
#| fig-height: 4
#| code-summary: "Code: Build a custom bar chart of V-matrix predictor weights."
predw_df <- grab_predictor_weights(prop99_syn) |>
  mutate(variable = fct_reorder(variable, weight))

ggplot(predw_df, aes(x = weight, y = variable)) +
  geom_col(fill = "#6a9bcc") +
  geom_text(aes(label = sprintf("%.3f", weight)), hjust = -0.12) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "V-matrix weight (predictor importance)", y = NULL)
```

Two readings of the same picture, one practical and one cautionary.

- *Practical reading.* Two lagged outcomes (`cigsale_1975` and `cigsale_1980`) carry the bulk of the matching information. The optimiser has decided that California's pre-period cigarette sales — *at multiple time points* — are the best fingerprint to match.
- *Cautionary reading.* The V matrix is **not a causal ranking**. It tells you which variables were *useful for matching the treated unit's pre-period*, not which variables *cause* the outcome.

**Common pitfall.** Treating the V matrix as a list of causal drivers. It is a list of *good pre-period predictors for one specific unit*, not a structural model of smoking.

**More pitfalls — what else can go wrong.**

- *Interpolation outside the convex hull.* The simplex constraint forces $\widehat{Y_{1t}(0)}$ to be a *convex combination* of donor outcomes, so any predictor for which California sits outside the range spanned by the donor pool will be implicitly extrapolated. When that happens, the optimiser tends to pile weight onto two or three extreme donors just to span the predictor range — a sign the donor pool is not rich enough. (See @abadie2021using, p. 408.)
- *Extreme / concentrated weights.* A weight vector that lives on only two or three donors is fragile: it makes the synthetic counterfactual hostage to those donors' idiosyncratic shocks, and a single leave-one-out perturbation can move the ATT noticeably. The robustness checks below — in-time placebo and leave-one-out — are the standard way to surface this fragility.
- *Donor-pool contamination.* If a "control" state itself experienced a policy shock during the post-period — a cigarette-tax hike, a clean-indoor-air law — then including it biases the synthetic counterfactual *toward* the treated condition and the estimated ATT shrinks toward zero. The Abadie-Diamond-Hainmueller donor pool excludes states that adopted large tobacco-control measures during 1989–2000 precisely for this reason.

## The estimate

**The ATT.** With the pipeline fit, the headline causal estimate is the mean of the per-year gap between observed California and synthetic California, restricted to 1989–2000. `grab_synthetic_control()` extracts a tidy long table with one row per year and two columns — `real_y` (observed) and `synth_y` (synthetic counterfactual) — so the ATT is a one-line `mean(real_y - synth_y)` over the post-period rows.

```{r}
#| label: sc-att
#| code-summary: "Code: Compute the 1989-2000 ATT as the mean post-period gap."
# grab_synthetic_control() returns a tidy tibble with observed (real_y)
# and synthetic (synth_y) cigsale for every year. We restrict to the
# post-period and compute the per-year gap.
sc_post <- grab_synthetic_control(prop99_syn) |>
  filter(time_unit > 1988) |>
  mutate(dif = real_y - synth_y)

# Average the per-year gap to recover the ATT.
mean(sc_post$dif)
```

The Synthetic Control ATT is approximately **`r sprintf("%.2f", mean(sc_post$dif))` packs/capita** averaged over 1989–2000. This is the book's primary causal estimate and within rounding of the canonical @abadie2010synthetic result.

```{r}
#| label: fig-sc-trends
#| fig-cap: "Synthetic Control: California (observed) vs synthetic California (weighted donor combination)."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Plot observed vs synthetic California trends with plot_trends."
plot_trends(prop99_syn)
```

The pre-period fit is excellent — the synthetic and observed series are nearly indistinguishable through 1988. A substantial gap opens immediately after 1989, widening to roughly 30 packs by 2000.

## Predictor balance: did the matching work?

`grab_balance_table()` shows California, synthetic California, and the unweighted donor average side-by-side on every predictor.

```{r}
#| label: tbl-sc-balance
#| tbl-cap: "Predictor balance: California vs synthetic California vs donor average."
#| code-summary: "Code: Display the predictor balance table side by side."
grab_balance_table(prop99_syn) |>
  gt_pretty(decimals = 2)
```

On every variable, *synthetic California* is far closer to California than the unweighted donor average is. The most dramatic improvement is on the lagged outcomes: `cigsale_1988` is roughly 90 for California vs ≈91 for the synthetic — a near-perfect match — while the unweighted donor average is around 114. That gap of ~24 packs is exactly the bias the naive pre-post method silently absorbed.

## Visualising the post-period gap

`plot_trends()` showed *both* observed and synthetic California on one canvas. The companion helper `plot_differences()` plots just the *gap*: $Y_{1t} - \widehat{Y_{1t}(0)}$, year by year. This isolates the treatment-effect curve in its cleanest form.

```{r}
#| label: fig-sc-differences
#| fig-cap: "Per-year gap between observed California and synthetic California."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Plot the per-year treatment gap with plot_differences."
plot_differences(prop99_syn)
```

Read the line as the *effect of Proposition 99 on California in year $t$*. The pre-period values hover near zero (the matching worked), the line drops sharply after 1989, and it stays negative — steadily widening — throughout the post-period. The 1989–2000 mean of this series is exactly the $-18.85$ packs ATT reported above.

## Inference via placebo permutation

A "standard error" computed as cross-year SD divided by $\sqrt{N}$ is *not* a real sampling-distribution-based standard error. The proper Synthetic Control uncertainty quantification is a permutation test.

**The recipe.** Refit the synthetic-control model treating *each donor state* as if *it* had been the treated unit. Compute the post-period gap for each placebo. Compare California's gap trajectory to those placebo trajectories. If California's gap is extreme relative to the placebos, the policy probably did something.

```{r}
#| label: fig-sc-placebos
#| fig-cap: "Placebo permutation: California in orange, donor placebos in grey (badly pre-fit donors pruned by default)."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot the placebo permutation with default pruning."
plot_placebos(prop99_syn)
```

The orange line is California; the grey lines are the donor placebos. By default, `plot_placebos()` *prunes* placebos whose pre-period mean squared prediction error (MSPE) exceeds twice California's — those donors fit their own pre-period so badly that comparing their post-period gap to California's would be misleading. After pruning, California's post-period gap sits visibly below every retained placebo, which is the visual signature of a "real" treatment effect.

The unpruned variant keeps every donor for full transparency:

```{r}
#| label: fig-sc-placebos-unpruned
#| fig-cap: "Same placebo plot with every donor retained — including badly pre-fit ones — to show the full pool of placebo trajectories."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot the placebo permutation without pruning bad-fit donors."
plot_placebos(prop99_syn, prune = FALSE)
```

With pruning off, the grey cloud is messier and a few badly-fit donors swing wildly — but California's post-period descent still ends up at the bottom of the bundle. The qualitative conclusion does not depend on the pruning rule.

## MSPE ratio and Fisher exact p-value

A sharper version of the same test is the **MSPE ratio** — the ratio of post-period to pre-period mean squared prediction error. If a unit has a tight pre-period fit *and* a large post-period gap, the ratio is large.

```{r}
#| label: tbl-sc-significance
#| tbl-cap: "MSPE ratios — top 5 units."
#| code-summary: "Code: Tabulate the top 5 MSPE ratios and Fisher p-values."
# grab_significance() returns one row per unit (treated + every placebo)
# with pre_mspe, post_mspe, the post/pre ratio, the unit's rank in that
# ratio, and the Fisher-style p-value (rank / n_units).
grab_significance(prop99_syn) |>
  arrange(desc(mspe_ratio)) |>
  head(5) |>
  gt_pretty(decimals = 3)
```

California's MSPE ratio is around **124** — more than two and a half times higher than the next-highest unit. California ranks **1st out of 39 units**. The Fisher exact $p$-value is rank divided by total units, so $1/39 \approx 0.026$. Under the null hypothesis that Proposition 99 had no effect, the probability of seeing a unit this extreme purely by chance is about 2.6%.

The same ratio plotted as a bar chart makes the rank-1 gap visible at a glance.

```{r}
#| label: fig-sc-mspe-ratio
#| fig-cap: "MSPE-ratio bar chart with California highlighted at rank 1 out of 39 units."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Plot the MSPE-ratio bar chart across all units."
plot_mspe_ratio(prop99_syn)
```

The orange bar at the top is California; every blue bar below it is a placebo donor. The gap between California and the second-place state is enormous. That gap is the visual signature of "a real treatment effect that the donor pool does not naturally replicate".

## Robustness: in-time placebo and leave-one-out

The placebo permutation above is the *in-space* robustness check — refit the model treating each donor as if it had been treated, and compare gap trajectories. Two complementary checks round out the canonical ADH (2010) robustness toolkit. The **in-time placebo** asks whether the gap California opens after 1989 is unusual compared to gaps the same pipeline would produce around an arbitrary pre-period date. The **leave-one-out** check asks whether the ATT depends on any single donor — a fragility-detector for the concentrated-weights pitfall flagged above.

### In-time placebo: pretend 1980 was the intervention

Refit the entire pipeline pretending the intervention was 1980 (eight years before the real one), restricting the panel to pre-1989 data so the actual policy can never enter the fit. If the synthetic-control machinery is *not* picking up spurious gaps, the placebo "ATT" between 1981 and 1988 should be near zero.

```{r}
#| label: fit-syn-intime-placebo
#| message: false
#| warning: false
#| code-summary: "Code: Refit the pipeline with a 1980 placebo intervention date."
prop99_intime <- prop99 |>
  filter(year < 1989) |>
  synthetic_control(
    outcome  = cigsale, unit = state, time = year,
    i_unit   = "California", i_time = 1980,
    generate_placebos = FALSE
  ) |>
  generate_predictor(
    time_window = 1970:1980,
    lnincome    = mean(lnincome, na.rm = TRUE),
    retprice    = mean(retprice, na.rm = TRUE),
    age15to24   = mean(age15to24, na.rm = TRUE)
  ) |>
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_weights(optimization_window = 1970:1980,
                   margin_ipop = .02, sigf_ipop = 7, bound_ipop = 6) |>
  generate_control()

intime_att <- grab_synthetic_control(prop99_intime) |>
  filter(time_unit > 1980) |>
  summarise(att = mean(real_y - synth_y)) |>
  pull(att)

intime_att
```

The in-time placebo "ATT" over 1981–1988 is about `r round(intime_att, 2)` packs/capita — an order of magnitude smaller than the −18.85 we found at the real intervention date. The pre-1989 series does *not* produce spurious gaps of the size we see post-1989.

### Leave-one-out: drop the highest-weight donor

Utah carries the largest donor weight (≈0.34). If the ATT is driven by Utah-specific noise rather than a real policy effect, dropping Utah and refitting should move the estimate substantially. Refit the pipeline with Utah excluded from the donor pool.

```{r}
#| label: fit-syn-loo-utah
#| message: false
#| warning: false
#| code-summary: "Code: Refit the pipeline with Utah dropped for a leave-one-out check."
prop99_loo <- prop99 |>
  filter(state != "Utah") |>
  synthetic_control(
    outcome  = cigsale, unit = state, time = year,
    i_unit   = "California", i_time = 1988,
    generate_placebos = FALSE
  ) |>
  generate_predictor(
    time_window = 1980:1988,
    lnincome    = mean(lnincome, na.rm = TRUE),
    retprice    = mean(retprice, na.rm = TRUE),
    age15to24   = mean(age15to24, na.rm = TRUE)
  ) |>
  generate_predictor(time_window = 1984:1988,
                     beer = mean(beer, na.rm = TRUE)) |>
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = 1988, cigsale_1988 = cigsale) |>
  generate_weights(optimization_window = 1970:1988,
                   margin_ipop = .02, sigf_ipop = 7, bound_ipop = 6) |>
  generate_control()

loo_att <- grab_synthetic_control(prop99_loo) |>
  filter(time_unit > 1988) |>
  summarise(att = mean(real_y - synth_y)) |>
  pull(att)

loo_att
```

The leave-one-out ATT (Utah dropped) is approximately `r round(loo_att, 2)` packs/capita — close to the −18.85 baseline. The headline finding does not hinge on any single donor; the remaining four high-weight states (Nevada, Montana, Colorado, Connecticut) span enough of California's pre-period to recover essentially the same synthetic counterfactual on their own. A fuller robustness exercise would iterate this drop over each of the five high-weight donors, but the single-drop check is the diagnostic that matters most in practice — it rules out the "one donor is doing all the work" failure mode.

## Inspecting the nested tidysynth object

`prop99_syn` is not a plain data frame — it is a *nested tibble* with one row per unit (treated unit + every donor refit as a placebo) and list-columns that hold every intermediate output of the optimisation.

*This one chunk prints with R's default nested-tibble formatter on purpose — the `<tibble [N × M]>` glyphs in the list-columns are the pedagogical point that a styled table would hide.*

```{r}
#| label: sc-nested
#| code-summary: "Code: Print the nested tidysynth tibble to expose its list-columns."
prop99_syn
```

Each list-column can be flattened with `tidyr::unnest()` for custom downstream work, or pulled out with one of the `grab_*()` helpers used above.

```{r}
#| label: tbl-sc-unnest
#| tbl-cap: "Unnested `.outcome` (first 8 rows)."
#| code-summary: "Code: Unnest the .outcome list-column and preview California's series."
# Flatten .outcome into a wide table: one row per (unit, year).
# The actual cigarette-sales column is named after each unit, so we
# select metadata + California's series for a clean preview.
prop99_syn |>
  tidyr::unnest(cols = c(.outcome)) |>
  select(.id, .placebo, .type, time_unit, California) |>
  head(8) |>
  gt_pretty(decimals = 2)
```

For the tidy *observed-vs-synthetic* table (which is what most analyses want), the dedicated helper is more convenient:

```{r}
#| label: tbl-sc-unnest-tidy
#| tbl-cap: "`grab_synthetic_control()` — observed vs synthetic (first 8 rows)."
#| code-summary: "Code: Preview the tidy observed vs synthetic table from grab_synthetic_control."
grab_synthetic_control(prop99_syn) |>
  head(8) |>
  gt_pretty(decimals = 2)
```

This is the whole point of the nested-tibble design: every step of the optimisation is *introspectable from R*, with no need to dig into S4 slots or `attr()` blobs.

## Recap

| Question | Answer |
|---|---|
| What does Synthetic Control estimate? | The ATT on California, 1989–2000 |
| What is the point estimate? | **`r sprintf("%.2f", mean(sc_post$dif))` packs/capita per year** |
| What is "synthetic California"? | A convex combination of four Mountain-West states (Utah, Nevada, Montana, Colorado) plus Connecticut |
| What predictors did the matching? | Mostly two lagged outcomes — `cigsale_1975` and `cigsale_1980` |
| How is the matching quality? | Excellent — synthetic and observed California are near-identical through 1988 |
| What is the inference statistic? | Fisher exact $p \approx 0.026$ (California ranks 1st out of 39 on the MSPE ratio) |
| What is the design-time pitfall? | Don't read the V matrix as a list of causal drivers — it is a list of *good pre-period predictors* |

Synthetic Control is the book's headline causal estimate, and the placebo / MSPE-ratio diagnostics both confirm that California's post-1989 trajectory is unusual relative to what other states experienced in the same window. In chapter 5 we hand the same donor information to a Bayesian model and ask whether a *credible interval* (a direct probability statement about the effect) tells the same story. Chapter 6 then revisits the same donor pool with `scpi`, replacing the Fisher rank-based $p$-value with finite-sample *prediction intervals* that decompose forecast error into in-sample weight uncertainty and out-of-sample post-treatment shocks. Chapter 7 closes the synthetic-control arc by relaxing SUTVA — letting donor sales themselves respond to California's policy through a spatial autoregressive layer — so that the cross-border spillover into Nevada becomes a derived quantity rather than an assumption.

## Key takeaways

- **Method:** Classical Synthetic Control (Abadie-Diamond-Hainmueller) estimates the ATT on a single treated unit by constructing a "synthetic" counterfactual as a weighted average of untreated donor units; donor weights $w$ are restricted to the unit simplex ($w_j \ge 0$, $\sum_j w_j = 1$), which rules out extrapolation outside the convex hull of the donor pool.
- **Method:** The weights solve a constrained quadratic program that minimises the V-weighted discrepancy $(X_1 - X_0 w)^\top V (X_1 - X_0 w)$ between treated and donor predictors, where the diagonal V matrix encodes the relative importance of each predictor for pre-period matching, not its causal contribution to the outcome.
- **Method:** Inference is design-based rather than sampling-based — a Fisher-style permutation test refits the model treating each donor as if it had been treated, and the MSPE ratio (post-period mean squared prediction error divided by pre-period MSPE) ranks the treated unit against this placebo distribution to yield an exact p-value of rank divided by number of units.
- **Lesson:** The 1989-2000 ATT for California is approximately -18.85 packs/capita per year, and California ranks 1st out of 39 units on the MSPE ratio, giving a Fisher exact p-value of roughly 0.026 — the post-1989 gap is more extreme than any of the 38 placebo-donor refits.
- **Lesson:** Synthetic California is a sparse blend — five donors (Utah, Nevada, Montana, Colorado, Connecticut) absorb essentially all of the donor-weight mass, and the V matrix places roughly 88% of its weight on just two lagged outcomes (`cigsale_1975` and `cigsale_1980`); lagged outcomes — pre-period values of the outcome itself entered as predictors — pin the synthetic series to California's own pre-1988 trajectory and drive the near-perfect pre-period fit visible in the trends plot.
- **Lesson:** The two robustness checks both pass — the in-time placebo that pretends 1980 was the intervention produces an "ATT" an order of magnitude smaller than -18.85, and the leave-one-out check that drops Utah (the highest-weight donor) returns an ATT close to the baseline, ruling out the failure mode where a single donor drives the estimate.
- **Caveat:** The V matrix is not a list of causal drivers — it ranks variables by their usefulness for matching *this* treated unit's pre-period, so a predictor can receive near-zero V-weight (as `beer`, `lnincome`, `age15to24`, and `retprice` do here) even if it genuinely affects cigarette consumption.
- **Caveat:** Sparse, concentrated donor weights make the synthetic counterfactual hostage to a handful of states' idiosyncratic shocks, and the simplex constraint means any predictor on which California lies outside the donor convex hull is silently extrapolated; both pathologies are why placebo permutation and leave-one-out diagnostics are required reading rather than optional polish.
- **Caveat:** Donor-pool contamination is a separate threat — if a "control" state itself adopted a major tobacco-control measure during 1989-2000 its post-period outcome would already reflect a treatment-like effect and bias the synthetic counterfactual toward the treated condition, which is why the canonical ADH donor pool excludes states with their own large policy shocks in the post-window.

## Further reading

- @abadie2010synthetic — the original synthetic-control treatment of Proposition 99.
- @abadie2021using — methodological feasibility, data requirements, and assumptions in a JEL review.
- @dunford2024tidysynth — documentation for the `tidysynth` package used here.
