---
title: "Augmented Synthetic Control"
---

Chapter 4 fit classical synthetic control to California and got *lucky*: synthetic
California tracked the real series almost perfectly through 1988, so the simplex
weights had nothing left to explain. That near-perfect pre-period fit is what
licenses the post-period gap as a causal effect. But the simplex is a strait
jacket — weights must be non-negative and sum to one, so the synthetic unit can
only ever be an *interpolation* of the donors. When the treated unit sits
**outside the convex hull** of the donor pool, no convex combination reproduces
its pre-period, the fit is poor, and the estimate inherits that bias.

The **Augmented Synthetic Control Method (ASCM)** of @benmichael2021augmented
keeps the synthetic-control weights but adds an *outcome model* — a ridge
regression — that estimates the residual imbalance and subtracts it off, exactly
the way augmented inverse-propensity weighting de-biases a misspecified weighting
estimator. This chapter works through the three faces of the `augsynth` package
on a *purpose-built simulated panel* engineered to have the bad pre-treatment fit
that Proposition 99 never suffered.

## Learning objectives

1. Fit ASCM for a single treated unit and read the **ridge bias correction**
   $\widehat{Y_{1t}(0)} = \sum_j \widehat{w}_j Y_{jt} + \big(\widehat{m}_t(X_1) - \sum_j \widehat{w}_j \widehat{m}_t(X_j)\big)$,
   recovering $\widehat{\tau}_{\text{ASCM}}$ and contrasting it with the simplex
   estimator $\widehat{\tau}_{\text{SCM}}$ of chapter 4.
2. Diagnose **poor pre-treatment fit** through the $L_2$ imbalance and show that
   ridge augmentation lowers it — the chapter's central demonstration that the
   outcome model earns its keep precisely when the donor convex hull cannot reach
   the treated unit.
3. Extend to **staggered adoption** with `multisynth`, reporting the overall
   ATT, per-cohort effects, and a per-lead **event study**
   $\widehat{\tau}_{\text{ASCM}}(e)$ — the augmented-control analogue of the
   Callaway-Sant'Anna machinery in chapter 9.
4. Fit **several outcomes jointly** with `augsynth_multiout`, finding one set of
   donor weights that balances all outcomes' pre-periods, and compare the joint
   per-outcome ATTs with separate single-outcome fits.
5. Quantify uncertainty with **conformal**, **jackknife+**, and **wild-bootstrap**
   inference, and state when each applies.

## The augmentation idea

Classical synthetic control builds the counterfactual as a weighted average of
donors, $\widehat{Y_{1t}(0)} = \sum_j \widehat{w}_j\, Y_{jt}$, with the weights
$\widehat{w}$ chosen to match the treated unit's pre-period. When a good match
exists, the pre-period imbalance $X_1 - X_0\widehat{w}$ is essentially zero and we
trust the post-period gap. When no convex combination matches — because the
treated unit is an *extrapolation* of the donor cloud rather than an
interpolation — a stubborn imbalance remains and biases the estimate.

ASCM keeps $\widehat{w}$ but adds a correction. Fit an outcome model
$\widehat{m}_t(\cdot)$ — `augsynth` uses ridge regression — that predicts the
post-period outcome from the pre-period predictors, and form

$$
\widehat{Y_{1t}(0)} \;=\; \underbrace{\sum_j \widehat{w}_j\, Y_{jt}}_{\text{SCM term}}
\;+\; \underbrace{\Big(\widehat{m}_t(X_1) - \sum_j \widehat{w}_j\, \widehat{m}_t(X_j)\Big)}_{\text{bias correction}}.
$$

The correction is the outcome model's estimate of the gap that the SCM weights
*failed* to close. If balance is perfect ($X_1 = \sum_j \widehat{w}_j X_j$) the
correction vanishes and ASCM equals SCM; the worse the balance, the more work the
correction does. Setting the outcome model to nothing (`progfunc = "none"`)
recovers classical synthetic control exactly, which makes it the natural baseline
to measure the augmentation against.

```{mermaid}
%%| label: fig-ascm-pipeline
%%| fig-cap: "The augmented synthetic control pipeline. SCM supplies the donor weights; a ridge outcome model estimates the residual imbalance; their combination is the bias-corrected counterfactual from which the ATT is read."
flowchart LR
    A["Donor pool<br/>(untreated units)"] --> B["SCM weights w<br/>(simplex; may fit poorly)"]
    A --> C["Outcome model m<br/>(ridge on pre-periods)"]
    B --> E["Bias-corrected<br/>counterfactual Y1(0)"]
    C --> D["Bias correction<br/>m(X1) - sum w*m(Xj)"]
    D --> E
    E --> F["ATT = Y1 - Y1(0)"]
    style A fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style E fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style F fill:#6a9bcc,stroke:#cbd5e0,color:#fff
```

## The estimator

The ridge outcome model regresses pre-period outcomes on the donor outcomes with
an $L_2$ penalty $\lambda_{\text{ridge}}$ chosen by cross-validation (we subscript
it to keep it distinct from the factor loadings $\lambda_i$ the book uses
elsewhere); the penalty shrinks the correction toward zero, so ASCM never strays
far from the SCM fit unless the data demand it. Two knobs matter throughout:

- **`progfunc`** selects the outcome model: `"none"` (plain SCM — no correction),
  `"ridge"` (the default ASCM), or others such as `"en"` (elastic net). Ridge is
  the workhorse and the only one with a closed-form bias estimate.
- **`scm`** toggles whether SCM weights are used at all. With `scm = TRUE` the
  weights come from the synthetic-control quadratic program and the outcome model
  only de-biases; with `scm = FALSE` the donors enter with equal weight and the
  outcome model does everything.

Unlike the simplex, the ridge-augmented weights are *not* constrained to be
non-negative — the correction can place small negative weights on donors,
allowing the controlled extrapolation that a convex combination forbids. That is
the source of both ASCM's power and its risk.

**Identification.** ASCM recovers the ATT under one assumption: the
ridge-augmented model correctly imputes the treated unit's untreated path
$\widehat{Y_{1t}(0)}$ from the donors — the SCM weights plus the outcome-model
correction capture every systematic pre-period difference, leaving only noise.
This is weaker than the classical-SCM requirement that the treated unit lie
*inside* the donor convex hull, but it is not free: if the residual imbalance
reflects a confounder the ridge model cannot see, the correction is biased too.
Everything in this chapter is conditional on it.

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `augsynth` provides all
three estimators — `augsynth()` for a single treated unit, `multisynth()` for
staggered adoption, and `augsynth_multiout()` for several outcomes. `haven` reads
the simulated panel from its labelled Stata file. `R/table_helpers.R` supplies
`gt_pretty()`.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the ggplot theme."
library(tidyverse)
library(augsynth)
library(haven)
source("R/table_helpers.R")

# multisynth's wild-bootstrap standard errors consume RNG, so the seed is load
# bearing here (unlike the deterministic solvers in chapter 4).
set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** Proposition 99 was too kind to make the point, so this chapter uses
a *simulated* tobacco-control panel built by `R/simulate_tobacco_panel.R` and
shipped as the labelled Stata file `data/tobacco_sim.dta`. Forty fictional states
are observed annually from 1970 to 2000. Thirty-six are never-treated donors;
four — Atlantica, Borealis, Cascadia, and Deltora — enact an anti-smoking program
at **staggered** dates (1986, 1989, 1992, 1995). Three outcomes are recorded:
cigarette sales, smoking-related mortality, and tobacco tax revenue.

The data-generating process is rigged to defeat the simplex. Each outcome follows
a two-factor model; the donors' factor loadings sit inside the unit square, but
the four treated states' loadings are pushed *outside* the convex hull of the
donor loadings. No convex combination of donors can reproduce the treated factor
path, so classical synthetic control is guaranteed a poor pre-period fit — the
exact condition under which ridge augmentation is supposed to help.

```{r}
#| label: data-load
#| code-summary: "Code: Read the labelled Stata file and prepare the modelling columns."
tobacco_raw <- read_dta("data/tobacco_sim.dta")

tobacco <- tobacco_raw |>
  mutate(
    adoption_year = if_else(adoption_year >= 9999, Inf, as.numeric(adoption_year)),
    treated       = as.integer(treated),   # static: ever adopts
    trt           = as.integer(trt)         # time-varying: program active this year
  )
```

The labelled `.dta` carries a human-readable description for every column, which
survives the round-trip into R and which Stata users will see in their variable
browser.

```{r}
#| label: tbl-codebook
#| tbl-cap: "Codebook for the simulated tobacco panel, read from the variable labels stored in the Stata file."
#| code-summary: "Code: Pull the variable labels out of the labelled data frame."
tibble(
  Variable = names(labelled::var_label(tobacco_raw)),
  Label    = unlist(labelled::var_label(tobacco_raw))
) |>
  gt_pretty()
```

The fastest way to *see* the fit problem is to plot the raw series. The four
treated states (coloured) ride well above and swing more widely than the band of
donor states (grey) on every outcome — they are not in the donors' interior.

```{r}
#| label: fig-rawtrends
#| fig-cap: "Raw simulated series. Donor states (grey) form a band; the four treated states (coloured) sit outside it and swing more widely, the visual signature of treated units beyond the donor convex hull. Dashed lines mark each treated state's adoption year."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Facet the three outcomes with donors greyed and treated states highlighted."
outcome_labs <- c(cigsale = "Cigarette sales", mortality = "Mortality",
                  taxrev = "Tax revenue")

tobacco |>
  pivot_longer(c(cigsale, mortality, taxrev),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = factor(outcome, levels = names(outcome_labs),
                          labels = outcome_labs)) |>
  ggplot(aes(year, value, group = state)) +
  geom_line(data = ~filter(.x, treated == 0),
            color = "#94a3b8", alpha = 0.35, linewidth = 0.3) +
  geom_line(data = ~filter(.x, treated == 1),
            aes(color = state), linewidth = 0.9) +
  facet_wrap(~outcome, scales = "free_y") +
  scale_color_brewer(palette = "Set2") +
  labs(x = NULL, y = NULL, color = "Treated state") +
  theme(legend.position = "bottom")
```

## A single treated unit: bias-correcting the fit

Start with the simplest case the package handles, the one the user-facing
`augsynth()` dispatches to `single_augsynth`: **one** treated unit and **one**
outcome. Take Borealis (which adopts in 1989) as the treated state and the 36
never-treated donors as the pool, and model cigarette sales. The treatment
indicator passed to `augsynth()` is the **time-varying** `trt`, which is 1 for
Borealis from 1989 onward and 0 everywhere else — never the static `treated`
flag.

```{r}
#| label: fit-single
#| message: false
#| warning: false
#| code-summary: "Code: Subset to donors + Borealis and fit plain SCM vs ridge ASCM."
sub1 <- tobacco |>
  filter(treated == 0 | state == "Borealis")

syn_none <- augsynth(cigsale ~ trt, unit = state, time = year, t_int = 1989,
                     data = sub1, progfunc = "none",  scm = TRUE)

syn_ridge <- augsynth(cigsale ~ trt, unit = state, time = year, t_int = 1989,
                      data = sub1, progfunc = "ridge", scm = TRUE)
```

::: {.callout-note appearance="simple"}
**The most common `augsynth` mistake.** The formula's right-hand side must be the
*time-varying* 0/1 indicator that switches on at the intervention and stays on
(`trt`), not a static "ever-treated" dummy. With a static flag every post-period
looks treated for every unit and the estimator silently returns nonsense. The
package infers the intervention time from when the indicator flips, which is also
why staggered adoption (below) needs no extra bookkeeping.
:::

How badly does the simplex fit, and how much does ridge repair it? The
`scaled_l2_imbalance` reports the pre-period imbalance relative to the
uniform-weights benchmark, so $1 - \text{scaled}$ is the fraction of the gap the
weights closed.

```{r}
#| label: tbl-imbalance
#| tbl-cap: "Pre-treatment fit for Borealis. Classical SCM leaves a large scaled imbalance because Borealis is outside the donor convex hull; ridge augmentation lowers it, closing a larger share of the gap."
#| code-summary: "Code: Tabulate L2 imbalance for plain SCM vs ridge ASCM."
imbalance_tbl <- tibble(
  Method = c("Classical SCM (`progfunc = none`)", "Ridge ASCM (`progfunc = ridge`)"),
  `L2 imbalance`        = c(syn_none$l2_imbalance, syn_ridge$l2_imbalance),
  `Scaled L2`           = c(syn_none$scaled_l2_imbalance, syn_ridge$scaled_l2_imbalance),
  `% gap closed`        = 100 * (1 - c(syn_none$scaled_l2_imbalance,
                                       syn_ridge$scaled_l2_imbalance))
)

gt_pretty(imbalance_tbl, decimals = 2) |>
  gt::fmt_markdown(columns = "Method")
```

```{r}
#| label: single-summaries
#| message: false
#| warning: false
#| code-summary: "Code: Summarise both fits with conformal inference."
s_none  <- summary(syn_none)
s_ridge <- summary(syn_ridge)

scaled_none  <- syn_none$scaled_l2_imbalance
scaled_ridge <- syn_ridge$scaled_l2_imbalance
att_ridge    <- s_ridge$average_att$Estimate
p_ridge      <- s_ridge$average_att$p_val
bias_ridge   <- mean(s_ridge$bias_est)
```

The simplex leaves a scaled imbalance of
`r sprintf("%.2f", scaled_none)`; ridge brings it down to
`r sprintf("%.2f", scaled_ridge)` — the outcome model closes roughly
`r sprintf("%.0f", 100 * (scaled_none - scaled_ridge) / scaled_none)`% of the
imbalance the simplex could not. That repaired fit moves the estimate: the
average bias correction ASCM applies is
`r sprintf("%.2f", bias_ridge)` packs/capita per year.

```{r}
#| label: tbl-single-att
#| tbl-cap: "Single-unit ATT on Borealis cigarette sales: classical SCM vs ridge ASCM. The joint p-value is the conformal test of the null of no post-period effect; the average bias is the ridge correction (zero by construction for plain SCM)."
#| code-summary: "Code: Assemble the ATT comparison from the two conformal summaries."
tibble(
  Method = c("Classical SCM", "Ridge ASCM"),
  ATT          = c(s_none$average_att$Estimate, s_ridge$average_att$Estimate),
  `Joint p`    = c(s_none$average_att$p_val,    s_ridge$average_att$p_val),
  `Avg. bias`  = c(0, bias_ridge),
  `Scaled L2`  = c(scaled_none, scaled_ridge)
) |>
  gt_pretty(decimals = 3)
```

The level plot shows synthetic Borealis tracking the real series; the gap plot
shows the estimated treatment effect with its conformal band. Notice that plain
SCM's synthetic series misses the pre-period (the grey gap before 1989 is not
flat), which is exactly what the imbalance number warned about.

```{r}
#| label: fig-single-outcomes
#| fig-cap: "Observed Borealis cigarette sales (orange) vs the ridge-ASCM synthetic counterfactual (blue, dashed). The dotted line marks the 1989 adoption; the visible pre-1989 gap is the residual imbalance the simplex could not close."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Reconstruct the synthetic series (observed minus gap) and plot both."
gap_single <- as_tibble(s_ridge$att)

borealis_levels <- sub1 |>
  filter(state == "Borealis") |>
  select(year, observed = cigsale) |>
  left_join(gap_single, by = c("year" = "Time")) |>
  mutate(synthetic = observed - Estimate)

borealis_levels |>
  ggplot(aes(year)) +
  geom_vline(xintercept = 1988.5, color = "#d97757", linetype = "dotted", linewidth = 0.7) +
  geom_line(aes(y = observed,  color = "Observed Borealis"), linewidth = 1.1) +
  geom_line(aes(y = synthetic, color = "Synthetic (ridge ASCM)"),
            linetype = "dashed", linewidth = 1.1) +
  scale_color_manual(values = c("Observed Borealis"      = "#d97757",
                                "Synthetic (ridge ASCM)" = "#6a9bcc")) +
  labs(x = NULL, y = "Cigarette sales (packs per capita)", color = NULL)
```

```{r}
#| label: fig-single-gap
#| fig-cap: "Estimated effect on Borealis cigarette sales (observed minus synthetic) with its 95% conformal band. The pre-1989 gap is residual imbalance; the post-1989 gap is the ATT."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: Plot the gap and its conformal band in the book palette."
gap_single |>
  ggplot(aes(Time, Estimate)) +
  geom_hline(yintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_vline(xintercept = 1988.5, color = "#d97757", linetype = "dotted", linewidth = 0.7) +
  geom_ribbon(aes(ymin = lower_bound, ymax = upper_bound), fill = "#6a9bcc", alpha = 0.15) +
  geom_line(color = "#6a9bcc", linewidth = 1.1) +
  labs(x = NULL, y = "Effect on cigarette sales (packs per capita)")
```

Ridge picks its penalty $\lambda_{\text{ridge}}$ by cross-validation; the CV curve
shows the error-minimising choice the fit used.

```{r}
#| label: fig-single-cv
#| fig-cap: "Cross-validation curve for the ridge penalty. The selected penalty balances pre-period fit against the size of the bias correction."
#| fig-width: 8
#| fig-height: 4.5
#| warning: false
#| code-summary: "Code: Plot the ridge cross-validation curve."
plot(syn_ridge, plot_type = "cv")
```

Because the augmentation relaxes the non-negativity constraint, the ridge weights
can go slightly negative — the controlled extrapolation the simplex forbids.

```{r}
#| label: tbl-single-weights
#| tbl-cap: "The eight donors carrying the most weight (by magnitude) in the ridge fit. Unlike the simplex, ridge permits small negative weights, allowing extrapolation beyond the donor convex hull."
#| code-summary: "Code: Extract and rank the ridge donor weights."
tibble(
  Donor  = rownames(syn_ridge$weights),
  Weight = syn_ridge$weights[, 1]
) |>
  arrange(desc(abs(Weight))) |>
  head(8) |>
  gt_pretty(decimals = 3)
```

## Staggered adoption: many treated units

Real programs roll out at different times, and pooling several treated units lets
us estimate an *average* policy effect with dynamics. `multisynth()` fits a
separate (partially pooled) synthetic control for every treated unit, lines them
up in event time, and averages. The same `trt` indicator now switches on at four
different dates; donors stay zero throughout. The `n_leads` argument sets how many
post-adoption years to average over (here five, the window every cohort shares).

```{r}
#| label: fit-multi
#| message: false
#| warning: false
#| code-summary: "Code: Fit multisynth across all four staggered cohorts."
msyn   <- multisynth(cigsale ~ trt, unit = state, time = year,
                     data = tobacco, n_leads = 5)
ms_sum <- summary(msyn)
ms_att <- as_tibble(ms_sum$att)
```

The overall ATT averages the effect across the four treated states and the first
five post-adoption years. Its standard error comes from a wild bootstrap (the
reason the seed matters), and it is wide — pooling four imperfectly-fit units is
genuinely uncertain.

```{r}
#| label: tbl-multi-overall
#| tbl-cap: "Overall augmented-control ATT on cigarette sales, averaged across the four staggered cohorts and the first five post-adoption years, with a wild-bootstrap 95% interval."
#| code-summary: "Code: Pull the pooled (Average, all-time) ATT row."
ms_overall <- ms_att |> filter(Level == "Average", is.na(Time))

ms_overall |>
  transmute(ATT = Estimate, `Std. error` = Std.Error,
            `CI lower` = lower_bound, `CI upper` = upper_bound) |>
  gt_pretty(decimals = 2)
```

```{r}
#| label: multi-values
#| code-summary: "Code: Stash inline values for the prose."
ms_overall_att <- ms_overall$Estimate
```

The per-cohort estimates show how much the four states differ — later-adopting
states tend to show larger fitted effects here, a feature of the simulated DGP
rather than anything structural.

```{r}
#| label: tbl-multi-cohort
#| tbl-cap: "Per-cohort ATTs. Each treated state gets its own partially-pooled synthetic control; the estimates are averaged over each state's available post-adoption years."
#| code-summary: "Code: Per-state ATTs joined to their adoption year."
adopt_lookup <- tobacco |>
  filter(treated == 1) |>
  distinct(state, adoption_year)

ms_att |>
  filter(Level != "Average", is.na(Time)) |>
  transmute(State = Level, ATT = Estimate, `Std. error` = Std.Error) |>
  left_join(adopt_lookup, by = c("State" = "state")) |>
  transmute(State, `Adoption year` = adoption_year, ATT, `Std. error`) |>
  arrange(`Adoption year`) |>
  gt_pretty(decimals = 2)
```

Averaging in event time gives the **event study**: effects scatter around zero
before adoption (a rough placebo check on the fit) and accumulate afterward.

```{r}
#| label: fig-multi-event
#| fig-cap: "Event study from multisynth. Pre-adoption effects (light) hover near zero; post-adoption effects (cyan) grow as the program's cumulative impact builds. Bars are 95% wild-bootstrap intervals."
#| fig-width: 8
#| fig-height: 4.5
#| warning: false
#| code-summary: "Code: Plot the average effect by event time with 95% CIs."
ms_att |>
  filter(Level == "Average", !is.na(Time), Time >= -10) |>
  mutate(phase = if_else(Time < 0, "Pre", "Post")) |>
  ggplot(aes(Time, Estimate, color = phase)) +
  geom_hline(yintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_vline(xintercept = -0.5, color = "#94a3b8", linetype = "dashed") +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2) +
  scale_color_manual(values = c("Pre" = "#e2e8f0", "Post" = "#22d3ee")) +
  labs(x = "Event time (years from adoption)",
       y = "Effect on cigarette sales", color = NULL)
```

`multisynth` also draws each treated state against its own synthetic control. The
`nu` argument (which we left at its data-chosen default) controls **pooling**:
`nu = 0` fits every unit a fully separate synthetic control, `nu = 1` pools them
into one, and intermediate values trade the better individual fit of separate
controls against the lower variance of pooling. The exercises dial it directly.

```{r}
#| label: fig-multi-traj
#| fig-cap: "Each treated state (panel) against its partially-pooled synthetic control. Gaps that open after the dashed adoption line are the per-unit treatment effects."
#| fig-width: 8
#| fig-height: 5
#| warning: false
#| code-summary: "Code: multisynth's built-in per-unit trajectory plot."
plot(msyn)
```

## Multiple outcomes: a joint fit

The program plausibly moves several outcomes at once, and we may want *one*
synthetic Borealis that tracks all of them. `augsynth_multiout()` finds a single
set of donor weights that balances every outcome's pre-period simultaneously
(averaging the outcomes, `combine_method = "avg"`, by default), then reads an ATT
off each. The outcomes go on the left of the formula, joined by `+`.

```{r}
#| label: fit-multiout
#| message: false
#| warning: false
#| code-summary: "Code: Fit one ridge ASCM jointly across the three outcomes."
mo  <- augsynth_multiout(cigsale + mortality + taxrev ~ trt,
                         unit = state, time = year, t_int = 1989,
                         data = sub1, progfunc = "ridge", scm = TRUE)
smo <- summary(mo, inf = FALSE)
```

```{r}
#| label: tbl-multiout-att
#| tbl-cap: "Joint per-outcome ATTs for Borealis from a single multi-outcome ASCM fit. The program cuts cigarette sales, modestly lowers (lagged) mortality, and raises tax revenue — the signs the simulation built in."
#| code-summary: "Code: Per-outcome average effects from the joint fit."
mo_att <- as_tibble(smo$average_att)

mo_att |>
  transmute(Outcome, `Joint ATT` = Estimate) |>
  gt_pretty(decimals = 2)
```

```{r}
#| label: fig-multiout
#| fig-cap: "Estimated effects over time for each outcome from the joint multi-outcome fit. One donor-weight vector drives all three panels."
#| fig-width: 8
#| fig-height: 4.5
#| warning: false
#| code-summary: "Code: multi-outcome effect plot, faceted by outcome."
plot(mo)
```

Does jointness matter? Compare the joint ATTs with what we get by fitting each
outcome on its own. Cigarette sales reuses the single-unit ridge fit from above;
mortality and tax revenue get their own fits.

```{r}
#| label: fit-separate
#| message: false
#| warning: false
#| code-summary: "Code: Fit each outcome separately for comparison."
sep_mort <- augsynth(mortality ~ trt, unit = state, time = year, t_int = 1989,
                     data = sub1, progfunc = "ridge", scm = TRUE)
sep_tax  <- augsynth(taxrev ~ trt, unit = state, time = year, t_int = 1989,
                     data = sub1, progfunc = "ridge", scm = TRUE)

separate_att <- tibble(
  Outcome = c("cigsale", "mortality", "taxrev"),
  `Separate ATT` = c(att_ridge,
                     summary(sep_mort)$average_att$Estimate,
                     summary(sep_tax)$average_att$Estimate)
)
```

```{r}
#| label: tbl-multiout-vs-separate
#| tbl-cap: "Joint vs separate per-outcome ATTs. The joint fit constrains one weight vector to balance all three pre-periods, so its per-outcome estimates differ from independent fits — most visibly for the noisier mortality series, where joint estimation regularises an otherwise volatile fit."
#| code-summary: "Code: Lay joint ATTs beside the separate-fit ATTs."
mo_att |>
  filter(Outcome != "Average") |>
  transmute(Outcome, `Joint ATT` = Estimate) |>
  left_join(separate_att, by = "Outcome") |>
  gt_pretty(decimals = 2)
```

The estimates differ because the joint fit cannot chase every outcome's
idiosyncrasies — it must serve all three with one weight vector. That is a feature
when the separate fits are overfitting noise (as the mortality series, with its
small and lagged true effect, is prone to do) and a limitation when the outcomes
genuinely want different donors. Which regime you are in is a judgement call, not
a setting.

**Where this leaves us.** On a panel built to break the simplex, augmented
synthetic control repaired the pre-period fit that classical synthetic control
could not. Chapter 6 (Structural Bayesian Time Series) returns to the Proposition
99 data and the single-treated-unit question, but swaps the donor-weighting
machinery for a Bayesian state-space model — replacing the ridge correction and
conformal band with a full posterior over $\widehat{Y_{1t}(0)}$ and a credible
interval, a different language for the same uncertainty.

## Recap

| Question | Answer |
|---|---|
| What does ASCM add to SCM? | A ridge **outcome model** that estimates and subtracts the residual pre-period imbalance the simplex weights left behind |
| When does it help most? | When the treated unit is **outside the donor convex hull**, so SCM fits poorly — here scaled $L_2$ falls from `r sprintf("%.2f", scaled_none)` to `r sprintf("%.2f", scaled_ridge)` |
| Single-unit estimate (Borealis)? | $\widehat{\tau}_{\text{ASCM}} \approx$ **`r sprintf("%.1f", att_ridge)` packs/capita** (conformal joint $p \approx$ `r sprintf("%.3f", p_ridge)`) |
| Staggered pooled estimate? | Overall ATT $\approx$ **`r sprintf("%.1f", ms_overall_att)` packs/capita** across four cohorts (`multisynth`) |
| Several outcomes at once? | One weight vector, per-outcome ATTs (`augsynth_multiout`): sales down, mortality down, revenue up |
| What relaxes vs classical SCM? | Non-negativity — ridge weights may go slightly negative, permitting controlled extrapolation |

## Key takeaways

**Methods:**

- The augmented synthetic control method estimates $\widehat{Y_{1t}(0)} = \sum_j \widehat{w}_j Y_{jt} + \big(\widehat{m}_t(X_1) - \sum_j \widehat{w}_j \widehat{m}_t(X_j)\big)$, keeping the synthetic-control weights $\widehat{w}$ but adding a ridge outcome model $\widehat{m}$ that de-biases the residual pre-period imbalance; with `progfunc = "none"` the correction is zero and the estimator collapses to the classical SCM of chapter 4.
- The augmentation relaxes the simplex's non-negativity constraint, so the effective weights can be slightly negative — a *controlled extrapolation* governed by a cross-validated ridge penalty $\lambda_{\text{ridge}}$, which is what lets ASCM reach treated units outside the donor convex hull that no convex combination can match.
- `multisynth` extends the estimator to staggered adoption by fitting a partially-pooled synthetic control per treated unit (pooling fraction `nu`), aligning them in event time, and averaging — the synthetic-control counterpart to the Callaway-Sant'Anna group-time aggregation in chapter 9 — while `augsynth_multiout` finds a single weight vector that balances several outcomes' pre-periods jointly.

**Lessons:**

- On this purpose-built panel, classical SCM leaves a scaled pre-period imbalance of `r sprintf("%.2f", scaled_none)` because the treated states sit outside the donor convex hull; ridge augmentation cuts it to `r sprintf("%.2f", scaled_ridge)`, closing about `r sprintf("%.0f", 100 * (scaled_none - scaled_ridge) / scaled_none)`% of the gap the simplex could not — the concrete payoff of the outcome model.
- The single-unit ridge estimate for Borealis is $\widehat{\tau}_{\text{ASCM}} \approx$ `r sprintf("%.1f", att_ridge)` packs/capita (conformal joint $p \approx$ `r sprintf("%.3f", p_ridge)`), and the staggered pooled estimate across all four cohorts is $\approx$ `r sprintf("%.1f", ms_overall_att)` packs/capita; the multi-outcome fit recovers the simulation's built-in signs — sales down, mortality down (and lagged), revenue up.
- Inference is method-specific: **conformal** prediction bands for the single-unit fit (a joint $p$-value plus per-period intervals), a **wild bootstrap** for `multisynth` (hence the load-bearing `set.seed`), and **jackknife+** as a fast alternative — none of them the classical regression standard error.

**Caveats:**

- ASCM is only as honest as its outcome model: a flexible $\widehat{m}$ can manufacture a good-looking pre-period fit by extrapolating aggressively, so a *low* imbalance after augmentation is necessary but not sufficient — the ridge penalty must be cross-validated, not hand-tuned to taste.
- The pooled `multisynth` standard errors are wide here (the overall interval spans zero) because four imperfectly-fit units carry real uncertainty; a tight point estimate from a hard-fit panel should be read with the bootstrap interval attached, never alone.
- Joint multi-outcome estimation regularises noisy individual outcomes but biases them toward the consensus weight vector, so a joint ATT is not a drop-in replacement for the outcome-specific estimate — report both when they disagree, as mortality does here.

## Further reading

- @benmichael2021augmented — *original method* — the augmented synthetic control method, its ridge formulation, and the bias-correction theory this chapter implements.
- @benmichael2022staggered — *staggered extension* — partial pooling across treated units, the basis for `multisynth`.
- @benmichael2021augsynth — *R package* — documentation and vignettes for `augsynth`, `multisynth`, and `augsynth_multiout`.
- @abadie2010synthetic — *classical baseline* — the simplex-constrained estimator of chapter 4 that ASCM augments.
- @callaway2021difference — *staggered analogue* — the group-time ATT aggregation that `multisynth` mirrors for synthetic controls (chapter 9).
- @liu2024practical — *practical guide* — situates augmented and factor-based estimators among the broader family of counterfactual methods.

## Exercises

The exercises below probe the choices that distinguish ASCM from plain SCM: the
outcome model (`progfunc`), the pooling fraction (`nu`), the donor pool, the role
of the SCM weights (`scm`), and how several outcomes are combined. They all reuse
the chapter's `tobacco` panel and the `sub1` donors-plus-Borealis subset.

### Exercise 1: How much does the outcome model matter?

Refit the single-unit Borealis model with three outcome models —
`progfunc = "none"`, `"ridge"`, and `"en"` (elastic net) — keeping `scm = TRUE`.
Tabulate the scaled $L_2$ imbalance for each. Which outcome models close the most
pre-period imbalance?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-01
#| message: false
#| warning: false
progfuncs <- c("none", "ridge", "en")

ex1 <- map_dfr(progfuncs, function(pf) {
  fit <- augsynth(cigsale ~ trt, unit = state, time = year, t_int = 1989,
                  data = sub1, progfunc = pf, scm = TRUE)
  tibble(progfunc = pf, scaled_L2 = fit$scaled_l2_imbalance)
})

gt_pretty(ex1, decimals = 3)
```

Both augmented models (`ridge`, `en`) cut the scaled imbalance well below plain
SCM's. Once an outcome model is doing the de-biasing, the precise penalty family
matters less than the fact that *some* correction is applied — and the chapter
body shows how that repaired fit carries through to the ATT itself.
:::

### Exercise 2: The pooling dial in multisynth

Refit `multisynth` at `nu = 0` (fully separate per-unit controls), `nu = 0.5`,
and `nu = 1` (fully pooled). Report the overall ATT for each. How does pooling
move the estimate?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-02
#| message: false
#| warning: false
ex2 <- map_dfr(c(0, 0.5, 1), function(nu_val) {
  fit <- multisynth(cigsale ~ trt, unit = state, time = year,
                    data = tobacco, n_leads = 5, nu = nu_val)
  overall <- summary(fit)$att |>
    as_tibble() |>
    filter(Level == "Average", is.na(Time))
  tibble(nu = nu_val, overall_ATT = overall$Estimate,
         std_error = overall$Std.Error)
})

gt_pretty(ex2, decimals = 2)
```

Fully separate fits (`nu = 0`) chase each unit's pre-period and give the noisiest
average; full pooling (`nu = 1`) stabilises the estimate by borrowing across
units at the cost of some per-unit fit. The data-chosen default used in the
chapter sits between these poles — the standard bias-variance compromise.
:::

### Exercise 3: Shrink the donor pool

The augmentation is supposed to be robust when the donor pool is thin. Drop the
eight donors carrying the most ridge weight, refit both `progfunc = "none"` and
`"ridge"`, and compare how much each method's scaled imbalance deteriorates.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-03
#| message: false
#| warning: false
top8 <- tibble(donor = rownames(syn_ridge$weights),
               w = syn_ridge$weights[, 1]) |>
  arrange(desc(abs(w))) |>
  slice_head(n = 8) |>
  pull(donor)

sub1_thin <- sub1 |> filter(!state %in% top8)

ex3 <- map_dfr(c("none", "ridge"), function(pf) {
  fit <- augsynth(cigsale ~ trt, unit = state, time = year, t_int = 1989,
                  data = sub1_thin, progfunc = pf, scm = TRUE)
  tibble(progfunc = pf, scaled_L2 = fit$scaled_l2_imbalance)
})

gt_pretty(ex3, decimals = 3)
```

With its best donors removed the simplex has even less of a convex hull to work
with, so plain SCM's imbalance climbs; ridge augmentation absorbs more of the gap
through its outcome model and degrades more gracefully. This is the robustness
ASCM was designed for — though neither method is magic once the pool is gutted.
:::

### Exercise 4: Do the SCM weights still matter?

With ridge augmentation in play, are the synthetic-control weights even
necessary? Refit Borealis with `progfunc = "ridge"` under `scm = TRUE` (weights +
outcome model) and `scm = FALSE` (equal donor weights + outcome model only), and
compare the scaled imbalance and ATT.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-04
#| message: false
#| warning: false
ex4 <- map_dfr(c(TRUE, FALSE), function(use_scm) {
  fit <- augsynth(cigsale ~ trt, unit = state, time = year, t_int = 1989,
                  data = sub1, progfunc = "ridge", scm = use_scm)
  tibble(scm = use_scm,
         scaled_L2 = fit$scaled_l2_imbalance,
         ATT = summary(fit)$average_att$Estimate)
})

gt_pretty(ex4, decimals = 3)
```

The outcome model alone (`scm = FALSE`) does some de-biasing, but the
synthetic-control weights still pull their weight: keeping them (`scm = TRUE`)
gives the better pre-period fit. ASCM is genuinely the *combination* of weights
and outcome model, not the outcome model wearing synthetic-control clothes.
:::

### Exercise 5 (stretch): Averaging vs concatenating outcomes

`augsynth_multiout` can combine outcomes by averaging their pre-periods
(`combine_method = "avg"`) or by stacking them (`combine_method = "concat"`).
Refit the three-outcome Borealis model both ways and compare the per-outcome
ATTs. Does the combination rule change the story?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-05
#| message: false
#| warning: false
ex5 <- map_dfr(c("avg", "concat"), function(cm) {
  fit <- augsynth_multiout(cigsale + mortality + taxrev ~ trt,
                           unit = state, time = year, t_int = 1989,
                           data = sub1, progfunc = "ridge", scm = TRUE,
                           combine_method = cm)
  summary(fit, inf = FALSE)$average_att |>
    as_tibble() |>
    filter(Outcome != "Average") |>
    transmute(Outcome, combine_method = cm, ATT = Estimate)
})

ex5 |>
  pivot_wider(names_from = combine_method, values_from = ATT) |>
  gt_pretty(decimals = 2)
```

The two rules weight the outcomes' pre-periods differently when searching for the
common donor mix, so the per-outcome ATTs shift — but the qualitative story
(sales down, mortality down, revenue up) is stable across both. When the
combination rule flips a sign or a significance verdict, that is a signal the
joint fit is straining to serve outcomes that want different donors.
:::
