---
title: "Classical Synthetic Control"
---

## The SCM idea

Synthetic Control stops using one control state. Instead, it builds a *weighted combination* of donor states that matches the treated unit's pre-period as closely as possible on a chosen set of predictors. The weighted combination is "synthetic California". The gap between observed California and synthetic California is the estimated effect.

**Why it works where DiD failed.** Difference-in-Differences against Nevada (chapter 4) needed parallel pre-trends with *one* neighbour. Synthetic Control needs parallel pre-trends with a *data-driven blend* of many neighbours. The optimisation does the matching, so the analyst no longer has to pick "the right" control state by hand.

## The four-stage tidysynth pipeline

```{mermaid}
flowchart LR
    A["1. synthetic_control()<br/>declare treated unit<br/>and intervention time"] --> B["2. generate_predictor()<br/>define matching variables<br/>(one call per time window)"]
    B --> C["3. generate_weights()<br/>optimise donor weights<br/>(quadratic programming)"]
    C --> D["4. generate_control()<br/>build synthetic California<br/>and post-period gap series"]
    D --> E["5. plot_/grab_ helpers<br/>trends, weights,<br/>placebos, MSPE ratio,<br/>Fisher exact p-value"]

    style A fill:#6a9bcc,stroke:#141413,color:#fff
    style B fill:#6a9bcc,stroke:#141413,color:#fff
    style C fill:#6a9bcc,stroke:#141413,color:#fff
    style D fill:#d97757,stroke:#141413,color:#fff
    style E fill:#00d4c8,stroke:#141413,color:#141413
```

Stages 1–4 produce the estimate. Stage 5 is a battery of inspection helpers — `plot_trends()`, `plot_differences()`, `plot_weights()`, `plot_placebos()`, `plot_mspe_ratio()`, `grab_unit_weights()`, `grab_predictor_weights()`, `grab_balance_table()`, `grab_significance()` — that turn the fitted object into figures and tables. We use all of them below.

## The equation

Let $X_1$ be the vector of $k$ pre-period predictors for the treated unit (California), and let $X_0$ be the $k \times J$ matrix holding the same predictors for the $J = 38$ donor states. The Synthetic Control estimator chooses donor weights $w$ to minimise the (V-weighted) discrepancy between treated and synthetic on the predictors:

$$w^* \, = \, \arg\min_{w \in \mathcal{W}} \, \big(X_1 - X_0 w\big)^\top V \big(X_1 - X_0 w\big),$$

subject to

$$\mathcal{W} = \big\{w \in \mathbb{R}^J \,:\, w_j \ge 0 \,\, \forall j, \,\, \textstyle\sum_{j=1}^J w_j = 1\big\}.$$

The diagonal matrix $V$ holds the **predictor** importance weights — the optimiser can care more about pre-period cigarette sales than about, say, beer consumption (we inspect $V$ below). Once $w^*$ is solved, the synthetic California outcome at any year $t$ is

$$\widehat{Y_{1t}(0)} = \sum_{j=1}^J w_j^* \, Y_{jt},$$

and the ATT over 1989–2000 is the mean post-period gap between observed California and that synthetic counterfactual.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(tidysynth)

set.seed(42)
```

```{r}
#| label: data-load
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()
```

## Fit the synthetic-control pipeline

```{r}
#| label: fit-syn
#| message: false
#| warning: false
prop99_syn <- prop99 |>
  # 1. Declare the panel structure: outcome, unit, time, treated unit
  #    ("California"), and the last full pre-period year (1988).
  #    generate_placebos = TRUE also fits the model treating each donor
  #    state as treated, for the permutation test below.
  synthetic_control(
    outcome  = cigsale, unit = state, time = year,
    i_unit   = "California", i_time = 1988,
    generate_placebos = TRUE
  ) |>
  # 2. Predictors averaged over the full pre-period (1980-1988).
  generate_predictor(
    time_window = 1980:1988,
    lnincome    = mean(lnincome, na.rm = TRUE),
    retprice    = mean(retprice, na.rm = TRUE),
    age15to24   = mean(age15to24, na.rm = TRUE)
  ) |>
  # 2b. beer is sparser, so use a narrower window where data is densest.
  generate_predictor(time_window = 1984:1988,
                     beer = mean(beer, na.rm = TRUE)) |>
  # 2c. Three "lagged outcomes" - cigsale at three pre-period dates.
  #     These pin synthetic California's pre-period trajectory.
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = 1988, cigsale_1988 = cigsale) |>
  # 3. Solve the constrained QP for donor weights w*. The three IPOP
  #    parameters are tuning knobs for the interior-point optimiser.
  generate_weights(optimization_window = 1970:1988,
                   margin_ipop = .02,
                   sigf_ipop   = 7,
                   bound_ipop  = 6) |>
  # 4. Compute the synthetic California series from w* and donor outcomes.
  generate_control()
```

**Predictor choices.** Seven predictors are passed in. Three are pre-period covariate averages over the full pre-period (`lnincome`, `retprice`, `age15to24` over 1980–1988). One uses a narrower window where data is densest (`beer` over 1984–1988). Three are *lagged outcomes* — cigarette sales themselves at 1975, 1980, and 1988. The lagged outcomes are the most important trick: anchoring the synthetic control on the treated unit's own pre-period *outcome levels* at multiple time points forces the synthetic series to track California's pre-1988 trajectory closely.

## Donor weights and predictor weights

The optimisation produces two weight vectors that drive the entire fit. Both are extractable as tidy tables.

```{r}
#| label: donor-weights
grab_unit_weights(prop99_syn) |>
  arrange(desc(weight)) |>
  head(8)
```

```{r}
#| label: predictor-weights
grab_predictor_weights(prop99_syn) |>
  arrange(desc(weight))
```

Two things to notice.

1. **Five states absorb essentially 100% of the donor weight.** Utah, Nevada, Montana, Colorado, Connecticut. Every other state gets effectively zero. California is matched mostly to other Western/sunbelt states with similar age structure and cigarette price levels, plus Connecticut as a smoking-rate counterweight from the east.
2. **The two earliest cigsale levels dominate the V matrix.** `cigsale_1975` and `cigsale_1980` together get roughly 88% of the predictor weight. The four behavioural and demographic covariates get less than 9% combined. The optimiser has effectively decided: "the best way to predict California's cigarette sales is using *other states' cigarette sales*."

For a one-line visual of both weight vectors, tidysynth ships a `plot_weights()` helper:

```{r}
#| label: fig-sc-weights
#| fig-cap: "Donor unit weights (left) and predictor V-matrix weights (right) from tidysynth."
#| fig-width: 9
#| fig-height: 5
plot_weights(prop99_syn)
```

### A closer look at the V matrix

The combined `plot_weights()` view is convenient, but the V matrix deserves a stand-alone chart because it answers a different question than the donor weights. Donor weights say *which states* mimic California; the V matrix says *which variables* the optimiser used to decide what "mimics" means.

```{r}
#| label: fig-sc-predictor-weights
#| fig-cap: "V matrix: cigsale_1975 and cigsale_1980 dominate; behavioural covariates get nearly zero weight."
#| fig-width: 8
#| fig-height: 4
predw_df <- grab_predictor_weights(prop99_syn) |>
  mutate(variable = fct_reorder(variable, weight))

ggplot(predw_df, aes(x = weight, y = variable)) +
  geom_col(fill = "#6a9bcc") +
  geom_text(aes(label = sprintf("%.3f", weight)), hjust = -0.12) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "V-matrix weight (predictor importance)", y = NULL) +
  theme_minimal()
```

Two readings of the same picture, one practical and one cautionary.

- *Practical reading.* Two lagged outcomes (`cigsale_1975` and `cigsale_1980`) carry the bulk of the matching information. The optimiser has decided that California's pre-period cigarette sales — *at multiple time points* — are the best fingerprint to match.
- *Cautionary reading.* The V matrix is **not a causal ranking**. It tells you which variables were *useful for matching the treated unit's pre-period*, not which variables *cause* the outcome.

**Common pitfall.** Treating the V matrix as a list of causal drivers. It is a list of *good pre-period predictors for one specific unit*, not a structural model of smoking.

## The estimate

```{r}
#| label: sc-att
# grab_synthetic_control() returns a tidy tibble with observed (real_y)
# and synthetic (synth_y) cigsale for every year. We restrict to the
# post-period and compute the per-year gap.
sc_post <- grab_synthetic_control(prop99_syn) |>
  filter(time_unit > 1988) |>
  mutate(dif = real_y - synth_y)

# Average the per-year gap to recover the ATT.
mean(sc_post$dif)
```

The Synthetic Control ATT is approximately **$-18.85$ packs/capita** averaged over 1989–2000. This is the book's primary causal estimate and within rounding of the canonical @abadie2010synthetic result.

```{r}
#| label: fig-sc-trends
#| fig-cap: "Synthetic Control: California (observed) vs synthetic California (weighted donor combination)."
#| fig-width: 8
#| fig-height: 5
plot_trends(prop99_syn)
```

The pre-period fit is excellent — the synthetic and observed series are nearly indistinguishable through 1988. A substantial gap opens immediately after 1989, widening to roughly 30 packs by 2000.

## Predictor balance: did the matching work?

`grab_balance_table()` shows California, synthetic California, and the unweighted donor average side-by-side on every predictor.

```{r}
#| label: sc-balance
grab_balance_table(prop99_syn)
```

On every variable, *synthetic California* is far closer to California than the unweighted donor average is. The most dramatic improvement is on the lagged outcomes: `cigsale_1988` is roughly 90 for California vs ≈91 for the synthetic — a near-perfect match — while the unweighted donor average is around 114. That gap of ~24 packs is exactly the bias the naive pre-post method silently absorbed.

## Visualising the post-period gap

`plot_trends()` showed *both* observed and synthetic California on one canvas. The companion helper `plot_differences()` plots just the *gap*: $Y_{1t} - \widehat{Y_{1t}(0)}$, year by year. This isolates the treatment-effect curve in its cleanest form.

```{r}
#| label: fig-sc-differences
#| fig-cap: "Per-year gap between observed California and synthetic California."
#| fig-width: 8
#| fig-height: 5
plot_differences(prop99_syn)
```

Read the line as the *effect of Proposition 99 on California in year $t$*. The pre-period values hover near zero (the matching worked), the line drops sharply after 1989, and it stays negative — steadily widening — throughout the post-period. The 1989–2000 mean of this series is exactly the $-18.85$ packs ATT reported above.

## Inference via placebo permutation

A "standard error" computed as cross-year SD divided by $\sqrt{N}$ is *not* a real sampling-distribution-based standard error. The proper Synthetic Control uncertainty quantification is a permutation test.

**The recipe.** Refit the synthetic-control model treating *each donor state* as if *it* had been the treated unit. Compute the post-period gap for each placebo. Compare California's gap trajectory to those placebo trajectories. If California's gap is extreme relative to the placebos, the policy probably did something.

```{r}
#| label: fig-sc-placebos
#| fig-cap: "Placebo permutation: California in orange, donor placebos in grey (badly pre-fit donors pruned by default)."
#| fig-width: 8
#| fig-height: 5
plot_placebos(prop99_syn)
```

The orange line is California; the grey lines are the donor placebos. By default, `plot_placebos()` *prunes* placebos whose pre-period mean squared prediction error (MSPE) exceeds twice California's — those donors fit their own pre-period so badly that comparing their post-period gap to California's would be misleading. After pruning, California's post-period gap sits visibly below every retained placebo, which is the visual signature of a "real" treatment effect.

The unpruned variant keeps every donor for full transparency:

```{r}
#| label: fig-sc-placebos-unpruned
#| fig-cap: "Same placebo plot with every donor retained — including badly pre-fit ones — to show the full pool of placebo trajectories."
#| fig-width: 8
#| fig-height: 5
plot_placebos(prop99_syn, prune = FALSE)
```

With pruning off, the grey cloud is messier and a few badly-fit donors swing wildly — but California's post-period descent still ends up at the bottom of the bundle. The qualitative conclusion does not depend on the pruning rule.

## MSPE ratio and Fisher exact p-value

A sharper version of the same test is the **MSPE ratio** — the ratio of post-period to pre-period mean squared prediction error. If a unit has a tight pre-period fit *and* a large post-period gap, the ratio is large.

```{r}
#| label: sc-significance
# grab_significance() returns one row per unit (treated + every placebo)
# with pre_mspe, post_mspe, the post/pre ratio, the unit's rank in that
# ratio, and the Fisher-style p-value (rank / n_units).
grab_significance(prop99_syn) |>
  arrange(desc(mspe_ratio)) |>
  head(5)
```

California's MSPE ratio is around **124** — more than two and a half times higher than the next-highest unit. California ranks **1st out of 39 units**. The Fisher exact $p$-value is rank divided by total units, so $1/39 \approx 0.026$. Under the null hypothesis that Proposition 99 had no effect, the probability of seeing a unit this extreme purely by chance is about 2.6%.

```{r}
#| label: fig-sc-mspe-ratio
#| fig-cap: "MSPE-ratio bar chart with California highlighted at rank 1 out of 39 units."
#| fig-width: 8
#| fig-height: 5
plot_mspe_ratio(prop99_syn)
```

The orange bar at the top is California; every blue bar below it is a placebo donor. The gap between California and the second-place state is enormous. That gap is the visual signature of "a real treatment effect that the donor pool does not naturally replicate".

## Inspecting the nested tidysynth object

`prop99_syn` is not a plain data frame — it is a *nested tibble* with one row per unit (treated unit + every donor refit as a placebo) and list-columns that hold every intermediate output of the optimisation.

```{r}
#| label: sc-nested
prop99_syn
```

Each list-column can be flattened with `tidyr::unnest()` for custom downstream work, or pulled out with one of the `grab_*()` helpers used above.

```{r}
#| label: sc-unnest
# Flatten .outcome into a wide table: one row per (unit, year).
# The actual cigarette-sales column is named after each unit, so we
# select metadata + California's series for a clean preview.
prop99_syn |>
  tidyr::unnest(cols = c(.outcome)) |>
  select(.id, .placebo, .type, time_unit, California) |>
  head(8)
```

For the tidy *observed-vs-synthetic* table (which is what most analyses want), the dedicated helper is more convenient:

```{r}
#| label: sc-unnest-tidy
grab_synthetic_control(prop99_syn) |> head(8)
```

This is the whole point of the nested-tibble design: every step of the optimisation is *introspectable from R*, with no need to dig into S4 slots or `attr()` blobs.

## Recap

| Question | Answer |
|---|---|
| What does Synthetic Control estimate? | The ATT on California, 1989–2000 |
| What is the point estimate? | **$-18.85$ packs/capita per year** |
| What is "synthetic California"? | A convex combination of five Western/sunbelt states (Utah, Nevada, Montana, Colorado, Connecticut) |
| What predictors did the matching? | Mostly two lagged outcomes — `cigsale_1975` and `cigsale_1980` |
| How is the matching quality? | Excellent — synthetic and observed California are near-identical through 1988 |
| What is the inference statistic? | Fisher exact $p \approx 0.026$ (California ranks 1st out of 39 on the MSPE ratio) |
| What is the design-time pitfall? | Don't read the V matrix as a list of causal drivers — it is a list of *good pre-period predictors* |

Synthetic Control is the book's headline causal estimate, and the placebo / MSPE-ratio diagnostics both confirm that California's post-1989 trajectory is unusual relative to what other states experienced in the same window. In chapter 6 we hand the same donor information to a Bayesian model and ask whether a *credible interval* (a direct probability statement about the effect) tells the same story.

## Further reading

- @abadie2010synthetic — the original synthetic-control treatment of Proposition 99.
- @abadie2021using — methodological feasibility, data requirements, and assumptions in a JEL review.
- @dunford2024tidysynth — documentation for the `tidysynth` package used here.
