---
title: "Structural Bayesian Time Series"
---

## Learning objectives

1. Fit a BSTS local-level trend model with a spike-and-slab prior over donor regressors using `CausalImpact`. Spike-and-slab is what makes the method tractable on the 194-donor pre-period and is the right tool whenever predictors outnumber pre-period observations.
2. Impute missing donor covariates with random-forest multiple imputation and reshape into the wide matrix `CausalImpact` expects. Real donor panels (income, beer, demographics) come with 16–55% missingness, and skipping imputation throws away most of the pre-period information.
3. Extract posterior inclusion probabilities to see which donors the BSTS fit actually relies on. Inclusion probabilities are the Bayesian analogue of the SC donor weights from chapter 4 and let the reader audit the counterfactual.
4. Diagnose MCMC convergence with Rhat and effective sample size before reading off credible intervals. The chapter's default of 1,000 iterations is deliberately tight, and a reader who skips this step risks reporting credible bands from an unconverged chain.
5. Read CausalImpact credible intervals as posterior probability statements, not frequentist confidence coverage. This is the only method in the book that produces direct probability statements about the effect, and conflating the two interpretations is one of the most common reporting errors in applied work.

## The BSTS / CausalImpact idea

Chapter 4 built California's counterfactual from a *convex* blend of donors and chapter 5 bias-corrected that blend with a ridge outcome model; this chapter borrows the donor information a third way. Fit a **Bayesian structural time-series (BSTS)** model on the pre-period. Use *other states' cigarette sales* (and optionally covariates) as predictors. Project the fitted model forward as the counterfactual. The posterior over (observed − projected) gives a credible interval for the policy effect. We will write the resulting point estimate as $\widehat{\tau}_{\text{BSTS}}$ — the posterior-mean ATT — to keep it distinct from chapter 4's simplex-weight estimator and chapter 7's `scpi` prediction-interval estimator.

This is the only method in the book that delivers a *credible* interval (CrI) — a direct probability statement about the parameter. Concretely, the 95% **credible interval** that `CausalImpact` reports is a Bayesian object: "given the model and the data, there is a 95% posterior probability that the effect lies in this range". This is **not** a frequentist 95% confidence interval, which would instead say "if we re-ran the entire study many times, 95% of the intervals we construct would cover the true effect". The two answer different questions and need not coincide numerically. We will write "CrI" rather than "CI" throughout to keep the distinction visible.

## The model in two pieces

The BSTS counterfactual is

$$Y_{1t} = \mu_t + \beta^\top x_t + \varepsilon_t, \quad t \le t^*$$

where $\mu_t$ is a local-level trend, $x_t$ are the control-series regressors (other states' `cigsale`, plus optional covariates), and $t^*$ is the intervention date. California's outcome is *a slowly-evolving trend* **plus** *a linear combination of donor-state series* **plus** *a random error*.

**The spike-and-slab prior.** The donor pool here is 194 columns wide (39 donor states × 5 variables minus California's outcome), but the pre-period is only 19 years long. An unregularised regression is mathematically impossible — there are an order of magnitude more candidate coefficients than observations. `CausalImpact` handles this with a **spike-and-slab** prior on $\beta$ (specifically, the *Stochastic Search Variable Selection* or SSVS form of @george1997approaches): each coefficient $\beta_j$ has prior probability $\pi^{\text{prior}}$ of being non-zero (drawn from a "slab" — a diffuse Gaussian) and probability $1 - \pi^{\text{prior}}$ of being exactly zero (the "spike"). The default `expected.model.size = 3` sets $\pi^{\text{prior}} = 3/194$, so the prior strongly favours sparse counterfactuals built from a handful of donors. After fitting, each donor has a **posterior inclusion probability**
$$\pi_j \;=\; \Pr(\beta_j \ne 0 \mid Y),$$
estimated by the share of post-burn-in MCMC iterations in which $\beta_j$ was non-zero. This $\pi_j$ is the natural Bayesian analogue of the simplex weights $w_j$ from chapter 4 — but the two are different objects: $w_j$ is a non-negative real number that sums to one across donors, while $\pi_j$ is a probability that lives in $[0, 1]$ for each donor independently. Do not equate them. @scott2014predicting motivate why this prior, originally designed for "nowcasting" with sparse predictors, is also the right device for short-pre-period counterfactual forecasting.

```{mermaid}
%%| label: fig-bsts-counterfactual
%%| fig-cap: "Decomposition of the BSTS counterfactual: a local-level trend, a regression on donor cigsale + covariates, and random error sum to Ŷ₁ₜ, which is then subtracted from the observed series to give the policy effect."
flowchart TB
    subgraph "BSTS counterfactual Ŷ₁ₜ"
        TREND["μₜ — local-level trend<br/>(absorbs dynamics no control can explain)"]
        REG["β·xₜ — regression on donor cigsale + covariates<br/>(borrows from donor pool)"]
        ERR["εₜ — random error"]
    end
    TREND --> Y["Ŷ₁ₜ"]
    REG --> Y
    ERR --> Y
    Y --> CMP["Observed Y₁ₜ − Ŷ₁ₜ<br/>= policy effect (with credible band)"]

    style TREND fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style REG fill:#00d4c8,stroke:#cbd5e0,color:#141413
    style ERR fill:#7a8395,stroke:#cbd5e0,color:#fff
    style Y fill:#7A209F,stroke:#cbd5e0,color:#fff
    style CMP fill:#d97757,stroke:#cbd5e0,color:#fff
```

The trend $\mu_t$ absorbs the dynamics that no control series can explain; the regression term $\beta^\top x_t$ borrows information from the donor pool. After the model is fit on $t \le t^*$, it is projected forward to give $\widehat{Y_{1t}(0)}$, and the posterior over $Y_{1t} - \widehat{Y_{1t}(0)}$ gives the credible interval for the policy effect. Averaging that pointwise gap over the post-period yields $\widehat{\tau}_{\text{BSTS}}$.

**Contrast with classical synthetic control (chapter 4).** This is the same donor-pool idea as classical synthetic control, but with a different prior over the weights *and* a different output object. Chapter 4 returns a single **point estimate** of the ATT plus a placebo rank; chapter 6 returns a full **posterior distribution** over the ATT, from which the point estimate $\widehat{\tau}_{\text{BSTS}}$ and the 95% CrI are just two summaries. The mechanical difference behind that is the constraint set: synthetic control constrains the weight vector $w$ to the unit simplex (non-negative, sum to one), so it gives **convex** weights (every "synthetic California" is a weighted average of donor states that lives inside their convex hull) and a Fisher exact $p$-value from placebo permutation; BSTS lets $\beta$ be any sparse real-valued vector under the spike-and-slab prior, so it gives possibly **negative** coefficients $\beta_j$ (not weights $w_j$ — the two play analogous roles, but live on different supports) and a posterior credible interval. The two estimates need not agree, and when they disagree the reason almost always lives in this prior choice over $w$ versus $\beta$.

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `CausalImpact` is Google's wrapper around the `bsts` (Bayesian Structural Time Series) package: it takes a wide matrix with the treated outcome in column 1 and donor series in the remaining columns, fits a Bayesian state-space model on the pre-period, projects forward as the counterfactual, and returns posterior means + 95% credible intervals on the pointwise and cumulative effects. `mice` [@vanbuuren2011mice] provides multiple imputation by chained equations — we use its random-forest mode (here with $m = 1$, so this is *single* random-forest imputation, i.e. one draw from what would otherwise be a multiple-imputation procedure) to fill the missing covariate values before pivoting to wide form. The `R/table_helpers.R` helper provides `gt_pretty()` for the summary table.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, set seed, and configure transparent ggplot theme."
library(tidyverse)
library(CausalImpact)
library(mice)
source("R/table_helpers.R")

set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** Like Synthetic Control, this method uses the **full 39-state × 31-year panel**: every state other than California is potentially a regressor in the Bayesian counterfactual model, and the four covariates (`lnincome`, `retprice`, `age15to24`, `beer`) are candidate predictors too. We do not pre-filter or restrict the window. The dataset is loaded as-is and then reshaped from long to wide in the next section, which is the form `CausalImpact` requires.

```{r}
#| label: data-load
#| code-summary: "Code: Load the Proposition 99 panel from the bundled RDS file."
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()
```

The loaded `prop99` is the same 1,209-row × 7-column tibble used in chapter 4 (39 states × 31 years × the `cigsale` outcome and four covariates). The next section deals with the missing covariate values and the long-to-wide reshape that CausalImpact requires.

## Imputation and wide pivot

CausalImpact wants a *wide* dataset with the treated outcome in column 1 and every control series in the remaining columns. The covariate columns have missing values (`lnincome` is missing 16% of rows, `age15to24` 32%, `beer` 55%), so we fill them with single random-forest imputation from `mice`.

```{r}
#| label: impute-and-pivot
#| message: false
#| warning: false
#| code-summary: "Code: Impute missing covariates with random forests and pivot panel to wide form."
# Fill in missing covariate values with one round of random-forest
# multiple imputation (m = 1, method = "rf"). printFlag suppresses
# console chatter.
prop99_imputed <- prop99 |>
  mice(m = 1, method = "rf", printFlag = FALSE) |>
  complete() |>
  as_tibble()

# Pivot the long panel to wide format: one column per (variable, state)
# pair. CausalImpact requires the treated outcome in column 1, hence
# relocate(cigsale_California) and drop the year index.
prop99_wide <- prop99_imputed |>
  pivot_wider(names_from  = state,
              values_from = c(cigsale, lnincome, beer, age15to24, retprice)) |>
  relocate(cigsale_California) |>
  select(-year)

dim(prop99_wide)
```

The reported `dim()` confirms the reshape: `prop99_wide` is **31 rows × 195 columns** — one row per year 1970–2000, and 39 states × 5 variables in the columns (`cigsale_*`, `lnincome_*`, `beer_*`, `age15to24_*`, `retprice_*` for each of the 39 states), with `year` dropped and `cigsale_California` relocated to column 1 because that's what `CausalImpact()` expects as the treated outcome.

## Fit the Bayesian structural TS model

**The fit.** The chunk below calls `CausalImpact(prop99_wide, pre.period, post.period)` with the pre and post windows expressed as *integer row indices* (not years): row 1 is 1970, row 19 is 1988 (the last pre-period year), row 20 is 1989, row 31 is 2000. Under the hood, `CausalImpact` uses an MCMC sampler — by default 1,000 iterations with the first 100 discarded as burn-in. For a tutorial on 19 pre-period observations that default is fine, but production work should rerun with `model.args = list(niter = 10000)` and inspect convergence diagnostics (we add a few of these below). The returned `impact_full$summary` is a two-row data frame — one row for the *average* effect over 1989–2000 and one row for the *cumulative* effect — with posterior means and 95% credible bounds for each. The helper code below the `CausalImpact()` call just formats those columns into the table.

```{r}
#| label: tbl-causalimpact
#| tbl-cap: "CausalImpact posterior summary; brackets are 95% credible intervals."
#| message: false
#| warning: false
#| code-summary: "Code: Fit CausalImpact on the full donor matrix and format posterior summary table."
# CausalImpact takes integer row indices, not years.
# Row 1 = 1970, row 19 = 1988 (last pre-period year).
# Row 20 = 1989, row 31 = 2000 (last observed year).
pre_idx  <- c(1, 19)
post_idx <- c(20, 31)

# We set set.seed(42) at the top of the chapter for the mice random-
# forest imputation. CausalImpact wraps its MCMC sampler so that re-
# running the same chunk on the same machine reproduces the posterior
# draws; for cross-machine reproducibility, set niter explicitly and
# rerun. The global R seed above governs reproducibility of any
# downstream code that consumes the posterior draws.

impact_full <- CausalImpact(prop99_wide,
                            pre.period  = pre_idx,
                            post.period = post_idx)

ci_summary <- impact_full$summary |>
  tibble::rownames_to_column("Horizon") |>
  dplyr::transmute(
    Horizon,
    Actual            = sprintf("%.1f", Actual),
    Prediction        = sprintf("%.1f [%.1f, %.1f]",
                                Pred, Pred.lower, Pred.upper),
    `Absolute effect` = sprintf("%.1f [%.1f, %.1f]",
                                AbsEffect, AbsEffect.lower, AbsEffect.upper),
    `Relative effect` = sprintf("%.1f%% [%.1f%%, %.1f%%]",
                                RelEffect * 100, RelEffect.lower * 100,
                                RelEffect.upper * 100),
    `Posterior p`     = sprintf("%.3f", p)
  )

gt_pretty(ci_summary)
```

**Reading the output.** The headline numbers below are extracted inline from `impact_full$summary` rather than hardcoded, so they update automatically if the data, prior, or MCMC budget changes.

```{r}
#| label: headline-numbers
#| echo: false
ci_avg <- impact_full$summary["Average", ]
ci_cum <- impact_full$summary["Cumulative", ]
fmt1 <- function(x) sprintf("%.1f", x)
```

- **Average ATT:** $\widehat{\tau}_{\text{BSTS}} \approx$ `r fmt1(ci_avg$AbsEffect)` packs/capita, 95% credible interval $[$ `r fmt1(ci_avg$AbsEffect.lower)`, `r fmt1(ci_avg$AbsEffect.upper)` $]$ — see "Common pitfall" below for why this interval is mildly anti-conservative under $m = 1$ imputation.
- **Cumulative effect:** $\approx$ `r fmt1(ci_cum$AbsEffect)` packs over 12 years, with 95% CrI $[$ `r fmt1(ci_cum$AbsEffect.lower)`, `r fmt1(ci_cum$AbsEffect.upper)` $]$, or roughly `r sprintf("%.0f%%", -ci_cum$RelEffect * 100)` of what would have been expected absent the policy.
- **Posterior probability of any causal effect:** $\approx$ `r sprintf("%.0f%%", (1 - ci_avg$p) * 100)`.

### Which donors did the spike-and-slab pick?

With 194 candidate regressors and an `expected.model.size = 3` prior, most coefficients are shrunk to zero in most MCMC iterations. The natural diagnostic is the **posterior inclusion probability** $\pi_j = \Pr(\beta_j \ne 0 \mid Y)$ — for each column, the share of post-burn-in MCMC iterations in which the column's coefficient was *not* in the spike. Calling `plot()` on the underlying `bsts` model with `"coefficients"` returns a horizontal bar plot of the top contributors.

```{r}
#| label: fig-inclusion
#| fig-cap: "Posterior inclusion probabilities for the donor regressors. Bars are the share of MCMC iterations in which each donor's coefficient was non-zero; longer bars are the donors the spike-and-slab prior kept in the model. The colour indicates the posterior-mean sign of the coefficient when it is included."
#| fig-width: 7
#| fig-height: 5
#| code-summary: "Code: Plot spike-and-slab posterior inclusion probabilities for donor regressors."
plot(impact_full$model$bsts.model, "coefficients")
```

Notice that with 194 candidate columns and a prior that expects only 3 of them to matter, the inclusion probabilities $\pi_j$ are individually low — even the top donor sits around 5–10%, not at the 80–100% levels you might see in a more identified problem. The top contributors here are typically *covariate* columns (state-level `retprice`, `age15to24`, `beer`) rather than the raw donor `cigsale_*` series — a striking contrast with chapter 4, where the simplex weights $w_j$ placed all the mass on Utah, Nevada, Montana, Colorado, and Connecticut's *cigarette-sales* series. Two warnings carry over from chapter 4: (i) a high $\pi_j$ does *not* mean a donor is "causally similar" to California — it means the donor's column is useful for *predicting* California pre-1989; and (ii) unlike SC's simplex weights, BSTS coefficients $\beta_j$ can be negative, so a column with a high $\pi_j$ and a negative posterior-mean coefficient is being used as a *contrast* rather than as a positive ingredient.

### MCMC convergence diagnostics

With only 1,000 iterations (the default) and 19 pre-period observations, it's worth sanity-checking that the sampler has actually converged before reading the credible interval off the screen. Two cheap checks: a visual one on the state-space components, and a numerical one on the regression coefficients of the donors that are *most often* included in the model.

```{r}
#| label: fig-bsts-components
#| fig-cap: "BSTS state-space components: trend (μₜ) and the contribution of the regression term. A visually flat, mean-reverting residual is consistent with sampler convergence; obvious drift or stickiness is not."
#| fig-width: 7
#| fig-height: 6
#| dev.args: !expr 'list(bg = "white")'
#| code-summary: "Code: Plot BSTS state-space components (trend and regression contribution)."
plot(impact_full$model$bsts.model, "components")
```

```{r}
#| label: tbl-mcmc-diagnostics
#| tbl-cap: "Convergence diagnostics for the donor regressors with the highest posterior inclusion probability. Rhat near 1.0 and ESS well above 100 are the standard rules of thumb; values far from these warrant a longer chain via `model.args = list(niter = 10000)`."
#| message: false
#| warning: false
#| code-summary: "Code: Compute Rhat and ESS diagnostics for top-included donor coefficients."
# Extract the post-burn-in coefficient draws (iterations x donors)
beta_draws <- impact_full$model$bsts.model$coefficients
burn       <- max(1L, floor(0.1 * nrow(beta_draws)))
beta_post  <- beta_draws[seq.int(burn + 1L, nrow(beta_draws)), , drop = FALSE]

# Inclusion probability per donor: share of iterations the coef is non-zero.
incl_prob  <- colMeans(beta_post != 0)
top_donors <- names(sort(incl_prob, decreasing = TRUE))[seq_len(5)]

# Helpers: split a single chain into halves, compute a Gelman-style Rhat,
# and an effective-sample-size estimate from the autocorrelation of the
# non-zero portion of the draws. These are coarse — meant as a sanity
# check, not a publication-grade diagnostic.
rhat_one <- function(x) {
  n <- length(x); if (n < 4) return(NA_real_)
  h <- floor(n / 2); a <- x[seq_len(h)]; b <- x[(h + 1):(2 * h)]
  W <- mean(c(var(a), var(b))); if (W <= 0) return(NA_real_)
  B <- h * var(c(mean(a), mean(b)))
  sqrt(((h - 1) / h * W + B / h) / W)
}
ess_one <- function(x) {
  nz <- x[x != 0]; n <- length(nz); if (n < 10) return(NA_real_)
  rho <- stats::acf(nz, plot = FALSE, lag.max = min(50, n - 1))$acf[-1]
  keep <- rho > 0.05; if (!any(keep)) return(n)
  tau <- 1 + 2 * sum(rho[seq_len(which.min(keep) - 1)])
  n / max(tau, 1)
}

mcmc_diag <- tibble::tibble(
  Donor                = top_donors,
  `Inclusion prob.`    = sprintf("%.2f", incl_prob[top_donors]),
  `Posterior mean (β)` = sprintf("%.3f",
                                 colMeans(beta_post[, top_donors, drop = FALSE])),
  Rhat                 = sprintf("%.2f",
                                 vapply(top_donors,
                                        function(d) rhat_one(beta_post[, d]),
                                        numeric(1))),
  ESS                  = sprintf("%.0f",
                                 vapply(top_donors,
                                        function(d) ess_one(beta_post[, d]),
                                        numeric(1)))
)

gt_pretty(mcmc_diag)
```

For 1,000 default iterations the components plot should look stable and the Rhat values should sit near 1; ESS values in the low hundreds are tolerable for a tutorial but on the thin side for publication. Re-running with `model.args = list(niter = 10000)` is the easy upgrade.

### What if we drop the covariates?

The covariate columns are heavily imputed (`beer` was 55% missing). A natural robustness check is to re-fit using only other states' `cigsale` series — a leaner donor pool that lets us see how much of the headline result is driven by the covariates. We rebuild a covariate-free wide matrix from the *original, un-imputed* `prop99` (so no imputation enters this fit at all) and pass it through `CausalImpact` with the same pre/post windows.

```{r}
#| label: tbl-causalimpact-nocov
#| tbl-cap: "CausalImpact fit with only donor `cigsale` columns (no imputed covariates)."
#| message: false
#| warning: false
#| code-summary: "Code: Refit CausalImpact using only donor cigsale columns as a covariate-free robustness check."
prop99_nocov <- prop99 |>
  dplyr::select(year, state, cigsale) |>
  tidyr::pivot_wider(names_from = state, values_from = cigsale) |>
  dplyr::relocate(California) |>
  dplyr::select(-year)

impact_nocov <- CausalImpact(prop99_nocov,
                             pre.period  = pre_idx,
                             post.period = post_idx)

ci_summary_nocov <- impact_nocov$summary |>
  tibble::rownames_to_column("Horizon") |>
  dplyr::transmute(
    Horizon,
    Actual            = sprintf("%.1f", Actual),
    Prediction        = sprintf("%.1f [%.1f, %.1f]",
                                Pred, Pred.lower, Pred.upper),
    `Absolute effect` = sprintf("%.1f [%.1f, %.1f]",
                                AbsEffect, AbsEffect.lower, AbsEffect.upper),
    `Relative effect` = sprintf("%.1f%% [%.1f%%, %.1f%%]",
                                RelEffect * 100, RelEffect.lower * 100,
                                RelEffect.upper * 100),
    `Posterior p`     = sprintf("%.3f", p)
  )

gt_pretty(ci_summary_nocov)
```

Without covariates the donor matrix collapses to 38 columns (one `cigsale` per donor state), so the spike-and-slab now selects among states only. Compared to the covariate fit above, the point estimate typically moves a few packs and the posterior probability of a non-zero effect shifts accordingly — the covariates absorb some of the variation the simpler model was attributing to Proposition 99, which can be read as either "added robustness" or "watered-down signal" depending on how much you trust the imputed beer-and-income covariates.

## The two-panel diagnostic

**What to look for.** Calling `plot()` on a `CausalImpact` object returns its default two-panel figure: the **top panel** shows observed California (solid line) against the Bayesian counterfactual (dashed) with its 95% credible band; the **bottom panel** shows the *cumulative* gap — the running sum of `observed − counterfactual` from the intervention date onwards, with its own credible band. Read the top panel for whether and when the policy effect opens up, and the bottom panel for whether the cumulative effect's credible band excludes zero by the end of the series.

```{r}
#| label: fig-causalimpact
#| fig-cap: "CausalImpact two-panel: pointwise observed vs Bayesian counterfactual, and cumulative effect over time."
#| fig-width: 8
#| fig-height: 7
#| code-summary: "Code: Render the CausalImpact two-panel pointwise and cumulative effect plot."
plot(impact_full)
```

The top panel shows the pointwise picture: observed California opens a steady gap below the Bayesian counterfactual starting in 1989, with a 95% credible band that widens as we forecast further from the training window. The bottom panel cumulates that gap over time. By 2000 the cumulative effect is approximately `r fmt1(ci_cum$AbsEffect)` packs/capita with a 95% credible interval of $[$ `r fmt1(ci_cum$AbsEffect.lower)`, `r fmt1(ci_cum$AbsEffect.upper)` $]$ — wide enough that zero sits inside the band, which is why the headline posterior probability of a non-zero effect is only $\approx$ `r sprintf("%.0f%%", (1 - ci_cum$p) * 100)`, not the $> 99\%$ the eyeball test of the top panel might suggest. This is the gap between *pointwise* certainty (the dashed line is clearly above the solid one) and *cumulative* certainty (twelve consecutive partly-overlapping pointwise bands compound into a wider cumulative band).

## Common pitfall

Imputing missing covariates without thinking about the imputation model. The random-forest fill we use here is a single-imputation shortcut for tutorial speed. With multiple imputation ($m > 1$) or a different model, the estimate can move by 1–3 packs. Two concrete failure modes are worth spelling out, because they affect the credibility of the headline number rather than just its width.

**Example 1 — Imputing donor covariates with a model that "sees" California.** Our `mice(prop99, ...)` call passes the entire panel — California rows included — into the imputation model. For this dataset, with only four covariates and most missingness clustered in `beer` (≈55%) and `age15to24` (≈32%) on donor states, the consequence is usually mild. But if you were to feed the *full panel with the post-period California outcome attached* into an imputer that also uses `cigsale` as a predictor, the imputer could learn from California's post-1989 trajectory and propagate that information back into donor states' missing covariate cells. CausalImpact would then build a counterfactual that mechanically tracks California's post-period — biasing the estimated effect *toward zero* by construction. The safe pattern is to impute donor covariates *before* the treated outcome is in scope, or to drop the treated unit from the imputer entirely and impute donors only.

**Example 2 — Single (`m = 1`) vs multiple (`m = 5`) imputation.** Single imputation treats the imputed values as if they were observed, then propagates that pretend-observed dataset through `CausalImpact`. The posterior credible band you get back is the model's uncertainty *given* the imputed cells — it knows nothing about the uncertainty *introduced by the imputation itself*. With proper multiple imputation (`mice(..., m = 5)` and Rubin's-rules pooling across the five resulting fits), the credible band widens by roughly 1–3 packs on this dataset because the across-imputation variance is added to the within-imputation variance. The point estimate barely moves; the interval does. For a tutorial we keep `m = 1` for speed and clarity, but a published number should use `m ≥ 5` and report a pooled estimate.

**Example 3 — Posterior inclusion probabilities are not causal weights.** As in chapter 4's V-matrix, a donor state with high $\pi_j$ is one that is *useful for predicting California's pre-period sales*, not one that is *causally similar* to California. The two are correlated but they are not the same; reading the inclusion-probability plot as a "ranking of causally-similar states" is the same mistake chapter 4 warned about with the V matrix. And $\pi_j$ should never be plotted on the same axis as chapter 4's $w_j$ — they are not the same quantity, only analogous diagnostics.

**Example 4 — No MCMC convergence check.** It is easy to take the 95% credible interval at face value without ever asking whether the sampler that produced it has converged. The default 1,000 iterations is on the thin side for a 195-column donor pool, and the diagnostics chunk above (Rhat, ESS, components plot) is the minimum honest check. If Rhat is well above 1 or ESS for a top donor is in the single digits, the credible band is not yet a credible posterior — rerun with `model.args = list(niter = 10000)`.

## Where to next

The credible interval around $\widehat{\tau}_{\text{BSTS}}$ here is a Bayesian object — its width depends on the prior. The natural follow-up question is whether a **frequentist** prediction interval, built without a prior over donor weights, would tell the same story. Chapter 7 answers that with the `scpi` framework of @cattaneo2021prediction and @cattaneo2025scpi, which constructs a finite-sample prediction interval around the *synthetic-control* counterfactual from chapter 4 by separating in-sample weight uncertainty from out-of-sample forecast error.

## Key takeaways

**Methods:**

- Bayesian structural time series (BSTS) — implemented in `CausalImpact` — fits a state-space model $Y_{1t} = \mu_t + \beta^\top x_t + \varepsilon_t$ on the pre-period, where $\mu_t$ is a local-level trend and $\beta^\top x_t$ is a regression on donor-state series and covariates; projecting the fit forward yields the posterior counterfactual $\widehat{Y_{1t}(0)}$, and averaging $Y_{1t} - \widehat{Y_{1t}(0)}$ over the post-period gives the posterior-mean ATT $\widehat{\tau}_{\text{BSTS}}$.
- A spike-and-slab (SSVS) prior on $\beta$ — with default `expected.model.size = 3` out of 194 candidate regressors — performs sparse Bayesian variable selection, producing **posterior inclusion probabilities** $\pi_j = \Pr(\beta_j \ne 0 \mid Y)$ that play the role classical synthetic control's simplex weights $w_j$ play, but allow negative coefficients and lie outside the donors' convex hull. The two indices are *not* interchangeable: $w_j \in [0,1]$ with $\sum_j w_j = 1$, while $\pi_j \in [0,1]$ independently for each donor.
- The output is a **posterior** distribution — a full probability distribution over the unknown effect given model and data — summarised by a 95% **credible interval (CrI)**, the Bayesian interval inside which the parameter lies with 95% probability; this is a direct probability statement about the effect, unlike the frequentist placebo $p$-values from chapter 4, which characterise the procedure rather than the parameter.

**Lessons:**

- On Prop 99, CausalImpact estimates an average ATT of $\widehat{\tau}_{\text{BSTS}} \approx$ `r fmt1(ci_avg$AbsEffect)` packs/capita (cumulative $\approx$ `r fmt1(ci_cum$AbsEffect)` packs over 1989–2000), with a posterior probability of any non-zero effect of $\approx$ `r sprintf("%.0f%%", (1 - ci_avg$p) * 100)` — qualitatively the same story as classical synthetic control, but expressed as a posterior probability rather than as a placebo rank.
- Pointwise certainty and cumulative certainty are not the same: the dashed counterfactual sits visibly above observed California from 1989 on, yet the 95% credible band on the *cumulative* effect by 2000 ($[$ `r fmt1(ci_cum$AbsEffect.lower)`, `r fmt1(ci_cum$AbsEffect.upper)` $]$) still contains zero because twelve overlapping pointwise bands compound into a wider cumulative one.
- Dropping the imputed covariates and refitting on donor `cigsale` columns alone moves the headline estimate by a few packs and shifts the posterior probability accordingly — a useful built-in robustness check that exposes how much of the result rides on the imputed `beer`, `lnincome`, and `age15to24` columns rather than on the cigarette-sales donor pool itself.

**Caveats:**

- The credible interval is conditional on the prior. The spike-and-slab `expected.model.size = 3` choice strongly favours sparse counterfactuals; a different shrinkage prior (e.g. horseshoe) or a different expected model size would yield different inclusion probabilities $\pi_j$ and a different — possibly wider — CrI, so prior sensitivity should be reported, not assumed away.
- Single ($m = 1$) random-forest imputation treats the imputed covariate cells as if observed, so the reported credible band understates uncertainty by 1–3 packs relative to proper multiple imputation ($m \ge 5$ with Rubin's-rules pooling); imputing donors with a model that "sees" the treated unit's post-period is the more dangerous failure, biasing $\widehat{\tau}_{\text{BSTS}}$ toward zero by construction.
- With 19 pre-period observations and 1,000 default MCMC iterations, convergence is not automatic — Rhat near 1 and ESS in the low hundreds are the minimum honest check, and `model.args = list(niter = 10000)` is the cheap upgrade for any non-tutorial use. As in chapter 4, high posterior inclusion probability $\pi_j$ means a donor *predicts* California well, not that it is causally similar.

## Further reading

- @brodersen2015inferring — the original `CausalImpact` paper.
- @brodersen-causalimpact-talk — a 50-minute walk-through of the CausalImpact intuition by the package's lead author.
- @causalimpact-pkg — package documentation and worked examples.
- @scott2014predicting — the BSTS package and its use of spike-and-slab regression for short-pre-period prediction.
- @george1997approaches — the SSVS spike-and-slab variable-selection prior that underlies the regression component.
- @vanbuuren2011mice — the `mice` package and the multiple-imputation-by-chained-equations framework, including why $m = 1$ is a shortcut and not the recommended default.

## Exercises

These exercises stretch the chapter's "what could move this credible interval?" question: change the MCMC budget, change the intervention date (placebo), interrogate the forecast-horizon uncertainty, derive the posterior probability of an effect from the raw posterior $p$, and finally run the *proper* multiple-imputation pipeline that the "Common pitfall" section warned against skipping. All exercises reuse `prop99`, `prop99_wide`, `impact_full`, `pre_idx`, and `post_idx` from the setup chunks above.

### Exercise 1: MCMC iteration sensitivity

The chapter's default fit uses `CausalImpact`'s default 1,000 MCMC iterations. Refit with `niter = 200` (deliberately under-sampled) and `niter = 2000` (a more honest setting), and compare the average-effect 95% credible-interval widths. How sensitive is the headline interval to MCMC budget?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-01
#| message: false
#| warning: false
ci_one <- function(n) {
  CausalImpact(prop99_wide,
               pre.period  = pre_idx,
               post.period = post_idx,
               model.args  = list(niter = n))
}

impact_short <- ci_one(200)
impact_long  <- ci_one(2000)

ci_width <- function(impact, label) {
  s <- impact$summary["Average", ]
  tibble(setting    = label,
         abs_effect = s$AbsEffect,
         lower      = s$AbsEffect.lower,
         upper      = s$AbsEffect.upper,
         width      = s$AbsEffect.upper - s$AbsEffect.lower)
}

bind_rows(
  ci_width(impact_short, "niter = 200"),
  ci_width(impact_full,  "niter = 1000 (chapter)"),
  ci_width(impact_long,  "niter = 2000")
) |>
  gt_pretty(decimals = 2)
```

The point estimate barely moves across the three settings — the posterior mean is well-identified even by `niter = 200`. The *credible interval width* is more sensitive: with only 200 draws, the band can be slightly biased by the empirical quantile estimator's variance. The chapter's default of 1,000 is a reasonable tutorial setting; for a publishable headline number, 5,000–10,000 iterations with explicit Rhat/ESS checks is the prudent floor.
:::

### Exercise 2: In-time placebo at 1985

Refit `CausalImpact` pretending the intervention was 1985 instead of 1989, with the panel restricted to pre-1989 data so the real policy can never enter the training window. The pseudo-post window is 1986–1988 (3 years). If BSTS is well calibrated, the placebo "ATT" should be small in absolute value.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-02
#| message: false
#| warning: false
# Restrict to pre-1989 data: keep rows 1-19 of the wide matrix.
prop99_wide_pre89 <- prop99_wide[1:19, ]

# Pretend the intervention was 1985, so:
#   Row 1 = 1970, row 16 = 1985 (last pre-period).
#   Row 17 = 1986, row 19 = 1988 (pseudo-post).
impact_placebo <- CausalImpact(prop99_wide_pre89,
                               pre.period  = c(1, 16),
                               post.period = c(17, 19))

impact_placebo$summary |>
  tibble::rownames_to_column("Horizon") |>
  dplyr::transmute(
    Horizon,
    Actual            = sprintf("%.1f", Actual),
    Prediction        = sprintf("%.1f [%.1f, %.1f]",
                                Pred, Pred.lower, Pred.upper),
    `Absolute effect` = sprintf("%.1f [%.1f, %.1f]",
                                AbsEffect, AbsEffect.lower, AbsEffect.upper),
    `Posterior p`     = sprintf("%.3f", p)
  ) |>
  gt_pretty()
```

The placebo average effect is small in absolute value and its credible interval comfortably straddles zero; the posterior $p$ is far above the conventional 0.05 threshold. The BSTS machinery does not manufacture a meaningful effect when no policy was passed in the pseudo-post window. Compare with the chapter's real fit: $\widehat{\tau}_{\text{BSTS}} \approx$ `r fmt1(ci_avg$AbsEffect)` packs/capita and $p \approx$ `r sprintf("%.2f", ci_avg$p)`.
:::

### Exercise 3: Pointwise credible-band width by post-year

The 95% pointwise credible band on the counterfactual widens as the forecast horizon grows — a textbook property of state-space models. Extract the per-year band from `impact_full$series` and plot the band width by year. By 2000, how much wider is the band than in 1989?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-03
#| fig-width: 8
#| fig-height: 4
band_df <- impact_full$series |>
  as.data.frame() |>
  tibble::rownames_to_column("idx") |>
  mutate(year     = 1970 + as.integer(idx) - 1,
         width_95 = point.pred.upper - point.pred.lower) |>
  filter(year >= 1989) |>
  select(year, width_95)

ggplot(band_df, aes(year, width_95)) +
  geom_line(linewidth = 1.1, color = "#6a9bcc") +
  geom_point(color = "#d97757", size = 2) +
  labs(x = "Year", y = "Pointwise 95% credible band width")

tibble(year       = c(1989, 2000),
       band_width = c(band_df$width_95[band_df$year == 1989],
                      band_df$width_95[band_df$year == 2000])) |>
  gt_pretty(decimals = 2)
```

The band roughly doubles across the 12-year post-period: by 2000, the model is much less sure where the counterfactual would have been than in 1989. This is the price you pay for projecting a state-space model far from its training window, and it is the reason a *cumulative* effect plot (the bottom panel of the chapter's `plot(impact_full)`) compounds these widening pointwise bands into an even wider cumulative band.
:::

### Exercise 4: From posterior $p$ to "probability of any effect"

The chapter quotes "posterior probability of any causal effect ≈ 92%." That number does not appear directly in the `impact_full$summary` table — it is $1 - p$, where `p` is the *posterior tail probability* that the simulated counterfactual is as extreme as the observed series under the fitted model (i.e. the share of posterior draws of $\widehat{Y_{1t}(0)}$ summed over the post-period that lie on the *same side* as the observed sum). Verify this by extracting `impact_full$summary$p` and computing $1 - p$ for both horizons.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch05-04
impact_full$summary |>
  tibble::rownames_to_column("Horizon") |>
  dplyr::transmute(Horizon,
                   p                = p,
                   `1 - p (chapter)` = 1 - p) |>
  gt_pretty(decimals = 3)
```

`p` is around 0.08 for both the average and cumulative horizons; $1 - p \approx 0.92$ reproduces the chapter's "≈ 92%" headline. Two things are worth flagging: (1) `p` is a *posterior tail probability*, not a frequentist $p$-value, so it does not interchange with the placebo $p$-value from chapter 4 — the chapter-4 number characterises the *procedure* under the null, this one characterises the *parameter* given the data. (2) The same number appears for both rows because CausalImpact constructs the average and cumulative effects from the same posterior draws.
:::

### Exercise 5 (stretch): Proper multiple imputation with Rubin's-rules pooling

The chapter's $m = 1$ fill is a shortcut that the "Common pitfall" section explicitly flags as anti-conservative. Run **proper** multiple imputation with $m = 5$ random-forest completions, fit `CausalImpact` on each, and pool the average-effect point estimate and 95% interval using **Rubin's rules**. Compare the pooled interval to the chapter's $m = 1$ interval.

::: {.callout-tip collapse="true"}
## Solution

Pooling logic: with $m$ fits each giving a point estimate $\hat\theta_i$ and within-imputation SE $\sqrt{W_i}$, the pooled point estimate is $\bar\theta = \frac{1}{m}\sum \hat\theta_i$, the pooled variance is $T = \bar W + (1 + 1/m)\,B$ where $\bar W$ is the average within-variance and $B$ the between-imputation variance, and the pooled 95% interval is $\bar\theta \pm 1.96\sqrt{T}$ (the $t$-correction is small for $m=5$ and is dropped here for clarity).

```{r}
#| label: ex-ch05-05
#| message: false
#| warning: false
set.seed(123)

# 5 mice random-forest completions, each pivoted to wide.
imp_long <- mice(prop99, m = 5, method = "rf", printFlag = FALSE)

fit_one <- function(i) {
  wide_i <- complete(imp_long, i) |>
    as_tibble() |>
    pivot_wider(names_from  = state,
                values_from = c(cigsale, lnincome, beer, age15to24, retprice)) |>
    relocate(cigsale_California) |>
    select(-year)
  ci <- CausalImpact(wide_i,
                     pre.period  = pre_idx,
                     post.period = post_idx)
  s  <- ci$summary["Average", ]
  tibble(theta   = s$AbsEffect,
         se      = (s$AbsEffect.upper - s$AbsEffect.lower) / (2 * 1.96))
}

fits <- map_dfr(seq_len(5), fit_one)

theta_bar <- mean(fits$theta)
W_bar     <- mean(fits$se^2)
B         <- var(fits$theta)
T_pooled  <- W_bar + (1 + 1/5) * B
se_pooled <- sqrt(T_pooled)

tibble(estimator = c("m=1 (chapter)", "m=5 Rubin-pooled"),
       point     = c(impact_full$summary$AbsEffect[1], theta_bar),
       lower     = c(impact_full$summary$AbsEffect.lower[1],
                     theta_bar - 1.96 * se_pooled),
       upper     = c(impact_full$summary$AbsEffect.upper[1],
                     theta_bar + 1.96 * se_pooled),
       width     = c(impact_full$summary$AbsEffect.upper[1] -
                     impact_full$summary$AbsEffect.lower[1],
                     2 * 1.96 * se_pooled)) |>
  gt_pretty(decimals = 2)
```

The point estimate barely moves but the pooled interval is moderately wider than the $m = 1$ interval, because Rubin's pooling adds the between-imputation variance $B$ to the within-imputation variance $\bar W$. This widening is what "Common pitfall — Example 2" anticipated. The $m=1$ shortcut is fine for tutorial pacing; it is not fine for a published causal estimate when meaningful covariate missingness is present.
:::
