---
title: "Structural Bayesian Time Series"
---

## The BSTS / CausalImpact idea

Fit a **Bayesian structural time-series (BSTS)** model on the pre-period. Use *other states' cigarette sales* (and optionally covariates) as predictors. Project the fitted model forward as the counterfactual. The posterior over (observed − projected) gives a credible interval for the policy effect.

This is the only method in the book that delivers a *credible* interval — a direct probability statement about the parameter — rather than a frequentist confidence band or a Fisher exact $p$-value.

## The model in two pieces

The BSTS counterfactual is

$$y_{1t} = \mu_t + \beta^\top x_t + \varepsilon_t, \quad t \le t^*$$

where $\mu_t$ is a local-level trend, $x_t$ are the control-series regressors (other states' `cigsale`, plus optional covariates), and $t^*$ is the intervention date. California's outcome is *a slowly-evolving trend* **plus** *a linear combination of donor-state series* **plus** *a random error*.

```{mermaid}
flowchart TB
    subgraph "BSTS counterfactual ŷ₁ₜ"
        TREND["μₜ — local-level trend<br/>(absorbs dynamics no control can explain)"]
        REG["β·xₜ — regression on donor cigsale + covariates<br/>(borrows from donor pool)"]
        ERR["εₜ — random error"]
    end
    TREND --> Y["ŷ₁ₜ"]
    REG --> Y
    ERR --> Y
    Y --> CMP["Observed y₁ₜ − ŷ₁ₜ<br/>= policy effect (with credible band)"]

    style TREND fill:#6a9bcc,stroke:#141413,color:#fff
    style REG fill:#00d4c8,stroke:#141413,color:#141413
    style ERR fill:#7a8395,stroke:#141413,color:#fff
    style Y fill:#1f2b5e,stroke:#141413,color:#fff
    style CMP fill:#d97757,stroke:#141413,color:#fff
```

The trend $\mu_t$ absorbs the dynamics that no control series can explain; the regression term $\beta^\top x_t$ borrows information from the donor pool. After the model is fit on $t \le t^*$, it is projected forward and the posterior over $y_{1t} - \hat{y}_{1t}$ gives the credible interval for the policy effect.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(CausalImpact)
library(mice)

set.seed(42)
```

```{r}
#| label: data-load
prop99 <- read_rds("data/proposition99.rds") |> as_tibble()
```

## Imputation and wide pivot

CausalImpact wants a *wide* dataset with the treated outcome in column 1 and every control series in the remaining columns. The covariate columns have missing values (`lnincome` is missing 16% of rows, `age15to24` 32%, `beer` 55%), so we fill them with single random-forest imputation from `mice`.

```{r}
#| label: impute-and-pivot
#| message: false
#| warning: false
# Fill in missing covariate values with one round of random-forest
# multiple imputation (m = 1, method = "rf"). printFlag suppresses
# console chatter.
prop99_imputed <- prop99 |>
  mice(m = 1, method = "rf", printFlag = FALSE) |>
  complete() |>
  as_tibble()

# Pivot the long panel to wide format: one column per (variable, state)
# pair. CausalImpact requires the treated outcome in column 1, hence
# relocate(cigsale_California) and drop the year index.
prop99_wide <- prop99_imputed |>
  pivot_wider(names_from  = state,
              values_from = c(cigsale, lnincome, beer, age15to24, retprice)) |>
  relocate(cigsale_California) |>
  select(-year)

dim(prop99_wide)
```

## Fit the Bayesian structural TS model

```{r}
#| label: fit-causalimpact
#| message: false
#| warning: false
# CausalImpact takes integer row indices, not years.
# Row 1 = 1970, row 19 = 1988 (last pre-period year).
# Row 20 = 1989, row 31 = 2000 (last observed year).
pre_idx  <- c(1, 19)
post_idx <- c(20, 31)

# Reset the seed so the BSTS MCMC draws are reproducible.
set.seed(42)

impact_full <- CausalImpact(prop99_wide,
                            pre.period  = pre_idx,
                            post.period = post_idx)
summary(impact_full)
```

**Reading the output.**

- **Average ATT:** approximately $-13$ packs/capita (posterior SD ≈ 11), 95% credible interval roughly $[-32, +5.7]$.
- **Cumulative effect:** about $-154$ packs over 12 years, or roughly 16% of what would have been expected absent the policy.
- **Posterior probability of any causal effect:** ≈ 92%.

If we drop the covariates and use only other states' cigarette sales as controls, the point estimate strengthens to around $-21$ packs and the posterior probability rises to ≈ 97%. The covariates absorb some of the variation the simpler model was attributing to Proposition 99 — which can be read as either "added robustness" or "watered-down signal" depending on how much you trust the imputed beer-and-income covariates.

## The two-panel diagnostic

```{r}
#| label: fig-causalimpact
#| fig-cap: "CausalImpact two-panel: pointwise observed vs Bayesian counterfactual, and cumulative effect over time."
#| fig-width: 8
#| fig-height: 7
plot(impact_full)
```

The top panel shows the pointwise picture: observed California opens a steady gap below the Bayesian counterfactual starting in 1989, with a 95% credible band that widens as we forecast further from the training window. The bottom panel cumulates that gap over time. By 2000 the cumulative effect is roughly $-150$ packs/capita with a credible interval that includes zero only at the very upper edge.

## Common pitfall

Imputing missing covariates without thinking about the imputation model. The random-forest fill we use here is a single-imputation shortcut for tutorial speed. With multiple imputation ($m > 1$) or a different model, the estimate can move by 1–3 packs — and worse, an imputation that uses California itself to fill donor covariates would build an artificial post-period correlation that biases the result *toward* zero.

## Recap

CausalImpact lands at $-13$ to $-21$ packs depending on whether covariates are included, with a 92–97% posterior probability of a non-zero effect. It is the only method in the book that delivers a *credible* interval (a direct probability statement about the parameter), not a frequentist confidence band.

| Question | Answer |
|---|---|
| What does CausalImpact estimate? | The ATT on California, 1989–2000 |
| What is the point estimate? | Average ≈ $-13$ packs/capita; cumulative ≈ $-154$ packs over 12 years |
| What is the uncertainty quantification? | 95% Bayesian credible interval ≈ $[-32, +5.7]$ packs; posterior probability of any effect ≈ 92% |
| How is the counterfactual built? | Local-level trend + Bayesian regression on donor-state cigarette sales (+ imputed covariates) |
| What is the design-time pitfall? | Single-imputation shortcuts on covariates with high missingness; never impute donors using the treated unit |

## Further reading

- @brodersen2015inferring — the original `CausalImpact` paper.
- @brodersen-causalimpact-talk — a 50-minute walk-through of the CausalImpact intuition by the package's lead author.
- @causalimpact-pkg — package documentation and worked examples.
