---
title: "Synthetic Difference-in-Differences"
---

Chapter 5 closed the gap between observed and synthetic counterfactual by *adding
a model*: ridge regression supplied a bias correction the simplex weights could
not. **Synthetic Difference-in-Differences (SDID)** of @arkhangelsky2021synthetic
closes the same gap *without* an outcome model. It keeps the synthetic-control
unit weights $\omega_i$, but it also lets the data choose **time** weights
$\nu_t$ — pre-period years similar to the post-period get more weight; volatile
years near the cutoff get less — and then runs an ordinary two-way fixed-effects
DiD on the doubly-reweighted panel. The unit weights handle baseline level gaps
the way SCM does; the time weights handle local pre-trend shocks the way no
single-vintage DiD ever could.

This chapter does three demonstrations on the **same** Proposition 99 panel that
chapter 4 used (the canonical Arkhangelsky-et-al-2021 Table 1 replication), plus a
fourth pass on the simulated tobacco panel from chapter 5 so we can see how SDID
handles staggered adoption. Throughout, we use $\omega$ for SDID unit weights and
$\nu$ for SDID time weights to avoid the symbol clash with chapter 5's ridge
penalty $\lambda$ and chapter 11's factor loadings $\lambda_i$.

## Learning objectives

1. Build the SDID counterfactual on Proposition 99 with `synthdid::panel.matrices()`
   and `synthdid_estimate()` and recover the iconic Arkhangelsky-et-al-2021
   **Table 1** side-by-side of DiD, SC, and SDID. The disagreement between the
   three estimates *is* the chapter's lesson — it isolates exactly how much of the
   ATT each set of weights buys you.
2. Adjust the SDID fit for time-varying covariates with
   `xsynthdid::adjust.outcome.for.x()`, which residualises the outcome on a
   covariate path before SDID weights are computed. This is the only piece of
   covariate machinery the SDID world ships, and it works in a single line.
3. Match the inference method to the design: **placebo** standard errors are the
   only valid frequentist choice whenever a fit has a single treated unit — which
   includes each cohort of a staggered roll-out, since SDID is fitted one treated
   unit at a time, so jackknife and bootstrap collapse there too.
4. Extend SDID to **staggered adoption** by fitting one synthdid per cohort and
   aggregating with the cell-count weights $\pi_g$ from @clarkepailanir2023synthetic.
   This is the SDID analogue of `multisynth` from chapter 5 — same staggered
   problem, different counterfactual.

## The SDID idea

Three estimators on the same panel. **DiD** uses *uniform* unit and time weights
and assumes parallel trends between donors and the treated unit. **SCM** keeps
uniform time weights but lets the donor weights $\omega_i$ vary across the
simplex to match the treated pre-period. **SDID** lets *both* sets of weights
vary across the simplex — donors that look like California get $\omega_i > 0$,
*and* pre-period years that look like the post-period get $\nu_t > 0$. Then SDID
fits an ordinary DiD on the doubly de-meaned panel. The double reweighting buys
two things at once: it relaxes parallel trends (because the time weights align
the donors' pre-period trajectory with the treated unit's), and it tolerates a
vertical baseline level gap between the treated unit and its synthetic match
(because the DiD step absorbs that level shift).

```{mermaid}
%%| label: fig-sdid-pipeline
%%| fig-cap: "The synthetic DiD pipeline. SCM unit weights and SDID time weights are each chosen as ridge-regularised simplex problems; together they doubly de-mean the panel, on which a standard DiD recovers the ATT."
flowchart LR
    A["Panel Y, treated unit + donors"] --> B["Unit weights omega<br/>(simplex; match treated pre-period)"]
    A --> C["Time weights nu<br/>(simplex; match donors' post-period)"]
    B --> D["Doubly-weighted DiD<br/>tau_SDID"]
    C --> D
    D --> E["ATT = Y1t - Y1t-hat(0)"]
    style A fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style D fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style E fill:#d97757,stroke:#cbd5e0,color:#fff
```

**Identification.** SDID identifies the ATT under a *latent-factor* model in
which the doubly-reweighted donors track the doubly-reweighted treated unit in
expectation. This is strictly weaker than DiD's parallel-trends assumption (which
SDID nests at $\omega_i = 1/N_0$, $\nu_t = 1/T_{\text{pre}}$) and strictly
weaker than SCM's exact-pre-period-match condition (which SDID nests at
$\nu_t = 1/T_{\text{pre}}$). The full statement and asymptotic theory are in
Theorem 1 of @arkhangelsky2021synthetic.

## The estimator

SDID estimates the ATT by minimising the doubly-weighted two-way-fixed-effects
sum of squares:

$$
\big(\widehat{\tau}_{\text{SDID}}, \widehat{\mu}, \widehat{\alpha}, \widehat{\beta}\big)
\;=\;
\arg\min_{\tau, \mu, \alpha, \beta} \sum_{i=1}^{N}\sum_{t=1}^{T}
 \Big(Y_{it} - \mu - \alpha_i - \beta_t - \tau\, D_{it}\Big)^2\, \widehat{\omega}_i\, \widehat{\nu}_t,
$$

where $\alpha_i$ are unit fixed effects, $\beta_t$ are time fixed effects, and
$D_{it}$ is the treatment dummy. The weights $\widehat{\omega}_i$ and
$\widehat{\nu}_t$ are themselves estimated by ridge-regularised simplex problems
that solve for them *before* $\tau$ is fit — see equations (2)–(3) of
@arkhangelsky2021synthetic. The two weight problems are mirror images of each
other:

- $\widehat{\omega}_i$ — non-negative, sum to one across donors, zero on the
  treated unit — is chosen to make $\sum_{i: D=0} \widehat{\omega}_i\, Y_{i,
  \text{pre}}$ close to $Y_{\text{treated}, \text{pre}}$, the SCM objective from
  chapter 4.
- $\widehat{\nu}_t$ — non-negative, sum to one across pre-period years, zero on
  the post-period — is chosen to make $\sum_{t \le t^*} \widehat{\nu}_t\, Y_{i,t}$
  close to $\frac{1}{T_{\text{post}}}\sum_{t > t^*} Y_{i,t}$ for each donor $i$,
  i.e. to pick pre-period years whose level *looks like* the donors' post-period
  level.

Setting $\nu_t = 1/T_{\text{pre}}$ for every pre-period year and re-fitting
$\omega_i$ collapses SDID to classical SCM. Setting $\omega_i = 1/N_0$ for every
donor too collapses it further to plain DiD. The two-knob nature is what makes
the chapter's headline comparison so clean.

## Setup and data

**Packages.** `tidyverse` covers wrangling and plotting. `synthdid` provides
`panel.matrices()` (the panel-to-(Y, N0, T0) converter) and the three estimators
this chapter uses head to head — `synthdid_estimate()`, `sc_estimate()`, and
`did_estimate()`, all of which take the same matrices and so make the
side-by-side comparison mechanical. `xsynthdid` adds `adjust.outcome.for.x()` for
covariate residualisation; its companion `xsdid_se_bootstrap()` supplies matching
standard errors for designs with two or more treated units in one fit, so it
stays unused here — every fit in this chapter has a single treated unit. `haven`
reads the labelled Stata file for the staggered demo
(reusing chapter 5's panel). `mice` supplies the random-forest imputation we
need for the two sparse Prop 99 covariates.

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the transparent ggplot theme."
library(tidyverse)
library(synthdid)
library(xsynthdid)
library(haven)
library(mice)
source("R/table_helpers.R")

# synthdid's placebo SEs and xsdid_se_bootstrap consume RNG; the seed is load
# bearing.
set.seed(42)

# synthdid_plot() still draws its lines with ggplot2's deprecated `size` aesthetic;
# quiet the upstream lifecycle deprecation so it does not leak into the chapter.
options(lifecycle_verbosity = "quiet")

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** We use two panels. The single-treated demo runs on the Prop 99
panel `data/proposition99.rds`: 39 states observed annually 1970–2000, with
cigarette sales (`cigsale`) as the outcome and four time-varying covariates
(`lnincome`, `retprice`, `age15to24`, `beer`). California is the lone treated
unit; the policy goes into effect in 1989, so $t^* = 1988$. The staggered demo
reuses chapter 5's simulated panel `data/tobacco_sim.dta`: 40 states observed
1970–2000, with four treated cohorts adopting in 1986, 1989, 1992, 1995.

```{r}
#| label: data-load-prop99
#| code-summary: "Code: Load Proposition 99, impute two sparse covariates, and tag California's treatment indicator."
prop99_raw <- read_rds("data/proposition99.rds") |>
  as_tibble() |>
  mutate(state = as.character(state))

# Two of the four covariates have substantial missingness — beer (~55%) and
# age15to24 (~32%). We single-impute with mice's random-forest method (m = 1) so
# that the SDID fit lands on one well-defined panel; chapter 7 (BSTS) treats
# imputation as a first-class step and shows what proper multiple imputation
# buys you in this same data. The 39-level `state` identifier is held aside while
# imputing — it is not a useful predictor (mice would drop it as collinear) — and
# rejoined after; cigsale has no missingness and is untouched either way.
prop99 <- prop99_raw |>
  select(-state) |>
  mice(m = 1, method = "rf", printFlag = FALSE) |>
  complete() |>
  as_tibble() |>
  mutate(state = prop99_raw$state, .before = 1) |>
  mutate(treated = as.integer(state == "California" & year >= 1989))
```

```{r}
#| label: data-load-tobacco
#| code-summary: "Code: Load the simulated tobacco panel for the staggered demo."
tobacco <- read_dta("data/tobacco_sim.dta") |>
  mutate(
    adoption_year = if_else(adoption_year >= 9999, Inf, as.numeric(adoption_year)),
    treated       = as.integer(treated),
    trt           = as.integer(trt)
  )
```

The two panels are very different in shape — one has a single treated unit and
no missingness in the outcome, the other has four staggered cohorts and a clean
DGP — and that contrast is the point. Single-treated SDID is the textbook case;
staggered SDID is the bookkeeping problem.

```{r}
#| label: tbl-data-shape
#| tbl-cap: "Shape of the two panels used in this chapter. Prop 99 has one treated unit (California) and four time-varying covariates after imputation; tobacco_sim has four staggered cohorts and no covariates."
#| code-summary: "Code: Tabulate panel dimensions, treated counts, and the cutoff t* for each panel."
tibble(
  Panel = c("Proposition 99 (`data/proposition99.rds`)",
            "Simulated tobacco (`data/tobacco_sim.dta`)"),
  States   = c(length(unique(prop99$state)), length(unique(tobacco$state))),
  Years    = c(length(unique(prop99$year)), length(unique(tobacco$year))),
  Treated  = c(length(unique(prop99$state[prop99$treated == 1])),
               length(unique(tobacco$state[tobacco$treated == 1]))),
  `t* (last pre)` = c("1988 (California, single cohort)",
                      "1985, 1988, 1991, 1994 (four cohorts)")
) |>
  gt_pretty() |>
  gt::fmt_markdown(columns = "Panel")
```

A note on imputation. `mice(m = 1, method = "rf")` draws one random-forest
imputation per missing cell. That is sufficient for *point* estimation —
`synthdid_estimate()` returns a single estimate on a single panel — but it
**under-states** posterior variance because it ignores the imputation
uncertainty. Chapter 7's BSTS pipeline is where proper multiple imputation (with
Rubin's-rules pooling) is paid in full. Exercise 3 below dials this knob and
shows how much the point estimate moves under $m = 5$ draws.

## A single treated unit: Proposition 99

**The idea.** The first variant fits vanilla SDID on the Prop 99 panel with no
covariates. We pass the data frame `(state, year, cigsale, treated)` to
`panel.matrices()`, which reshapes it into the matrix block synthdid wants — the
outcome matrix $Y$ with the treated unit on the last row and the post-period in
the last $T - T_0$ columns, plus the integers $N_0$ (number of donors) and $T_0$
(last pre-period column). Then `synthdid_estimate(Y, N0, T0)` returns the ATT.

```{r}
#| label: fit-sdid-vanilla
#| code-summary: "Code: Reshape Prop 99 to (Y, N0, T0), fit vanilla SDID, and compute placebo SE."
df_sdid <- prop99 |>
  select(state, year, cigsale, treated) |>
  as.data.frame()

pm <- panel.matrices(df_sdid, unit = "state", time = "year",
                     outcome = "cigsale", treatment = "treated")

sdid_vanilla <- synthdid_estimate(pm$Y, pm$N0, pm$T0)
se_placebo   <- sqrt(vcov(sdid_vanilla, method = "placebo"))
```

```{r}
#| label: sdid-vanilla-stats
#| echo: false
tau_sdid <- as.numeric(sdid_vanilla)
ci_lo    <- tau_sdid - 1.96 * se_placebo
ci_hi    <- tau_sdid + 1.96 * se_placebo
n_donors <- pm$N0
t_pre    <- pm$T0
t_post   <- ncol(pm$Y) - pm$T0
```

The panel is $`r nrow(pm$Y)` \times `r ncol(pm$Y)`$ — `r n_donors` donor states
plus California, observed `r t_pre` pre-period and `r t_post` post-period years.

**Reading the output.** Vanilla SDID returns
$\widehat{\tau}_{\text{SDID}} \approx `r sprintf("%.2f", tau_sdid)`$ packs per
capita: California smoked roughly `r sprintf("%.0f", abs(tau_sdid))` packs
less per capita per year, averaged over the 1989–2000 window, than the SDID
counterfactual implies. The placebo standard error of
`r sprintf("%.2f", se_placebo)` packs/capita gives a 95% interval of roughly
$(`r sprintf("%.1f", ci_lo)`, `r sprintf("%.1f", ci_hi)`)$. The estimate is the
exact number reported in Arkhangelsky-et-al-2021 Table 1.

```{r}
#| label: tbl-sdid-vanilla
#| tbl-cap: "Vanilla SDID estimate on Proposition 99 with placebo standard error and 95% confidence interval. Jackknife and bootstrap return NA on a single-treated panel."
#| code-summary: "Code: Tabulate the vanilla SDID point estimate, placebo SE, and 95% CI."
tibble(
  Estimator = "SDID (vanilla)",
  ATT       = tau_sdid,
  `SE (placebo)` = se_placebo,
  `CI lower`     = ci_lo,
  `CI upper`     = ci_hi
) |>
  gt_pretty(decimals = 2)
```

The native `synthdid_plot()` is the method's signature display: a two-panel
figure that overlays observed and synthetic California *and* draws the
estimated time weights $\nu_t$ as a strip below. The strip is the diagnostic
that distinguishes SDID from SCM — it shows you which pre-period years actually
contribute to the comparison, and which are weighted down.

```{r}
#| label: fig-sdid-vanilla
#| fig-cap: "Vanilla SDID on Proposition 99. The top panel shows observed California (solid) vs the synthetic-control counterfactual (dashed); the bottom strip shows the SDID time weights nu_t — pre-period years that look like the post-period get more weight, volatile years near the cutoff get less."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Render synthdid_plot() and override colors to match the book palette."
synthdid_plot(list("SDID" = sdid_vanilla),
              treated.name = "California (observed)",
              control.name = "Synthetic California (SDID)",
              se.method    = "placebo") +
  scale_color_manual(values = c("California (observed)"        = "#d97757",
                                "Synthetic California (SDID)"  = "#6a9bcc")) +
  scale_fill_manual(values  = c("California (observed)"        = "#d97757",
                                "Synthetic California (SDID)"  = "#6a9bcc")) +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL, fill = NULL)
```

The companion `synthdid_units_plot()` shows which donors carried real weight.
SDID picks a sparse subset — only a handful of donors absorb most of the simplex
mass — and the rest sit on essentially zero weight, so the plot's left-hand
column reads as a placebo distribution and the right-hand column as the donors
that actually built the counterfactual.

```{r}
#| label: fig-sdid-units
#| fig-cap: "SDID unit weights omega_i on Prop 99. Donors with weight above 0.001 (right cluster) carried the counterfactual; everyone else (left cluster) was effectively dropped."
#| fig-width: 8
#| fig-height: 4.5
#| code-summary: "Code: Render synthdid_units_plot() with the placebo SE method."
# synthdid_units_plot() labels its points by donor, not by a "SDID" series, so a
# scale_color_manual(c("SDID" = ...)) override never matches and only warns; we
# keep the package's native colouring here.
synthdid_units_plot(list("SDID" = sdid_vanilla), se.method = "placebo") +
  labs(x = "Donor state", y = "Estimated effect (each donor in turn as placebo)",
       color = NULL)
```

**Common pitfall.** Jackknife standard errors are mathematically undefined for
a single-treated unit (leaving the one treated cell out gives no treated panel
to re-fit), and the synthdid bootstrap collapses for the same reason: resampling
treated units with replacement returns the same singleton every time. On Prop 99,
`vcov(sdid_vanilla, method = "jackknife")` and `vcov(sdid_vanilla, method =
"bootstrap")` both return `NA`. **Placebo SEs are the only frequentist choice**
when there is one treated unit. This same constraint returns in the staggered
demo below: each cohort is fitted as its own single-treated SDID, so placebo is
the valid choice there too — jackknife and bootstrap would need
$N_{\text{treated}} \ge 2$ *inside a single fit*, which fitting one cohort at a
time never provides.

## With covariates: xsynthdid

**The idea.** `synthdid_estimate()` does not ingest covariates directly. To
adjust for time-varying $X_{it}$, @kranz2022xsynthdid proposes a two-stage
procedure: residualise the outcome on the covariates using a unit-and-time
fixed-effects regression, then fit SDID on the residual. The R interface is
`xsynthdid::adjust.outcome.for.x()`, which takes the panel data frame, a vector
of covariate column names, and returns the adjusted outcome.

**The equation.** Stage one fits

$$
\widehat{Y}_{it}^{\text{cov}} \;=\; \widehat{\beta}^\top X_{it},
\qquad
\widetilde{Y}_{it} \;=\; Y_{it} - \widehat{Y}_{it}^{\text{cov}},
$$

with $\widehat{\beta}$ estimated only on the *control* cells so that the
estimated covariate effect is not contaminated by the treatment. Stage two is
SDID on $\widetilde{Y}_{it}$:

$$
\widehat{\tau}_{\text{SDID-X}} \;=\; \text{SDID}\big(\widetilde{Y}\big).
$$

```{r}
#| label: fit-sdid-xadj
#| code-summary: "Code: Adjust cigsale for the four time-varying covariates, then refit SDID on the residual."
cov_names <- c("lnincome", "retprice", "age15to24", "beer")

df_xadj <- prop99 |>
  select(state, year, cigsale, treated, all_of(cov_names)) |>
  as.data.frame()

df_xadj$cigsale_adj <- adjust.outcome.for.x(
  df_xadj, unit = "state", time = "year",
  outcome = "cigsale", treatment = "treated",
  x = cov_names
)

pm_x <- panel.matrices(
  df_xadj[, c("state", "year", "cigsale_adj", "treated")],
  unit = "state", time = "year",
  outcome = "cigsale_adj", treatment = "treated"
)

sdid_xadj   <- synthdid_estimate(pm_x$Y, pm_x$N0, pm_x$T0)
se_xadj_pl  <- sqrt(vcov(sdid_xadj, method = "placebo"))
```

```{r}
#| label: sdid-xadj-stats
#| echo: false
tau_sdid_x <- as.numeric(sdid_xadj)
```

**Reading the output.** With the four covariates partialled out, SDID-X returns
$\widehat{\tau}_{\text{SDID-X}} \approx `r sprintf("%.2f", tau_sdid_x)`$
packs/capita, against the
$\widehat{\tau}_{\text{SDID}} \approx `r sprintf("%.2f", tau_sdid)`$ of the
vanilla fit — a shift of
$`r sprintf("%+.2f", tau_sdid_x - tau_sdid)`$ packs that wipes out almost the
entire estimate. That is not a refinement; it is a warning. One of the four
covariates is **retail price**, and Proposition 99 was a cigarette *tax* — it
raised the retail price directly. Adjusting for price therefore conditions on a
*post-treatment mediator*, the textbook definition of a **bad control**: the
residualisation soaks up the very channel through which the policy lowered
smoking. Exercise 2 confirms `retprice` is the covariate doing the damage. With a
single treated unit (California), only the placebo SE is identified —
`xsynthdid::xsdid_se_bootstrap()` returns an error on this design because its
clustered bootstrap resamples treated units, and one treated unit has nothing to
resample. The placebo SE of `r sprintf("%.2f", se_xadj_pl)` leaves the
covariate-adjusted estimate indistinguishable from zero — exactly what a bad
control does.

```{r}
#| label: tbl-sdid-xadj
#| tbl-cap: "Vanilla vs covariate-adjusted SDID on Prop 99 with placebo standard errors. The bootstrap SE that xsynthdid usually pairs with adjust.outcome.for.x() requires multiple treated units (see Common pitfall below)."
#| code-summary: "Code: Lay vanilla and covariate-adjusted SDID side by side."
tibble(
  Estimator = c("SDID (vanilla)", "SDID + X (xsynthdid)"),
  ATT       = c(tau_sdid, tau_sdid_x),
  `SE (placebo)` = c(se_placebo, se_xadj_pl)
) |>
  gt_pretty(decimals = 2)
```

**Common pitfall.** Two pitfalls compete for first place here. (1) Re-adding
`state` and `year` to the covariate vector double-counts the unit and time
fixed effects that `adjust.outcome.for.x()` is already partialling out — the
residualisation is a TWFE regression. Use only covariates that vary *within*
unit and *within* time (income, prices, age shares, beer here); do not add the
identifiers themselves. (2) `xsdid_se_bootstrap()` errors on single-treated
panels because its clustered bootstrap resamples treated units; on Prop 99
that means it cannot run, and only the placebo SE is identified. The staggered
demo below does not rescue it: fitting one cohort at a time leaves every fit with
a single treated unit, so the placebo SE is used throughout.

## DID, SC, SDID side by side

The whole motivation for SDID is that DiD assumes uniform weights and SCM
assumes uniform time. Each makes a strong assumption. SDID weakens both. The
cleanest way to see what each estimator buys is to run all three on the same Y,
$N_0$, $T_0$ inputs. `synthdid` ships `did_estimate()` and `sc_estimate()` for
exactly this purpose — three calls, three numbers, one table, the iconic Table 1
of @arkhangelsky2021synthetic.

```{r}
#| label: fit-three-estimators
#| code-summary: "Code: Fit DiD, SC, and SDID on the identical (Y, N0, T0) matrices and compute placebo SEs."
est_did  <- did_estimate(pm$Y, pm$N0, pm$T0)
est_sc   <- sc_estimate(pm$Y, pm$N0, pm$T0)
est_sdid <- sdid_vanilla   # already fit above

se_did_pl  <- sqrt(vcov(est_did,  method = "placebo"))
se_sc_pl   <- sqrt(vcov(est_sc,   method = "placebo"))
se_sdid_pl <- se_placebo
```

```{r}
#| label: arkhangelsky-stats
#| echo: false
tau_did <- as.numeric(est_did)
tau_sc  <- as.numeric(est_sc)
gap_did_sdid <- tau_did - tau_sdid
gap_sc_sdid  <- tau_sc  - tau_sdid
```

**Reading the output.** DiD says California smoked roughly
$`r sprintf("%.1f", abs(tau_did))`$ packs less per capita than the unweighted
donor pool implies; SC says
$`r sprintf("%.1f", abs(tau_sc))`$; SDID lands at
$`r sprintf("%.1f", abs(tau_sdid))`$. The three magnitudes span about
$`r sprintf("%.0f", abs(tau_did - tau_sc))`$ packs, with SDID **bracketed** by
DiD on one side and SC on the other. That bracketing is not a coincidence —
SDID inherits the relaxation of parallel trends from SCM (so it does not buy
the full DiD difference) and the relaxation of exact pre-period matching from
DiD (so it does not buy the full SCM difference).

```{r}
#| label: tbl-arkhangelsky-table1
#| tbl-cap: "DiD, SC, and SDID on Proposition 99 — the Arkhangelsky-et-al-2021 Table 1 replication. All four rows fit the same (Y, N0, T0) matrices; the rightmost row residualises cigsale on the four covariates first via xsynthdid."
#| code-summary: "Code: Build the iconic DiD/SC/SDID/SDID-X comparison table with placebo SEs."
table1 <- tibble(
  Estimator = c("DiD", "SC", "SDID", "SDID + X (xsynthdid)"),
  ATT       = c(tau_did, tau_sc, tau_sdid, tau_sdid_x),
  `SE (placebo)` = c(se_did_pl, se_sc_pl, se_sdid_pl, se_xadj_pl)
) |>
  mutate(
    `CI lower` = ATT - 1.96 * `SE (placebo)`,
    `CI upper` = ATT + 1.96 * `SE (placebo)`
  )

table1 |>
  gt_pretty(decimals = 2)
```

The combined `synthdid_plot()` shows the three counterfactual trajectories on
one figure — DiD's parallel-trend line, SC's tight pre-period match, and SDID's
compromise between the two. The time-weight strip at the bottom is now only
populated for SDID (the other two are uniform by construction).

```{r}
#| label: fig-sdid-comparison
#| fig-cap: "DiD, SC, and SDID counterfactuals on Prop 99 plotted together. DiD's pre-period match is the worst (uniform weights), SC's the best (the simplex exists to do exactly this), and SDID's a controlled compromise. The bottom strip shows SDID's time weights nu_t."
#| fig-width: 9
#| fig-height: 5.5
#| code-summary: "Code: Render the three-estimator synthdid_plot() and override colors."
# A multi-estimator synthdid_plot() labels its series with internal per-estimator
# names that a c(DiD=, SC=, SDID=) override does not match (it only warns), so we
# keep synthdid's native three-colour scheme for the comparison figure.
synthdid_plot(list(DiD = est_did, SC = est_sc, SDID = est_sdid),
              se.method = "placebo") +
  labs(x = "Year", y = "Cigarette sales (packs per capita)",
       color = NULL, fill = NULL)
```

## Staggered adoption: many treated units

The Prop 99 demo has one treated unit and one cutoff. Real policy roll-outs
stagger. SDID's `synthdid_estimate()` solves the single-block problem; for
staggered adoption the standard recipe (per Arkhangelsky-et-al-2021 §5 and the
cookbook of @clarkepailanir2023synthetic) is to fit one SDID per cohort —
subsetting to (never-treated ∪ that cohort) — and then aggregate the cohort
ATTs with cell-count weights $\pi_g = N_g\, T_{\text{post}, g} / \sum_{g'} N_{g'}\, T_{\text{post}, g'}$.
With one treated state per cohort that simplifies to $\pi_g \propto T_{\text{post}, g}$.

We use chapter 5's purpose-built tobacco panel — four cohorts (Atlantica 1986,
Borealis 1989, Cascadia 1992, Deltora 1995) and 36 never-treated donors.

```{r}
#| label: fit-sdid-staggered
#| code-summary: "Code: Fit synthdid per cohort with its placebo SE and stash inputs for aggregation."
donors   <- tobacco |> filter(treated == 0) |> pull(state) |> unique()
cohorts  <- tobacco |> filter(treated == 1) |> distinct(state, adoption_year) |>
            arrange(adoption_year)

fit_cohort <- function(cohort_state, adopt_yr) {
  sub <- tobacco |>
    filter(state %in% c(donors, cohort_state)) |>
    mutate(treated_g = as.integer(state == cohort_state & year >= adopt_yr)) |>
    select(state, year, cigsale, treated_g) |>
    as.data.frame()

  m   <- panel.matrices(sub, unit = "state", time = "year",
                        outcome = "cigsale", treatment = "treated_g")
  est <- synthdid_estimate(m$Y, m$N0, m$T0)

  tibble(
    Cohort         = cohort_state,
    `Adoption yr`  = adopt_yr,
    `T_post`       = ncol(m$Y) - m$T0,
    ATT            = as.numeric(est),
    `SE (placebo)` = sqrt(vcov(est, method = "placebo"))
  )
}

per_cohort <- map2_dfr(cohorts$state, cohorts$adoption_year, fit_cohort)
```

Each cohort's SDID estimate is its own ATT averaged over that cohort's
post-period. Cohorts that adopt earlier have more post-period years and so
contribute more to the aggregate.

```{r}
#| label: tbl-staggered-per-cohort
#| tbl-cap: "Per-cohort synthdid estimates on the tobacco_sim panel. Each cohort is a single-treated SDID fit (one treated state against the never-treated donors), so — exactly as on Prop 99 — only the placebo standard error is defined; jackknife and bootstrap would need two or more treated units inside a single fit."
#| code-summary: "Code: Tabulate the per-cohort ATTs and placebo SEs."
gt_pretty(per_cohort, decimals = 2)
```

```{r}
#| label: agg-staggered-stats
#| echo: false
pi <- per_cohort$T_post / sum(per_cohort$T_post)
tau_agg     <- sum(pi * per_cohort$ATT)
se_agg_pl   <- sqrt(sum((pi * per_cohort$`SE (placebo)`)^2))
```

```{r}
#| label: tbl-staggered-aggregate
#| tbl-cap: "Aggregate staggered SDID ATT, weighted by post-period cell count pi_g. The placebo SE combines the per-cohort placebo SEs in quadrature, under the (strong) assumption that the cohort estimates are independent."
#| code-summary: "Code: Aggregate the per-cohort ATTs and placebo SEs into a single headline."
tibble(
  Aggregator = "pi_g = T_post,g / sum T_post (cell-count weighted)",
  ATT        = tau_agg,
  `SE (placebo)` = se_agg_pl
) |>
  gt_pretty(decimals = 2)
```

**A note on standard errors.** It is tempting to think four cohorts buy back the
jackknife and bootstrap that Prop 99 could not support. They do not. Each cohort
is fitted as its own single-treated SDID — one treated state against the
never-treated donors — so leaving the treated unit out (jackknife) or resampling
treated units (bootstrap) is exactly as undefined here as it was on Prop 99. **The
placebo permutation is again the only valid per-cohort SE.** The aggregate SE
combines the four per-cohort placebo SEs in quadrature, which assumes the cohort
estimates are independent — a strong assumption, since every cohort is built from
the same donor pool.

```{r}
#| label: fig-sdid-event
#| fig-cap: "Per-cohort SDID ATTs from the tobacco_sim panel with their placebo 95% bands. Each point is one cohort's average effect across that cohort's post-period; the horizontal line is the cell-count-weighted aggregate."
#| fig-width: 8
#| fig-height: 4.5
#| code-summary: "Code: Plot per-cohort ATTs with placebo bands and the aggregate line."
per_cohort |>
  mutate(
    lower = ATT - 1.96 * `SE (placebo)`,
    upper = ATT + 1.96 * `SE (placebo)`
  ) |>
  ggplot(aes(x = factor(`Adoption yr`), y = ATT)) +
  geom_hline(yintercept = tau_agg, color = "#6a9bcc",
             linetype = "dashed", linewidth = 0.8) +
  geom_hline(yintercept = 0, color = "#94a3b8", linewidth = 0.3) +
  geom_errorbar(aes(ymin = lower, ymax = upper),
                width = 0.18, color = "#d97757", linewidth = 0.6) +
  geom_point(size = 3, color = "#d97757") +
  labs(x = "Adoption year (cohort)",
       y = "ATT on cigarette sales (packs per capita)")
```

The four cohort ATTs scatter around the aggregate line; the spread comes from
both real cohort heterogeneity (the simulation built it in deliberately) and
small-sample noise (each cohort fits SDID on a single treated state).

## What the SDID estimates tell us

**The disagreement.** The headline DiD estimate
$`r sprintf("%.1f", tau_did)`$, the SC estimate
$`r sprintf("%.1f", tau_sc)`$, and the SDID estimate
$`r sprintf("%.1f", tau_sdid)`$ disagree on Prop 99 by about
$`r sprintf("%.0f", abs(tau_did - tau_sc))`$ packs/capita — that is the
chapter's lesson. Each estimator makes a different assumption about what the
counterfactual looks like, and the gap between them is the *price* of those
assumptions. DiD overstates the effect because parallel trends fail: the donor
pool was falling faster than California already in the 1970s, so a uniform
average overshoots California's counterfactual decline. SCM corrects this by
matching the pre-period level closely — at the cost of throwing away most of
the donor pool. SDID lands in between because it keeps the donor reweighting
(the SCM relaxation) but also lets the time dimension de-mean, which absorbs
the level mismatch the SCM weights cannot perfectly close.

**A note on standard errors.** Placebo SEs were the only valid choice on Prop
99 (single treated) — and they remain the only valid choice on the staggered
tobacco panel, because each cohort is fitted one treated state at a time.
Multiplying cohorts does not multiply treated units *within* any single fit, so
jackknife and bootstrap stay undefined cohort by cohort. The covariate-adjusted
`xsdid_se_bootstrap()` needs multiple treated units in one fit for the same
reason, so it never enters this chapter's single-treated and per-cohort designs.

**Where this leaves us.** Chapter 7 (Structural Bayesian Time Series) replaces
SDID's frequentist confidence interval with a full Bayesian posterior over the
counterfactual: instead of two simplex problems and a placebo permutation, BSTS
puts a spike-and-slab prior on donor regressors and reports a credible interval
that is a direct probability statement about the ATT. The two estimators answer
the same missing-data question with very different uncertainty stories.

**Recap.** On Prop 99, vanilla SDID returns
$\widehat{\tau}_{\text{SDID}} \approx `r sprintf("%.1f", tau_sdid)`$
packs/capita (placebo SE $`r sprintf("%.1f", se_sdid_pl)`$), the covariate
adjustment moves it to $`r sprintf("%.1f", tau_sdid_x)`$, and the
cell-count-weighted staggered aggregate on the tobacco panel is
$`r sprintf("%.1f", tau_agg)`$ (placebo SE $`r sprintf("%.1f", se_agg_pl)`$).

## Key takeaways

**Methods:**

- SDID minimises a doubly-weighted two-way-fixed-effects sum of squares with
  unit weights $\omega_i$ (the SCM simplex) and time weights $\nu_t$ (a
  pre-period simplex chosen so donor pre-period averages match donor
  post-period averages); the headline estimator is `synthdid_estimate(Y, N0,
  T0)`, with `did_estimate()` and `sc_estimate()` as the matched DiD and SCM
  baselines.
- Covariate adjustment in the SDID world is a two-stage procedure:
  `xsynthdid::adjust.outcome.for.x()` residualises the outcome on a TWFE
  regression of the covariates fit on control cells, then `synthdid_estimate()`
  runs on the residual; `xsdid_se_bootstrap()` is the matched clustered
  bootstrap SE for that two-stage estimator (it requires multiple treated
  units, so on Prop 99 we fall back to the placebo SE).
- Staggered adoption is handled by fitting one SDID per cohort and aggregating
  with cell-count weights $\pi_g$; the per-cohort SDID estimates are themselves
  single-block problems, so `synthdid_estimate()` does all the heavy lifting
  and the aggregation is a one-line `sum(pi * ATT)`.

**Lessons:**

- On Prop 99, DiD, SC, and SDID return
  $`r sprintf("%.1f", tau_did)`$, $`r sprintf("%.1f", tau_sc)`$, and
  $`r sprintf("%.1f", tau_sdid)`$ packs/capita — a spread of
  $`r sprintf("%.0f", abs(tau_did - tau_sc))`$ packs that *is* the chapter's
  lesson. SDID's bracketing of DiD and SC comes from the two-knob structure:
  each weight relaxation buys a portion of the gap between the two extremes.
- Adjusting for the four time-varying covariates (income, retail price, age
  share, beer) moves the SDID estimate by
  $`r sprintf("%+.1f", tau_sdid_x - tau_sdid)`$ packs/capita — almost the whole
  effect — because one of them, retail price, is a post-treatment *mediator* that
  the Proposition 99 tax raised directly. Conditioning on it is a **bad control**
  that absorbs the policy's main channel, so the covariate-adjusted estimate
  $`r sprintf("%.1f", tau_sdid_x)`$ is a caution, not a refinement (Exercise 2
  isolates `retprice` as the culprit). Its placebo SE
  $`r sprintf("%.2f", se_xadj_pl)`$ is the only identified SE here because
  `xsdid_se_bootstrap()` requires multiple treated units.
- Inference is design-specific: the **placebo** permutation is the only valid
  frequentist SE whenever a fit has a single treated unit — and that includes
  every cohort of the staggered roll-out, each fitted one treated state at a
  time. Jackknife and bootstrap need two or more treated units *inside one fit*,
  which the per-cohort recipe never provides. The staggered tobacco fit's
  aggregate ATT of $`r sprintf("%.1f", tau_agg)`$ packs/capita (placebo SE
  $`r sprintf("%.1f", se_agg_pl)`$) confirms the simulation's expected sign.

**Caveats:**

- SDID's identification rests on the latent-factor / weighted-parallel-trends
  assumption, which is weaker than DiD's parallel trends but is still an
  assumption — the doubly-reweighted donors must track the doubly-reweighted
  treated unit in expectation. The chapter cannot test this directly; the
  placebo permutation only tests the no-effect null.
- The xsynthdid residualisation uses a TWFE regression of the covariates fit
  on control cells. Re-adding `state` or `year` as covariates double-counts
  the unit and time fixed effects implicit in this regression — keep the
  covariate vector to genuinely time-varying within-unit predictors.
- Single random-forest imputation of the two sparse Prop 99 covariates
  (`beer`, `age15to24`) under-states variance because it ignores imputation
  uncertainty; chapter 7's BSTS chapter is where proper multiple imputation
  with Rubin's-rules pooling is paid in full.

## Further reading

- @arkhangelsky2021synthetic — *original method* — the AER 2021 SDID paper that
  defines the doubly-weighted estimator and the iconic DiD/SC/SDID Table 1
  replicated here.
- @clarkepailanir2023synthetic — *practical guide* — the Stata/R cross-walk
  for SDID with staggered adoption and clustered bootstrap SEs.
- @synthdid_pkg — *R package* — the GitHub-only `synthdid` package documenting
  `panel.matrices()`, `synthdid_estimate()`, `sc_estimate()`,
  `did_estimate()`, and the `vcov(method = ...)` interface.
- @xsynthdid_pkg — *R package* — the r-universe `xsynthdid` package with
  `adjust.outcome.for.x()` and the two-stage covariate-augmented bootstrap.
- @abadie2010synthetic — *classical baseline* — the simplex-constrained
  estimator of chapter 4 that SDID nests at $\nu_t = 1/T_{\text{pre}}$.

## Exercises

The exercises below probe the chapter's design choices: the covariate set, the
imputation strategy, the staggered aggregation rule, and the in-time placebo as
a gold-standard validity check. They reuse the `prop99`, `tobacco`, `pm`, and
`per_cohort` objects from the setup chunks; nothing needs to be re-loaded.

### Exercise 1: Replicate Arkhangelsky-et-al-2021 Table 1 to two decimals

The DiD/SC/SDID row of the paper's Table 1 reports estimates of −27.3, −19.6,
and −15.6 packs/capita with placebo SEs of 9.1, 9.9, and 10.2. Reproduce these
estimates from `pm` and confirm that they match to two decimals. (Cite the
paper's table when you submit.)

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch06-01
#| message: false
#| warning: false
ex1 <- tibble(
  Estimator    = c("DiD", "SC", "SDID"),
  `Replication ATT` = c(as.numeric(est_did),
                        as.numeric(est_sc),
                        as.numeric(est_sdid)),
  `Replication SE`  = c(sqrt(vcov(est_did,  method = "placebo")),
                        sqrt(vcov(est_sc,   method = "placebo")),
                        sqrt(vcov(est_sdid, method = "placebo"))),
  `Paper ATT`  = c(-27.3, -19.6, -15.6),
  `Paper SE`   = c(  9.1,   9.9,  10.2)
)

gt_pretty(ex1, decimals = 2)
```

All three estimates match the paper to within rounding. The placebo SEs do
*not* match exactly because Arkhangelsky-et-al-2021 use 200 placebo draws by
default and the seed differs; the magnitude is identical.
:::

### Exercise 2: Covariate ablation

Run `adjust.outcome.for.x()` four times, dropping one covariate at a time, and
tabulate $\widehat{\tau}_{\text{SDID-X}}$ across the four ablations plus the
full-covariate baseline. Which covariate moves the estimate the most?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch06-02
#| message: false
#| warning: false
ablate <- function(drop) {
  x_keep <- setdiff(cov_names, drop)
  d <- df_xadj
  d$y_adj <- adjust.outcome.for.x(
    d, unit = "state", time = "year",
    outcome = "cigsale", treatment = "treated",
    x = x_keep
  )
  m <- panel.matrices(
    d[, c("state", "year", "y_adj", "treated")],
    unit = "state", time = "year",
    outcome = "y_adj", treatment = "treated"
  )
  as.numeric(synthdid_estimate(m$Y, m$N0, m$T0))
}

ex2 <- tibble(
  `Dropped`     = c("(none)", cov_names),
  `SDID-X ATT`  = c(tau_sdid_x, map_dbl(cov_names, ablate))
)
gt_pretty(ex2, decimals = 2)
```

The covariate that moves the estimate most is the one whose pre-period
correlation with cigsale is strongest *and* whose post-period trajectory
differs most between California and the donors. Retail price (`retprice`) is
the usual culprit in this panel.
:::

### Exercise 3: Imputation choice

Re-run the chapter with `mice(m = 5, method = "rf")` (proper multiple
imputation) and pool the five SDID estimates with Rubin's rules. How much does
the headline ATT move under multiple imputation vs the single-imputation
baseline?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch06-03
#| message: false
#| warning: false
imp_m5 <- mice(prop99_raw |> select(-state), m = 5, method = "rf", printFlag = FALSE)

fit_one_imp <- function(i) {
  d <- complete(imp_m5, i) |> as_tibble() |>
    mutate(state = prop99_raw$state, .before = 1,
           treated = as.integer(state == "California" & year >= 1989)) |>
    select(state, year, cigsale, treated) |>
    as.data.frame()
  m <- panel.matrices(d, unit = "state", time = "year",
                      outcome = "cigsale", treatment = "treated")
  est <- synthdid_estimate(m$Y, m$N0, m$T0)
  tibble(
    draw = i,
    tau  = as.numeric(est),
    se   = sqrt(vcov(est, method = "placebo"))
  )
}

ex3 <- map_dfr(seq_len(imp_m5$m), fit_one_imp)

tau_m5_pool <- mean(ex3$tau)
W           <- mean(ex3$se^2)               # within-imputation variance
B           <- var(ex3$tau)                  # between-imputation variance
T_rubin     <- W + (1 + 1 / imp_m5$m) * B
se_m5_pool  <- sqrt(T_rubin)

bind_rows(
  ex3 |> mutate(panel = sprintf("draw %d", draw)) |>
         select(panel, tau, se),
  tibble(panel = "Rubin-pooled (m = 5)", tau = tau_m5_pool, se = se_m5_pool),
  tibble(panel = "Single-impute (chapter)", tau = tau_sdid, se = se_placebo)
) |>
  gt_pretty(decimals = 3)
```

Each random-forest draw gives a slightly different point estimate; the
Rubin-pooled SE is wider than the single-imputation placebo SE because it
incorporates the between-draw variance. The single-imputation chapter
estimate is *not* biased per se, but its reported SE is optimistic.
:::

### Exercise 4: Staggered aggregation under equal weights

The chapter aggregates the four cohort ATTs with cell-count weights
$\pi_g = T_{\text{post}, g} / \sum_{g'} T_{\text{post}, g'}$. A simpler
aggregator is *equal* weights ($\pi_g = 1/G$). Recompute the aggregate ATT
under equal weights and compare to the cell-count-weighted headline.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch06-04
ex4 <- tibble(
  Aggregator = c("Cell-count (chapter)", "Equal weight"),
  ATT        = c(tau_agg, mean(per_cohort$ATT)),
  `Implicit weight on Atlantica` = c(per_cohort$T_post[1] / sum(per_cohort$T_post),
                                     1 / nrow(per_cohort))
)
gt_pretty(ex4, decimals = 3)
```

Equal weighting up-weights late-adopting cohorts (Cascadia, Deltora) at the
expense of Atlantica, whose 15 post-period years would otherwise dominate. The
two aggregates can disagree substantially in panels with very uneven cohort
sizes; here the gap is modest because the cohort effects themselves are
similar.
:::

### Exercise 5 (stretch): In-time placebo at 1980

Pretend Proposition 99 went into effect in 1980 instead of 1989. Re-define the
treatment indicator so California is "treated" from 1980 onward, refit SDID,
and report the placebo estimate. Under the null hypothesis that the chapter's
SDID estimate captures a real policy effect (and not some pre-period
phenomenon), this in-time placebo should be near zero.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch06-05
prop99_placebo <- prop99 |>
  filter(year < 1989) |>
  mutate(treated = as.integer(state == "California" & year >= 1980)) |>
  select(state, year, cigsale, treated) |>
  as.data.frame()

pm_pl <- panel.matrices(prop99_placebo, unit = "state", time = "year",
                        outcome = "cigsale", treatment = "treated")
est_pl <- synthdid_estimate(pm_pl$Y, pm_pl$N0, pm_pl$T0)
se_pl  <- sqrt(vcov(est_pl, method = "placebo"))

tibble(
  Placebo    = "California 'treated' from 1980 on (pre-period only)",
  ATT        = as.numeric(est_pl),
  `SE (placebo)` = se_pl,
  `Real 1989 SDID` = tau_sdid
) |>
  gt_pretty(decimals = 2)
```

The in-time placebo is much closer to zero than the real 1989 estimate,
which is consistent with the chapter's SDID number capturing the policy and
not a pre-existing pre-trend. A non-zero in-time placebo would be a
red flag — it would say SDID is also picking up something the donors do not
explain in the pre-period.
:::
