---
title: "Structural Bayesian Time Series"
---

## Overview

*Placeholder.* Structural Bayesian time-series models (e.g., Google's `CausalImpact`) estimate the causal effect of an intervention as the difference between observed post-intervention outcomes and a Bayesian counterfactual forecast built from pre-intervention dynamics and contemporaneous control series.

## The model

*Placeholder.* A state-space decomposition into local-trend, seasonal, and regression components, fit by `bsts`.

## Estimation in R

```{r}
#| label: setup
#| message: false
#| warning: false
#| eval: false
library(tidyverse)
library(CausalImpact)
library(bsts)
```

```{r}
#| label: ci-example
#| eval: false
# Placeholder: build a CausalImpact input frame with the treated series
# and donor controls, call CausalImpact(), summarize and plot.
```

## Interpreting posterior intervals

*Placeholder.*

## Regional case study

*Placeholder.*

## Comparison with synthetic control

*Placeholder.* CausalImpact and classical synthetic control answer related questions with different assumptions; this section compares estimates on the same data.

## Further reading

*Placeholder.*
