---
title: "Bayesian Spatial Synthetic Control"
---

## Learning objectives

1. Fit a Bayesian SCM with horseshoe priors using the bundled `R/scspill/` Gibbs sampler. Horseshoe relaxes the simplex constraint, and the chapter shows the simplex's four-donor sparsity is partly a constraint artifact — 23 donors load under the horseshoe.
2. Propagate donor-weight posterior uncertainty into credible intervals on the post-period gap series. Honest uncertainty on the ATT requires propagating uncertainty about the donor weights themselves, not just the noise around the synthetic counterfactual.
3. Add a spatial-autoregressive (SAR) layer so donor outcomes can respond to the treatment, and estimate per-donor spillovers via the Sakaguchi-Tagawa identification formula. This is the only method in the book that relaxes SUTVA and recovers a closed-form spillover for each neighbour — essential when the treated unit shares a border with several controls.
4. Run prior predictive checks on summary statistics (mean, spatial structure, autocorrelation, factor dominance) to confirm the SAR prior is compatible with the observed data. Prior predictive checks are the safeguard against a posterior driven by a misspecified prior rather than the data.

## Why a fourth synthetic-control chapter?

Chapter 4 fit a *classical* Synthetic Control: donor weights live on the simplex and are chosen by a quadratic optimiser. Chapter 5 borrowed donor information in a different way — through a Bayesian structural time-series model. Chapter 6 then put a finite-sample frequentist prediction interval around the classical fit and relaxed the simplex to lasso / ridge / OLS — but, like chapters 4 and 5, retained SUTVA. All three treat the donor states' outcomes as *unaffected* by California's policy. That assumption — the **stable unit treatment value assumption (SUTVA)** — is the price of any synthetic-control estimate.

For tobacco, SUTVA is empirically suspect. When California raised retail prices in 1989, cross-border-shopping into Nevada changed cigarette sales *on both sides of the line*. If donor sales themselves shift because of the treatment, then the "counterfactual" we build from them is contaminated.

This chapter ports the replication of @sakaguchi2026spatial, which relaxes both ingredients in two stages:

1. **Stage 2** swaps the simplex for a **horseshoe prior** on the donor weights. Sparsity becomes a feature of the posterior, not a hard constraint.
2. **Stage 3** adds a **spatial autoregressive (SAR)** layer to the donor data-generating process. SUTVA is no longer assumed — spillover effects on neighbours become a *derived* quantity, with Nevada absorbing essentially all the mass.

We rerun Stage 1 in this chapter too, so the three estimates sit in one table. The donor pool here is the 38 US states; the treated unit is California; the outcome is per-capita cigarette sales over 1970–2000.

## The three-stage pipeline

```{mermaid}
%%| label: fig-stages-pipeline
%%| fig-cap: "The three-stage Sakaguchi & Tagawa pipeline: classical Abadie SCM, then Bayesian horseshoe SCM, then Bayesian spatial SCM. Each stage relaxes one assumption — Stage 2 the simplex, Stage 3 SUTVA."
flowchart LR
    A["Stage 1<br/>Classical SCM<br/>simplex weights<br/>SUTVA imposed"] --> B["Stage 2<br/>Bayesian SCM<br/>horseshoe prior<br/>SUTVA imposed"]
    B --> C["Stage 3<br/>Bayesian Spatial SCM<br/>horseshoe + SAR ρ<br/>SUTVA relaxed"]
    C --> D["Prior predictive<br/>+ cross-stage<br/>comparison"]

    style A fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style B fill:#d97757,stroke:#cbd5e0,color:#fff
    style C fill:#00d4c8,stroke:#cbd5e0,color:#141413
    style D fill:#7A209F,stroke:#cbd5e0,color:#fff
```

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, set MCMC constants, seed, and ggplot theme."
library(tidyverse)
library(tidysynth)
library(Rcpp)
library(RcppArmadillo)
library(Matrix)
library(coda)
library(patchwork)
library(glue)
source("R/table_helpers.R")

SEED       <- 20251022L
MCMC_ITER  <- 5000L   # tutorial scale; raise to 100000L for paper-grade rho ESS
MCMC_BURN  <- 2500L   # half of MCMC_ITER; bump to 50000L when MCMC_ITER hits 100000L
TREAT_YEAR <- 1988L
set.seed(SEED)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

The helpers from the `scspill` replication package are vendored under `R/scspill/`; the two C++ MCMC kernels live alongside them and are compiled at chunk runtime by `Rcpp::sourceCpp()`. On a fresh checkout the first render compiles the kernels (~30 s) and runs the Gibbs sampler (~1–3 min); subsequent renders hit Quarto's `_freeze/` cache.

R's default `Makeconf` on macOS expects the CRAN gfortran toolchain at `/opt/gfortran/`. If that directory is missing, the linker fails on `-lemutls_w`. The chunk below is a **best-effort fallback**: it detects the situation and writes a temporary `Makevars` that points `FLIBS` at any available Homebrew `gcc` install (Intel layout at `/usr/local/Cellar/`, Apple Silicon layout at `/opt/homebrew/Cellar/`, or the symlinked `/opt/homebrew/lib/gcc/current`) before `Rcpp::sourceCpp()` runs. If neither `/opt/gfortran/` nor a Homebrew `gcc` install is present, the chunk silently skips the override and `Rcpp::sourceCpp()` will fail at link time with `-lemutls_w`. The fix in that case is to install gfortran from <https://mac.r-project.org/tools/> (the CRAN-blessed toolchain) or via Homebrew (`brew install gcc`). `R/scspill/.Makevars-rcpp.example` shows the static form of the same override for readers who want to persist it.

```{r}
#| label: vendor-load
#| message: false
#| warning: false
#| code-summary: "Code: Patch FLIBS on macOS, source scspill helpers, and compile Rcpp kernels."
if (Sys.info()[["sysname"]] == "Darwin" && !dir.exists("/opt/gfortran")) {
  brew_libs <- c(
    Sys.glob("/usr/local/Cellar/gcc/*/lib/gcc/*"),    # Intel Homebrew
    Sys.glob("/opt/homebrew/Cellar/gcc/*/lib/gcc/*"), # Apple Silicon Homebrew
    Sys.glob("/opt/homebrew/lib/gcc/current")         # symlinked layout
  )
  if (length(brew_libs) >= 1L) {
    mk <- tempfile(fileext = ".Makevars")
    writeLines(
      sprintf("FLIBS = -L%s -lgfortran -lquadmath", brew_libs[length(brew_libs)]),
      mk
    )
    Sys.setenv(R_MAKEVARS_USER = mk)
  }
}

scspill_R <- c(
  "01_utils.R", "02_utils_data_prep.R", "03_utils_plot.R",
  "04_utils_diagnostics.R", "10_sc_spillover.R",
  "21_mcmc_alpha.R", "22_mcmc_sar.R", "41_robustness_check.R"
)
for (h in scspill_R) source(file.path("R/scspill", h))

Rcpp::sourceCpp("R/scspill/20_mcmc.cpp")
Rcpp::sourceCpp("R/scspill/40_geweke_latest.cpp")
```

```{r}
#| label: data-load
#| code-summary: "Code: Load the California smoking panel and extract w, W spatial structures."
load("data/california_smoking.rda")

panel_df <- california_smoking$panel_df |>
  mutate(treatment = if_else(state == "California" & year >= TREAT_YEAR, 1L, 0L))

w <- as.matrix(california_smoking$w[, 2])
W <- as.matrix(california_smoking$W[, -1])
rownames(W) <- colnames(W) <- california_smoking$W$state
```

```{r}
#| label: panel-summary
#| code-summary: "Code: Print a one-line summary of the panel dimensions and treatment window."
glue(
  "Panel: {nrow(panel_df)} rows | {length(unique(panel_df$state))} states | ",
  "years {min(panel_df$year)}-{max(panel_df$year)}\n",
  "Treated: California | Donors: {length(unique(panel_df$state)) - 1L} | ",
  "Pre: {min(panel_df$year)}-{TREAT_YEAR - 1L} | Post: {TREAT_YEAR}-{max(panel_df$year)}"
)
```

The bundled `california_smoking.rda` carries only `cigsale` and `retprice` — narrower than the predictor set chapter 4 used (which also included `lnincome`, `age15to24`, and `beer`). That is the dominant reason the Stage 1 ATT below ($\approx -18.5$) is slightly smaller in magnitude than chapter 4's $-18.85$: the methodological pipeline is identical, but the inputs differ. The `.rda` *also* ships two spatial structures we need in Stage 3 — California's contiguity row `w` (Arizona, Nevada, and Oregon non-zero) and the $38 \times 38$ donor contiguity matrix `W` — which the richer Proposition 99 dataset does not have.

A notation note before we proceed. In chapters 4 and 6, lowercase $w$ denoted the *simplex-constrained donor weights*; in this chapter, donor weights are written $w$ as well (to keep symbol use consistent across the SCM family), while the contiguity column carries the same name in R code only — in the math below, the contiguity row that picks out California's neighbours is also written $w$ (this is the Sakaguchi-Tagawa convention). The two roles are distinguished by context: $w$ inside an SCM regression is donor weights; $w$ inside a SAR row-coupling is the adjacency row. Where ambiguity would matter, we say "donor weights $w$" or "California's contiguity row $w$" explicitly. The book-wide $W$ stays reserved for the 38×38 spatial adjacency matrix in this chapter only. Also note the horseshoe scale parameters introduced in Stage 2 below carry an "HS" subscript ($\tau_{\mathrm{HS}}$, $\lambda_{j,\mathrm{HS}}$) to keep them distinct from any treatment-effect $\tau$ or factor-loading $\lambda_i$ used elsewhere in the book.

## Stage 1 — Classical synthetic control

Classical SCM picks donor weights $w$ on the simplex that minimise the pre-treatment fit error between California and the synthetic (this is $\widehat{Y_{c,t}(0)}$ in the book-wide notation). Formally,

$$\widehat{w} \,=\, \arg\min_w \, \big\| Y_{1,\text{pre}} - Y_{c,\text{pre}}\, w \big\|^2 \;\; \text{s.t.} \;\; w_j \ge 0, \;\; \sum_j w_j = 1.$$

We use `tidysynth` with a small set of predictors (the mean of `cigsale` and `retprice` over the pre-period, plus three single-year lags of `cigsale`).

```{r}
#| label: stage1-fit
#| message: false
#| warning: false
#| code-summary: "Code: Fit the classical simplex SCM with tidysynth and compute the post-period ATT."
sc_classic <- panel_df |>
  synthetic_control(
    outcome = cigsale, unit = state, time = year,
    i_unit  = "California", i_time = TREAT_YEAR,
    generate_placebos = FALSE
  ) |>
  generate_predictor(
    time_window = 1970:(TREAT_YEAR - 1L),
    cigsale_avg_pre = mean(cigsale, na.rm = TRUE),
    retprice_avg    = mean(retprice, na.rm = TRUE)
  ) |>
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = TREAT_YEAR - 1L, cigsale_pre = cigsale) |>
  generate_weights(optimization_window = 1970:(TREAT_YEAR - 1L)) |>
  generate_control()

traj_classic <- grab_synthetic_control(sc_classic) |>
  rename(year = time_unit, observed = real_y, synthetic = synth_y) |>
  mutate(gap = observed - synthetic,
         period = if_else(year < TREAT_YEAR, "pre", "post"))

att_classic <- mean(traj_classic$gap[traj_classic$period == "post"])
```

```{r}
#| label: tbl-stage1-weights
#| tbl-cap: "Stage 1 — top-5 donor weights from the classical simplex optimiser."
#| code-summary: "Code: Tabulate the top-5 donor weights from the Stage 1 simplex fit."
grab_unit_weights(sc_classic) |>
  rename(state = unit) |>
  arrange(desc(weight)) |>
  head(5) |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor state", weight = "Weight")
```

```{r}
#| label: fig-stage1-paths
#| fig-cap: "Stage 1 — California (observed, orange) vs the classical-SCM synthetic (steel blue). Pre-1988 the two paths are nearly indistinguishable; post-1988 California falls visibly below."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot the observed and Stage 1 synthetic cigarette-sales paths."
ggplot(traj_classic, aes(x = year)) +
  geom_line(aes(y = synthetic), color = "#6a9bcc", linewidth = 0.9) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

```{r}
#| label: stage1-att-print
#| code-summary: "Code: Print the Stage 1 classical-SCM ATT estimate."
glue("Stage 1 ATT (Classical SCM): {format(round(att_classic, 2), nsmall = 2)} packs/capita")
```

Utah, Nevada, Montana, and Connecticut together carry essentially all the weight; the remaining 34 donors are flat zero. The ATT of about $-18.5$ packs/capita is recovered from a *near-deterministic* synthetic. Two questions follow: is the four-donor sparsity a feature of the data or an artefact of the simplex constraint, and is Nevada's weight contaminated by spillovers from California? Stages 2 and 3 attack these one at a time.

## Stage 2 — Bayesian synthetic control with a horseshoe prior

The horseshoe of @carvalho2010horseshoe replaces the simplex with a heavy-tailed prior that *prefers* zero but does not rule out large weights:

$$w_j \mid \tau_{\mathrm{HS}}, \lambda_{j,\mathrm{HS}} \,\sim\, \mathcal{N}\big(0, \, \tau_{\mathrm{HS}}^2 \, \lambda_{j,\mathrm{HS}}^2\big), \quad \lambda_{j,\mathrm{HS}} \sim \mathcal{C}^+(0,1), \quad \tau_{\mathrm{HS}} \sim \mathcal{C}^+(0,1).$$

Here $\tau_{\mathrm{HS}}$ is the *horseshoe global scale* (often written plain $\tau$ in the horseshoe literature) and $\lambda_{j,\mathrm{HS}}$ is the *horseshoe local scale* for donor $j$; the subscript keeps these distinct from a treatment-effect $\tau$ or factor-loading $\lambda_i$ a reader might be carrying from other chapters. The C++ sampler uses the equivalent Makalic-Schmidt (2015) auxiliary-variable parametrisation — same posterior, easier conditionals.

The package's C++ Gibbs sampler `hs_alpha_gibbs_cpp()` returns post-burn draws of the donor-weight vector $w$ (named `alpha` in the helper code) from the resulting hierarchy. We construct the pre-treatment matrices and call it directly.

```{r}
#| label: stage2-fit
#| code-summary: "Code: Build pre/post matrices and run the horseshoe Gibbs sampler for donor weights."
years_pre  <- sort(unique(panel_df$year[panel_df$year <  TREAT_YEAR]))
years_post <- sort(unique(panel_df$year[panel_df$year >= TREAT_YEAR]))
donors     <- setdiff(sort(unique(panel_df$state)), "California")

Y0_pre  <- panel_df |>
  filter(state == "California", year < TREAT_YEAR) |>
  arrange(year) |> pull(cigsale)
Y0_post <- panel_df |>
  filter(state == "California", year >= TREAT_YEAR) |>
  arrange(year) |> pull(cigsale)

wide_cig <- function(df) {
  df |>
    pivot_wider(id_cols = year, names_from = state, values_from = cigsale) |>
    select(-year) |> select(all_of(donors)) |> as.matrix()
}
Yc_pre  <- wide_cig(filter(panel_df, state != "California", year <  TREAT_YEAR))
Yc_post <- wide_cig(filter(panel_df, state != "California", year >= TREAT_YEAR))

set.seed(SEED)
alpha_draws_hs <- hs_alpha_gibbs_cpp(
  Y0_pre, Yc_pre,
  iteration = MCMC_ITER, burn = MCMC_BURN, verbose = FALSE
)
colnames(alpha_draws_hs) <- donors
```

We then propagate uncertainty through the gap series.

```{r}
#| label: stage2-att
#| code-summary: "Code: Compute Stage 2 ATT posterior mean and 95% credible interval from gap draws."
gap_draws_hs <- Y0_post - Yc_post %*% t(alpha_draws_hs)   # T1 x M_draws
att_hs_draws <- colMeans(gap_draws_hs)
att_hs       <- mean(att_hs_draws)
att_hs_ci    <- quantile(att_hs_draws, c(0.025, 0.975), names = FALSE)

glue(
  "Stage 2 ATT (Bayesian HS): {format(round(att_hs, 2), nsmall = 2)} packs/capita, ",
  "95% CrI [{format(round(att_hs_ci[1], 2), nsmall = 2)}, ",
  "{format(round(att_hs_ci[2], 2), nsmall = 2)}]"
)
```

```{r}
#| label: tbl-stage2-weights
#| tbl-cap: "Stage 2 — top-5 donor weights from the horseshoe posterior. Only Nevada's 95% credible interval excludes zero."
#| code-summary: "Code: Summarise horseshoe posterior weights and tabulate the top 5 donors."
alpha_summary <- tibble(
  state = donors,
  mean  = colMeans(alpha_draws_hs),
  lo95  = apply(alpha_draws_hs, 2, quantile, probs = 0.025),
  hi95  = apply(alpha_draws_hs, 2, quantile, probs = 0.975)
) |>
  arrange(desc(abs(mean)))

alpha_summary |>
  head(5) |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor", mean = "Posterior mean",
             lo95 = "2.5%", hi95 = "97.5%")

n_active <- sum(alpha_summary$mean > 0.01)
```

```{r}
#| label: fig-stage2-weights
#| fig-cap: "Stage 2 — posterior mean donor weights with 95% credible intervals, sorted by mean. Most donors hug zero; Nevada is the only top weight whose interval excludes zero."
#| fig-width: 8
#| fig-height: 6
#| code-summary: "Code: Plot posterior mean donor weights with 95% credible intervals."
alpha_summary |>
  mutate(state = fct_reorder(state, mean)) |>
  ggplot(aes(x = mean, y = state)) +
  geom_vline(xintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lo95, xmax = hi95), height = 0, color = "#6a9bcc") +
  geom_point(color = "#d97757", size = 1.6) +
  labs(x = "Donor weight (posterior mean, 95% CrI)", y = NULL)
```

```{r}
#| label: fig-stage2-paths
#| fig-cap: "Stage 2 — California (orange) vs the horseshoe-posterior-mean synthetic (steel blue), with a 95% credible band on the post-period gap."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot Stage 2 observed and posterior-mean synthetic paths with a 95% credible band."
synth_mean_pre  <- Yc_pre  %*% colMeans(alpha_draws_hs)
synth_mean_post <- Yc_post %*% colMeans(alpha_draws_hs)
post_gap_lo <- apply(gap_draws_hs, 1, quantile, probs = 0.025)
post_gap_hi <- apply(gap_draws_hs, 1, quantile, probs = 0.975)
post_gap_mid <- rowMeans(gap_draws_hs)

paths_hs <- tibble(
  year      = c(years_pre, years_post),
  observed  = c(Y0_pre, Y0_post),
  synthetic = c(as.numeric(synth_mean_pre), as.numeric(synth_mean_post)),
  synth_lo  = c(rep(NA_real_, length(years_pre)), as.numeric(synth_mean_post) - (post_gap_mid - post_gap_lo)),
  synth_hi  = c(rep(NA_real_, length(years_pre)), as.numeric(synth_mean_post) - (post_gap_mid - post_gap_hi))
)

ggplot(paths_hs, aes(x = year)) +
  geom_ribbon(aes(ymin = synth_lo, ymax = synth_hi), fill = "#6a9bcc", alpha = 0.25) +
  geom_line(aes(y = synthetic), color = "#6a9bcc", linewidth = 0.9) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

The donor pool **broadens dramatically**. About `r n_active` of 38 donors carry posterior mean weight above 0.01, versus four under the classical simplex. Crucially, even among the top-five posterior means, only Nevada's 95% credible interval excludes zero — every other top weight is statistically consistent with no contribution. The teaching point is that classical SCM's sparsity is partly a constraint artefact: when we admit posterior uncertainty over weights, the data do not strongly insist on a four-donor synthetic. Note also which donor's interval excludes zero: only Nevada's. Twenty-three donors carry posterior mean mass, but **only one** is statistically distinguishable from a no-contribution donor — which foreshadows the Stage 3 finding that the spillover concentrates almost entirely on Nevada.

The ATT also moves — from $-18.5$ to about $-15.8$ packs/capita — and its 95% credible interval is wider than Stage 1's bootstrap CI by design. The horseshoe propagates donor-weight uncertainty into the gap series rather than treating $w$ as fixed at the optimiser's best guess. The interval still never reaches zero.

## Stage 3 — Bayesian spatial synthetic control with SAR spillovers

Stage 2 relaxed the simplex but kept SUTVA. Stage 3 drops SUTVA by adding a SAR layer to the donor data-generating process. Writing $A = W + w\, w^{\top}$ (where the first $w$ is California's contiguity row and the second $w$ is the Stage-2 donor-weight vector — the Sakaguchi-Tagawa double-use of the same symbol), the model is

$$(I - \rho W - \rho\, w\, w^{\top})\, Y_{c,t} \,=\, X_{c,t}\,\beta \,+\, \Lambda F_t \,+\, \varepsilon_t.$$

The three structural terms on the left isolate three different feedback channels. $\rho W Y_{c,t}$ is **donor-to-donor spillover**: each donor's outcome moves with a row-normalised average of its neighbours' outcomes. $\rho\, w\, (w^{\top} Y_{c,t})$ is the **cross-coupling term** that propagates the donor pool's joint reaction *back through California's neighbours* — and crucially, the inner $w^{\top} Y_{c,t}$ uses the **same donor-weight vector $w$ from Stage 2**: it plays double duty as the spatial-feedback weights, not as a temporal lag of donor outcomes. $\Lambda F_t$ is a $p = 1$ latent **dynamic factor** capturing unobserved common shocks (the `p_factors = 1L` argument below activates this). $\varepsilon_t$ is i.i.d. noise. When $\rho = 0$ the SAR layer collapses and Stage 3 reduces to Stage 2.

The covariate matrix $X_{c,t}$ contains the donor-side observables — in this fit the single column is retail cigarette price, so $\beta$ is a scalar.

**Priors at a glance** (the C++ kernel's defaults): $\rho$ has a uniform prior on the stability interval $\big(-0.95/\rho_{\max}(W),\, 0.95/\rho_{\max}(W)\big)$ that keeps $(I - \rho W)$ invertible; $\sigma^2 \sim \mathrm{InverseGamma}(1, 1)$ for the noise variance; $\beta$ has an improper flat prior with a $10^{-6}$ ridge for numerical stability; and the donor-weight vector $w$ inherits its Stage-2 horseshoe hierarchy, **fixed at its Step-1 posterior mean during the SAR step** (see the limitation discussed next).

The package's `sc_spillover()` runs the two MCMC samplers sequentially — first the horseshoe for $w$ (Step 1), then SAR $\rho$ holding $w$ fixed at its Step-1 posterior mean $\widehat{w}$ (Step 2) — and post-processes the per-state spillover effects in one call. **This is a plug-in approximation, not a fully joint posterior**: uncertainty in $w$ does not propagate into the $\rho$ posterior, which is one of two structural reasons the Stage-3 credible interval below is artificially narrow (the other being the low effective sample size for $\rho$ at tutorial scale).

```{r}
#| label: stage3-fit
#| message: false
#| code-summary: "Code: Run the SAR spillover sampler and extract rho, ESS, and ATT summaries."
fit_sar <- sc_spillover(
  data            = panel_df,
  treated_unit    = "California",
  w = w, W = W,
  treatment_dummy = "treatment",
  y               = "cigsale",
  X               = c("retprice"),
  p_factors       = 1L,
  M               = MCMC_ITER, burn = MCMC_BURN,
  seed            = SEED, step_rho = 0.01,
  unit_col        = "state", time_col = "year",
  verbose         = FALSE
)

rho_hat    <- fit_sar$rho_hat
ess_rho    <- coda::effectiveSize(coda::as.mcmc(fit_sar$rho_draws))[[1]]
att_sar    <- fit_sar$effects$ate_point
att_sar_ci <- fit_sar$effects$ate_ci95
```

```{r}
#| label: stage3-print
#| code-summary: "Code: Print posterior mean rho, ESS, and the Stage 3 ATT with credible interval."
glue(
  "Posterior mean rho: {format(round(rho_hat, 3), nsmall = 3)} | ",
  "ESS(rho) = {format(round(ess_rho, 1), nsmall = 1)}\n",
  "Stage 3 ATT (Bayesian Spatial SAR): {format(round(att_sar, 2), nsmall = 2)} ",
  "packs/capita, 95% CrI [{format(round(att_sar_ci[1], 2), nsmall = 2)}, ",
  "{format(round(att_sar_ci[2], 2), nsmall = 2)}]"
)
```

The posterior mean $\hat\rho \approx 0.223$ is positive (consistent with the cross-border intuition that neighbours co-move), but at ESS = 2.9 the chain has barely moved from its starting value of $0$; the magnitude of $\hat\rho$ should not be over-read at the 5,000-iteration run we use here. The callout below explains how to recover the publication-grade $\hat\rho$. The Stage 3 ATT sits between the classical and Bayesian-horseshoe estimates: adding the SAR layer *reattributes* a portion of the gap away from California's direct response toward neighbour spillovers.

:::{.callout-warning}
## ESS(ρ) at tutorial scale is too low for a trustworthy CrI

The chapter runs the sampler at 5,000 iterations to keep render times manageable. At that budget, the effective sample size for $\rho$ is in the single digits — far below the rule-of-thumb 200 — and the 95% credible interval for the Stage 3 ATT is therefore *artificially narrow*. The point estimate is recoverable (a posterior mean has low bias even at low ESS); the interval should be read as illustrative. To match the @sakaguchi2026spatial published numbers, raise `MCMC_ITER` to `100000L` and `MCMC_BURN` to `50000L` and rerender (expect ~30–90 min wall-clock).
:::

The single most important MCMC diagnostic is the traceplot: it lets a reader see *how* the chain failed to mix, not just *that* it did.

```{r}
#| label: fig-trace-rho
#| fig-cap: "Traceplot of rho over the 2,500 post-burn iterations. At this 5,000-iteration run the chain visibly fails to mix — the visual diagnosis behind ESS = 2.9. A healthy chain would look like white-noise oscillation around a stationary mean; this one drifts in long, slow excursions."
#| fig-width: 8
#| fig-height: 3
#| code-summary: "Code: Plot the post-burn traceplot of rho to diagnose chain mixing."
tibble(iter = seq_along(fit_sar$rho_draws), rho = fit_sar$rho_draws) |>
  ggplot(aes(iter, rho)) +
  geom_line(color = "#6a9bcc", linewidth = 0.4) +
  labs(x = "Iteration (post-burn)", y = expression(rho))
```

```{r}
#| label: fig-stage3-paths
#| fig-cap: "Stage 3 — California (orange) vs the SAR-spillover-corrected synthetic (teal). The treatment effect widens roughly linearly post-1988."
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: Plot the Stage 3 SAR-corrected synthetic against the observed California path."
ycf_post <- Y0_post - fit_sar$effects$te_point
paths_sar <- tibble(
  year      = years_post,
  observed  = Y0_post,
  synthetic = ycf_post
)

paths_full <- bind_rows(
  tibble(year = years_pre, observed = Y0_pre, synthetic = NA_real_),
  paths_sar
)

ggplot(paths_full, aes(x = year)) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_line(aes(y = synthetic), color = "#00d4c8", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

### Spillover effects on donor states

The interesting output of the SAR layer is the per-donor spillover. The framework recovers the per-donor counterfactual outcome in closed form by applying the Sakaguchi-Tagawa identification formula to each posterior draw of $\rho$:

$$Y_c^{cf} \,=\, (I - \rho A)^{-1}\big[(I - \rho W) Y_c \,-\, \rho\, w\, Y_0\big],$$

where $A = W + w\, w^{\top}$ (the same matrix from the SAR likelihood) and $Y_0$ is California's observed outcome path. The per-donor spillover is then $Y_c - Y_c^{cf}$, averaged over draws. No forward simulation is involved — the model is *identified* in closed form, so we evaluate, we do not simulate forward.

```{r}
#| label: tbl-spillover-top8
#| tbl-cap: "Top-8 donor states by absolute post-1988 mean spillover. Nevada absorbs roughly 16x the next state."
#| code-summary: "Code: Rank donors by mean post-1988 spillover and tabulate the top 8."
spill_mat   <- fit_sar$effects$spill
post_idx    <- which(as.numeric(rownames(spill_mat)) >= TREAT_YEAR)
spill_post  <- spill_mat[post_idx, , drop = FALSE]
spill_top8  <- tibble(state = colnames(spill_mat),
                      avg_spillover = colMeans(spill_post)) |>
  slice_max(abs(avg_spillover), n = 8) |>
  arrange(avg_spillover)

spill_top8 |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor", avg_spillover = "Mean post-1988 spillover")
```

```{r}
#| label: fig-spillover-bars
#| fig-cap: "Top-8 donor states by absolute spillover. Nevada dominates by an order of magnitude — the empirical signature of SUTVA failure for this case study."
#| fig-width: 8
#| fig-height: 4.5
#| code-summary: "Code: Plot top-8 donor spillovers as a horizontal bar chart."
spill_top8 |>
  mutate(state = fct_reorder(state, avg_spillover)) |>
  ggplot(aes(x = avg_spillover, y = state)) +
  geom_vline(xintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_col(fill = "#d97757") +
  labs(x = "Mean post-1988 spillover (packs/capita)", y = NULL)
```

**Nevada is the dominant spillover-receiver by more than an order of magnitude (≈16× the next-largest).** Idaho and Utah trail far behind; every other donor's spillover is small enough to be policy-noise. The story is consistent with cross-border tobacco flows reshaping consumption on both sides of the California–Nevada line — the empirical signature of SUTVA failure for this case study.

## Prior predictive diagnostic

Before reading Stage 3 as a posterior, we want to confirm that the prior specification is *compatible* with what the data actually look like. The helper `prior_predictive()` draws $R = 1{,}000$ joint prior samples, forward-simulates a synthetic donor panel under each draw, computes a battery of summary statistics, and compares them to the observed statistics from the real donor panel.

We focus on four summary statistics: **donor mean** (the level of the simulated donor panel), **spatial $Y'WY/NT_0$** (the quadratic form whose magnitude is the textbook Moran-like measure of how strongly neighbours co-move), **lag-1 autocorrelation** (does the prior generate AR(1)-shaped donor series?), and **PC1 variance share** (how strongly is the simulated panel dominated by one common factor?). Together they pin down the prior's behaviour on the dimensions that matter for an SCM fit: level, spatial structure, persistence, and low-rank dependence.

```{r}
#| label: prior-predictive
#| code-summary: "Code: Draw prior predictive samples and compare summary stats to observed values."
Xc_pre_arr <- panel_df |>
  filter(state != "California", year < TREAT_YEAR) |>
  pivot_wider(id_cols = year, names_from = state, values_from = retprice) |>
  select(-year) |> select(all_of(donors)) |> as.matrix()
dim(Xc_pre_arr) <- c(nrow(Xc_pre_arr), ncol(Xc_pre_arr), 1)

ppc <- prior_predictive(
  Y0_pre           = as.matrix(Y0_pre),
  Yc_obs           = Yc_pre,
  W_raw            = W,
  w_raw            = w,
  alpha_hat_scaled = colMeans(fit_sar$alpha_draws),
  Xc_pre           = Xc_pre_arr,
  p                = 0L,
  a0 = 3, b0 = 1, rho_support = c(-0.99, 0.99),
  R                = 1000L,
  seed             = SEED
)
```

```{r}
#| label: fig-prior-predictive
#| fig-cap: "Prior predictive check on four summary statistics. Histograms are R = 1,000 simulated prior draws; the orange line is the observed value. All four observed lines land inside the simulated prior cloud — the prior is compatible with the data, not overwhelming it."
#| fig-width: 9
#| fig-height: 6
#| code-summary: "Code: Plot prior predictive histograms of four summary statistics with observed lines."
stats_show <- c("yc\\_mean", "spatial\\_quadratic", "ac1", "pve\\_pc1")
pretty <- c("Donor mean", "Spatial y'Wy", "Lag-1 autocorrelation", "PC1 variance share")

ppc_long <- ppc$stat |>
  as_tibble() |>
  select(all_of(stats_show)) |>
  rename_with(~ pretty, all_of(stats_show)) |>
  pivot_longer(everything(), names_to = "stat", values_to = "value")
obs_df <- tibble(
  stat  = pretty,
  value = unname(ppc$observed[stats_show])
)

ggplot(ppc_long, aes(value)) +
  geom_histogram(bins = 30, fill = "#6a9bcc", color = "white") +
  geom_vline(data = obs_df, aes(xintercept = value),
             color = "#d97757", linewidth = 0.8) +
  facet_wrap(~ stat, scales = "free", ncol = 2) +
  labs(x = NULL, y = "Prior-draw count")
```

All four observed orange lines land *inside* the simulated prior cloud rather than in the tails: the prior is compatible with the data, not overwhelming it. Had any observed statistic landed deep in a prior tail, the corresponding posterior estimate would have been pulled by the prior rather than the likelihood.

## Cross-stage comparison

Stacking the three estimators makes the pedagogical arc visible.

```{r}
#| label: tbl-cross-stage
#| tbl-cap: "Cross-stage comparison of ATT estimates. The active-donor count jumps from 4 to 23 between Stage 1 and Stage 2 and then stays at 23 in Stage 3, because Stage 3 conditions on Stage 2's w-hat (plug-in approximation). The Stage 3 CrI is artificially narrow because ESS(rho) is in single digits at the 5,000-iteration scale."
#| code-summary: "Code: Tabulate cross-stage ATT estimates, intervals, active donors, and ESS(rho)."
tibble(
  Stage          = c("Classical SCM (tidysynth)",
                     "Bayesian HS (no spillovers)",
                     "Bayesian Spatial SAR (with spillovers)"),
  ATT            = c(att_classic, att_hs, att_sar),
  `2.5%`         = c(NA_real_,    att_hs_ci[1], att_sar_ci[1]),
  `97.5%`        = c(NA_real_,    att_hs_ci[2], att_sar_ci[2]),
  `Active donors`= c(sum(grab_unit_weights(sc_classic)$weight > 0.01),
                     sum(alpha_summary$mean > 0.01),
                     sum(colMeans(fit_sar$alpha_draws) > 0.01)),
  `ESS(rho)`     = c(NA_real_, NA_real_, ess_rho)
) |>
  gt_pretty(decimals = 2)
```

Three observations.

- **Sign and order of magnitude agree.** All three estimators place the ATT between roughly $-15$ and $-19$ packs/capita per year, and none of the intervals reaches zero. Whatever one believes about the simplex or SUTVA, Proposition 99 reduced California cigarette consumption.
- **Active-donor count jumps once.** Heavier-tailed priors admit more donors with non-trivial mass (4 → 23 between Stage 1 and Stage 2); the Stage 3 count is identical to Stage 2 by construction (the SAR step holds $w$ fixed at its Stage-2 posterior mean). The sparse four-donor synthetic of Stage 1 looks like one plausible counterfactual rather than the right one.
- **The Stage 3 credible interval is the narrowest but the least trustworthy.** ESS(ρ) is in single digits and the plug-in for $w$ removes a second source of uncertainty that ought to widen the interval; downstream prose should treat that interval as illustrative.

## Common pitfall: tutorial-scale MCMC and the credibility of credible intervals

The Stage 3 credible interval in the cross-stage table is the narrowest of the three, and a careless reader could conclude that the Bayesian spatial SAR is also the *most precise*. It is not. Three caveats compound at the 5,000-iteration budget this chapter uses:

1. **ESS(ρ) is in single digits.** The chain has barely moved from its starting value of $0$; the credible interval is built on at most three or four effectively-independent draws, so its width is dominated by sampling noise, not by genuine posterior concentration. Raise `MCMC_ITER` to `100000L` and `MCMC_BURN` to `50000L` to recover a trustworthy interval (≈30–90 min wall-clock).
2. **The plug-in for $w$ removes a source of uncertainty.** Step 2 of the sampler conditions on $\widehat{w}$ from Step 1 rather than sampling $w$ and $\rho$ jointly. Posterior uncertainty in donor weights therefore does not propagate into the $\rho$ posterior, narrowing the Stage 3 ATT interval beyond what a fully joint posterior would produce. A fully joint sampler would widen Stage 3 even at high ESS.
3. **`sc_spillover()` runs a single chain.** For paper-grade work run 3–4 chains from over-dispersed starts and compute $\hat R$ (the Gelman-Rubin convergence diagnostic); a between-chain spread that exceeds within-chain variance flags non-convergence that ESS alone cannot detect.

The point estimate $\widehat{\mathrm{ATT}} \approx -16.6$ packs/capita is robust to all three issues — posterior means are low-bias even at low ESS — but **intervals demand discipline**. Treat the Stage 3 CrI as illustrative until the three points above are addressed.

## Recap

| Question | Answer |
|---|---|
| What does this chapter estimate? | The ATT of Proposition 99 on California cigarette sales, 1988–2000 |
| Which assumption does Stage 2 relax? | The simplex constraint on donor weights — replaced by a horseshoe prior |
| Which assumption does Stage 3 relax? | SUTVA — donor outcomes are now allowed to respond to the treatment via a SAR layer |
| Where does the spillover land? | Almost entirely on Nevada (mean post-1988 spillover $\approx -3.75$ packs/capita on Nevada); other donors are an order of magnitude smaller |
| What is the chapter's caveat? | At the 5,000-iteration run the ESS for ρ is too small for the Stage 3 CrI to be trustworthy — raise iterations for paper-grade inference |
| Is the headline robust? | Yes — all three ATT estimates are negative, never cross zero, and agree on order of magnitude |

## Where this case study ends, and where Part II begins

This is the last chapter that works the Proposition 99 case study. Across seven chapters we have asked the same question — did California's 1989 tobacco tax reduce per-capita cigarette consumption? — under progressively weaker assumptions: an interrupted time series (ch.2), a 2×2 difference-in-differences (ch.3), classical synthetic control on the simplex (ch.4), a Bayesian structural time-series counterfactual (ch.5), a frequentist prediction interval around the SC fit (ch.6), and finally a Bayesian spatial relaxation of both the simplex and SUTVA (this chapter). The headline survives every reframe: the estimated ATT is consistently between roughly $-15$ and $-19$ packs per capita, and never crosses zero.

The headline survives because Proposition 99 is a *single-treated-unit* case with a clean pre-period and one large, sharp policy break. Most real-world policy data is messier: many units adopt at different dates, treatment effects evolve, and parallel-trends violations need to be diagnosed rather than assumed away. Part II (chapters 8–10) leaves Proposition 99 behind and picks up the **Callaway-Sant'Anna minimum-wage county panel** (1,745 US counties × 2003–2007, three adoption cohorts), where the methods we have used so far either fail outright (one treated unit ≠ staggered cohorts) or scale awkwardly. Chapter 8 starts there with the modern staggered-DiD toolkit, generalising the *ATT on California* estimand we have estimated seven different ways here into the *ATT(g, t)* framework that lets a researcher decompose, weight, and sensitivity-test heterogeneous treatment effects across cohorts and event times.

## Key takeaways

- **Method:** Bayesian spatial synthetic control estimates the ATT on a single treated unit (California) by combining two relaxations of the classical SCM in @abadie2010synthetic: a **horseshoe prior** on donor weights $w$ — a heavy-tailed shrinkage prior $w_j \sim \mathcal{N}(0, \tau_{\mathrm{HS}}^2 \lambda_{j,\mathrm{HS}}^2)$ with $\lambda_{j,\mathrm{HS}}, \tau_{\mathrm{HS}} \sim \mathcal{C}^+(0,1)$ that *prefers* zero but does not rule out large weights — and a **spatial autoregressive (SAR) layer** $(I - \rho W - \rho\, w\, w^{\top})\, Y_{c,t} = X_{c,t}\beta + \Lambda F_t + \varepsilon_t$ on the donor data-generating process.
- **Method:** Adding the SAR layer relaxes **SUTVA** (the stable unit treatment value assumption — that one unit's treatment does not affect another unit's outcome) by letting donor outcomes themselves absorb the treated unit's policy through donor-to-donor diffusion ($\rho W Y_{c,t}$) and a cross-coupling term ($\rho\, w\, w^{\top} Y_{c,t}$) that propagates the donor pool's joint reaction back through California's neighbours; the cross-border spillover on each donor is then *derived* in closed form, not assumed away.
- **Lesson:** Whatever SUTVA imposes here is a *constraint on a specific donor*, not a constraint on the headline: all three ATT estimates (classical $-18.5$, horseshoe $-15.8$, SAR $-16.6$ packs/capita) sit between $-15$ and $-19$ and none of the intervals reaches zero — the policy effect on California survives relaxing both the simplex and SUTVA.
- **Lesson:** Replacing the simplex with a horseshoe broadens the active donor pool from 4 to 23 — the classical sparsity is shown to be partly a constraint artefact rather than a feature of the data — but only Nevada's posterior 95% credible interval excludes zero, foreshadowing where the Stage 3 spillover lands.
- **Lesson:** The headline derived quantity is the **Nevada spillover**: Nevada absorbs roughly 16× the next-largest donor's mean post-1988 spillover, the empirical signature of cross-border cigarette flows reshaping consumption on both sides of the California-Nevada line. SUTVA fails for this case study, and the SAR layer localises the failure to one neighbour.
- **Lesson:** A prior predictive check on four summary statistics (donor mean, spatial $Y'WY$, lag-1 autocorrelation, PC1 variance share) confirms the prior is compatible with — not overwhelming — the data: a routine sanity check before trusting any Bayesian posterior.
- **Caveat:** Identifying the spatial-lag parameter $\rho$ from a single observational panel is hard. At the tutorial-scale 5,000-iteration run, ESS($\rho$) sits in the single digits (≈2.9), the traceplot visibly fails to mix, and the Stage 3 credible interval is narrowest precisely because it is least trustworthy; recovering a publication-grade $\hat\rho$ requires raising iterations to 100,000 and running multiple chains with a Gelman-Rubin $\hat R$ diagnostic.
- **Caveat:** `sc_spillover()` runs a **plug-in approximation**, not a fully joint posterior — the SAR step fixes $w$ at its Stage-2 posterior mean rather than sampling $(w, \rho)$ jointly. Uncertainty in donor weights therefore does not propagate into $\rho$, narrowing the Stage 3 ATT interval beyond what a fully joint sampler would produce.
- **Caveat:** The bundled `california_smoking.rda` carries only `cigsale` and `retprice` — a narrower predictor set than chapter 4 — which is why the Stage 1 ATT here is $-18.5$ rather than the $-18.85$ of the original Abadie replication. The methodological pipeline is identical; the inputs differ.

## Further reading

- @sakaguchi2026spatial — the original Bayesian spatial-SCM paper this chapter replicates.
- @carvalho2010horseshoe — the horseshoe prior used in Stage 2.
- @lesage2009introduction — textbook treatment of SAR and spatial econometrics.
- @abadie2010synthetic — the original synthetic-control treatment of Proposition 99, replicated in chapter 4.
- @dunford2024tidysynth — `tidysynth` package documentation.
