---
title: "Bayesian Spatial Synthetic Control"
---

## Why a third synthetic-control chapter?

Chapter 5 fit a *classical* Synthetic Control: donor weights live on the simplex and are chosen by a quadratic optimiser. Chapter 6 borrowed donor information in a different way — through a Bayesian structural time-series model. Both treat the donor states' outcomes as *unaffected* by California's policy. That assumption — the **stable unit treatment value assumption (SUTVA)** — is the price of any synthetic-control estimate.

For tobacco, SUTVA is empirically suspect. When California raised retail prices in 1989, cross-border-shopping into Nevada changed cigarette sales *on both sides of the line*. If donor sales themselves shift because of the treatment, then the "counterfactual" we build from them is contaminated.

This chapter ports the replication of @sakaguchi2026spatial, which relaxes both ingredients in two stages:

1. **Stage 2** swaps the simplex for a **horseshoe prior** on the donor weights. Sparsity becomes a feature of the posterior, not a hard constraint.
2. **Stage 3** adds a **spatial autoregressive (SAR)** layer to the donor data-generating process. SUTVA is no longer assumed — spillover effects on neighbours become a *derived* quantity, with Nevada absorbing essentially all the mass.

We rerun Stage 1 in this chapter too, so the three estimates sit in one table. The donor pool here is the 38 US states; the treated unit is California; the outcome is per-capita cigarette sales over 1970–2000.

## The three-stage pipeline

```{mermaid}
%%| label: fig-stages-pipeline
%%| fig-cap: "The three-stage Sakaguchi & Tagawa pipeline: classical Abadie SCM, then Bayesian horseshoe SCM, then Bayesian spatial SCM. Each stage relaxes one assumption — Stage 2 the simplex, Stage 3 SUTVA."
flowchart LR
    A["Stage 1<br/>Classical SCM<br/>simplex weights<br/>SUTVA imposed"] --> B["Stage 2<br/>Bayesian SCM<br/>horseshoe prior<br/>SUTVA imposed"]
    B --> C["Stage 3<br/>Bayesian Spatial SCM<br/>horseshoe + SAR ρ<br/>SUTVA relaxed"]
    C --> D["Prior predictive<br/>+ cross-stage<br/>comparison"]

    style A fill:#6a9bcc,stroke:#cbd5e0,color:#fff
    style B fill:#d97757,stroke:#cbd5e0,color:#fff
    style C fill:#00d4c8,stroke:#cbd5e0,color:#141413
    style D fill:#7A209F,stroke:#cbd5e0,color:#fff
```

## Key concepts at a glance

The rest of the chapter leans on a small vocabulary. The **definition** is always visible — open the *example* and *analogy* cards when you need them, leave them collapsed for a quick scan.

**1. Average Treatment Effect on the Treated (ATT)** — $\mathrm{ATT} = E[Y_{it}(1) - Y_{it}(0) \mid D_{it} = 1]$. The causal effect averaged over the units that *actually* received the treatment. In synthetic control there is one treated unit, so the ATT is the average gap between observed California and synthetic California over 1988–2000.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Stage 1 gives $\widehat{\mathrm{ATT}} \approx -18.5$ packs/capita/year. Stages 2 and 3 target the same ATT but build the synthetic California differently.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

A patient took a new drug; we never see the same patient untreated. We average across similar untreated patients to imagine what the treated patient *would have* done.

</details>
</div>

**2. Synthetic control method (SCM)** — $\widehat{Y}_{1,t}(0) = \sum_j \alpha_j Y_{j,t}$. The counterfactual for the treated unit is a weighted blend of donor units. Classical SCM constrains $\alpha$ to the simplex.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Classical SCM (Stage 1) puts essentially all the weight on four donors — Utah, Nevada, Montana, Connecticut. The synthetic California is a blend of those four states.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

A perfumer mixes a few base scents to imitate one fragrance. The recipe (the weights) is chosen so the imitation matches the original on every pre-treatment day.

</details>
</div>

**3. Donor pool** — the set of units eligible to build the synthetic counterfactual. They must be untreated throughout the study window and similar enough to the treated unit on pre-treatment characteristics.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Here the donor pool is the 38 US states other than California that did not change their tobacco taxes around 1988.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

A casting call: 38 actors audition to play "California without Prop 99". The director picks a weighted blend rather than one body double.

</details>
</div>

**4. Horseshoe prior** — $\alpha_j \mid \tau, \lambda_j \sim \mathcal{N}(0, \, \tau^2 \lambda_j^2)$ with $\tau, \lambda_j \sim \mathcal{C}^+(0,1)$. A heavy-tailed prior that simultaneously favours sparsity (most weights near zero) and allows a handful of large weights to escape the shrinkage.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

In Stage 2 the horseshoe spreads non-trivial posterior mass across roughly 20 of 38 donors. Connecticut and Nevada carry the largest means; only Nevada's 95% credible interval excludes zero.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

A juror who starts every defendant near "not guilty" but is willing to convict the very few against whom the evidence is overwhelming.

</details>
</div>

**5. SUTVA** — the stable unit treatment value assumption: a donor's no-treatment outcome does *not* depend on whether other units were treated. SUTVA fails if California's policy reshapes Nevada's sales.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Stage 3 estimates a post-treatment spillover of roughly $-3.75$ packs/capita on Nevada — an order of magnitude larger than on any other donor. Classical SCM silently treats this as part of the donor signal.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

Two adjacent restaurants. The new health code at restaurant A drives away customers, some of whom walk into restaurant B. Treating B as unaffected understates the policy's reach.

</details>
</div>

**6. Spatial autoregressive (SAR) model** — $y = \rho \, W y + X \beta + \varepsilon$. Each unit's outcome depends on a row-normalised neighbour average. The scalar $\rho$ measures how strongly each unit co-moves with its neighbours after controlling for covariates.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Stage 3's posterior mean is $\hat\rho \approx 0.22$. A 1-unit change in the neighbour-averaged sales is associated with about a 0.22-unit change in own sales.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

How loudly your neighbours' music sets the volume of yours, holding your taste fixed. $\rho = 0$ means you ignore them; $\rho \to 1$ means your stereo basically copies theirs.

</details>
</div>

**7. Spillover effect** — the effect on a *donor* unit of the treatment imposed on the treated unit. Under SUTVA this is assumed zero. Under the SAR layer it emerges as a derived quantity once $\rho$ and $W$ are estimated.

<div class="concept-pair">
<details class="concept-card concept-example"><summary>Example</summary>

Nevada absorbs essentially all the spillover mass ($\approx -3.75$ packs/capita). Idaho and Utah each receive roughly $-0.23$. The remaining 35 donors are essentially untouched.

</details>
<details class="concept-card concept-analogy"><summary>Analogy</summary>

A neighbour's leaky pipe. Most of the room stays dry, but the one wall sharing plumbing with the neighbour soaks through.

</details>
</div>

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(tidysynth)
library(Rcpp)
library(RcppArmadillo)
library(Matrix)
library(coda)
library(patchwork)
library(glue)
source("R/table_helpers.R")

SEED       <- 20251022L
MCMC_ITER  <- 5000L
MCMC_BURN  <- 2500L
TREAT_YEAR <- 1988L
set.seed(SEED)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

The helpers from the `scspill` replication package are vendored under `R/scspill/`; the two C++ MCMC kernels live alongside them and are compiled at chunk runtime by `Rcpp::sourceCpp()`. On a fresh checkout the first render compiles the kernels (~30 s) and runs the Gibbs sampler (~1–3 min); subsequent renders hit Quarto's `_freeze/` cache.

R's default `Makeconf` on macOS expects the CRAN gfortran toolchain at `/opt/gfortran/`. If that directory is missing, the linker fails on `-lemutls_w`. The chunk below detects the situation and writes a temporary `Makevars` that points `FLIBS` at any available Homebrew `gcc` install before `Rcpp::sourceCpp()` runs. `R/scspill/.Makevars-rcpp.example` shows the static form of the same override for readers who want to persist it.

```{r}
#| label: vendor-load
#| message: false
#| warning: false
if (Sys.info()[["sysname"]] == "Darwin" && !dir.exists("/opt/gfortran")) {
  brew_libs <- Sys.glob("/usr/local/Cellar/gcc/*/lib/gcc/*")
  if (length(brew_libs) >= 1L) {
    mk <- tempfile(fileext = ".Makevars")
    writeLines(
      sprintf("FLIBS = -L%s -lgfortran -lquadmath", brew_libs[length(brew_libs)]),
      mk
    )
    Sys.setenv(R_MAKEVARS_USER = mk)
  }
}

scspill_R <- c(
  "01_utils.R", "02_utils_data_prep.R", "03_utils_plot.R",
  "04_utils_diagnostics.R", "10_sc_spillover.R",
  "21_mcmc_alpha.R", "22_mcmc_sar.R", "41_robustness_check.R"
)
for (h in scspill_R) source(file.path("R/scspill", h))

Rcpp::sourceCpp("R/scspill/20_mcmc.cpp")
Rcpp::sourceCpp("R/scspill/40_geweke_latest.cpp")
```

```{r}
#| label: data-load
load("data/california_smoking.rda")

panel_df <- california_smoking$panel_df |>
  mutate(treatment = if_else(state == "California" & year >= TREAT_YEAR, 1L, 0L))

w <- as.matrix(california_smoking$w[, 2])
W <- as.matrix(california_smoking$W[, -1])
rownames(W) <- colnames(W) <- california_smoking$W$state
```

```{r}
#| label: panel-summary
glue(
  "Panel: {nrow(panel_df)} rows | {length(unique(panel_df$state))} states | ",
  "years {min(panel_df$year)}-{max(panel_df$year)}\n",
  "Treated: California | Donors: {length(unique(panel_df$state)) - 1L} | ",
  "Pre: {min(panel_df$year)}-{TREAT_YEAR - 1L} | Post: {TREAT_YEAR}-{max(panel_df$year)}"
)
```

The bundled `california_smoking.rda` carries only `cigsale` and `retprice` — narrower than the predictor set chapter 5 used (which also included `lnincome`, `age15to24`, and `beer`). That is the dominant reason the Stage 1 ATT below ($\approx -18.5$) is slightly smaller in magnitude than chapter 5's $-18.85$: the methodological pipeline is identical, but the inputs differ. The `.rda` *also* ships two spatial structures we need in Stage 3 — California's contiguity row `w` (Arizona, Nevada, and Oregon non-zero) and the $38 \times 38$ donor contiguity matrix `W` — which the richer Proposition 99 dataset does not have.

## Stage 1 — Classical synthetic control

Classical SCM picks donor weights on the simplex that minimise the pre-treatment fit error between California and the synthetic. Formally,

$$\widehat\alpha \,=\, \arg\min_\alpha \, \big\| Y_{1,\text{pre}} - Y_{c,\text{pre}}\, \alpha \big\|^2 \;\; \text{s.t.} \;\; \alpha_j \ge 0, \;\; \sum_j \alpha_j = 1.$$

We use `tidysynth` with a small set of predictors (the mean of `cigsale` and `retprice` over the pre-period, plus three single-year lags of `cigsale`).

```{r}
#| label: stage1-fit
#| message: false
#| warning: false
sc_classic <- panel_df |>
  synthetic_control(
    outcome = cigsale, unit = state, time = year,
    i_unit  = "California", i_time = TREAT_YEAR,
    generate_placebos = FALSE
  ) |>
  generate_predictor(
    time_window = 1970:(TREAT_YEAR - 1L),
    cigsale_avg_pre = mean(cigsale, na.rm = TRUE),
    retprice_avg    = mean(retprice, na.rm = TRUE)
  ) |>
  generate_predictor(time_window = 1975, cigsale_1975 = cigsale) |>
  generate_predictor(time_window = 1980, cigsale_1980 = cigsale) |>
  generate_predictor(time_window = TREAT_YEAR - 1L, cigsale_pre = cigsale) |>
  generate_weights(optimization_window = 1970:(TREAT_YEAR - 1L)) |>
  generate_control()

traj_classic <- grab_synthetic_control(sc_classic) |>
  rename(year = time_unit, observed = real_y, synthetic = synth_y) |>
  mutate(gap = observed - synthetic,
         period = if_else(year < TREAT_YEAR, "pre", "post"))

att_classic <- mean(traj_classic$gap[traj_classic$period == "post"])
```

```{r}
#| label: tbl-stage1-weights
#| tbl-cap: "Stage 1 — top-5 donor weights from the classical simplex optimiser."
grab_unit_weights(sc_classic) |>
  rename(state = unit) |>
  arrange(desc(weight)) |>
  head(5) |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor state", weight = "Weight")
```

```{r}
#| label: fig-stage1-paths
#| fig-cap: "Stage 1 — California (observed, orange) vs the classical-SCM synthetic (steel blue). Pre-1988 the two paths are nearly indistinguishable; post-1988 California falls visibly below."
#| fig-width: 8
#| fig-height: 5
ggplot(traj_classic, aes(x = year)) +
  geom_line(aes(y = synthetic), color = "#6a9bcc", linewidth = 0.9) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

```{r}
#| label: stage1-att-print
glue("Stage 1 ATT (Classical SCM): {format(round(att_classic, 2), nsmall = 2)} packs/capita")
```

Utah, Nevada, Montana, and Connecticut together carry essentially all the weight; the remaining 34 donors are flat zero. The ATT of about $-18.5$ packs/capita is recovered from a *near-deterministic* synthetic. Two questions follow: is the four-donor sparsity a feature of the data or an artefact of the simplex constraint, and is Nevada's weight contaminated by spillovers from California? Stages 2 and 3 attack these one at a time.

## Stage 2 — Bayesian synthetic control with a horseshoe prior

The horseshoe of @carvalho2010horseshoe replaces the simplex with a heavy-tailed prior that *prefers* zero but does not rule out large weights:

$$\alpha_j \mid \tau, \lambda_j \,\sim\, \mathcal{N}\big(0, \, \tau^2 \lambda_j^2\big), \quad \lambda_j \sim \mathcal{C}^+(0,1), \quad \tau \sim \mathcal{C}^+(0,1).$$

The package's C++ Gibbs sampler `hs_alpha_gibbs_cpp()` returns post-burn draws of $\alpha$ from the resulting hierarchy. We construct the pre-treatment matrices and call it directly.

```{r}
#| label: stage2-fit
years_pre  <- sort(unique(panel_df$year[panel_df$year <  TREAT_YEAR]))
years_post <- sort(unique(panel_df$year[panel_df$year >= TREAT_YEAR]))
donors     <- setdiff(sort(unique(panel_df$state)), "California")

Y0_pre  <- panel_df |>
  filter(state == "California", year < TREAT_YEAR) |>
  arrange(year) |> pull(cigsale)
Y0_post <- panel_df |>
  filter(state == "California", year >= TREAT_YEAR) |>
  arrange(year) |> pull(cigsale)

wide_cig <- function(df) {
  df |>
    pivot_wider(id_cols = year, names_from = state, values_from = cigsale) |>
    select(-year) |> select(all_of(donors)) |> as.matrix()
}
Yc_pre  <- wide_cig(filter(panel_df, state != "California", year <  TREAT_YEAR))
Yc_post <- wide_cig(filter(panel_df, state != "California", year >= TREAT_YEAR))

set.seed(SEED)
alpha_draws_hs <- hs_alpha_gibbs_cpp(
  Y0_pre, Yc_pre,
  iteration = MCMC_ITER, burn = MCMC_BURN, verbose = FALSE
)
colnames(alpha_draws_hs) <- donors
```

We then propagate uncertainty through the gap series.

```{r}
#| label: stage2-att
gap_draws_hs <- Y0_post - Yc_post %*% t(alpha_draws_hs)   # T1 x M_draws
att_hs_draws <- colMeans(gap_draws_hs)
att_hs       <- mean(att_hs_draws)
att_hs_ci    <- quantile(att_hs_draws, c(0.025, 0.975), names = FALSE)

glue(
  "Stage 2 ATT (Bayesian HS): {format(round(att_hs, 2), nsmall = 2)} packs/capita, ",
  "95% CrI [{format(round(att_hs_ci[1], 2), nsmall = 2)}, ",
  "{format(round(att_hs_ci[2], 2), nsmall = 2)}]"
)
```

```{r}
#| label: tbl-stage2-weights
#| tbl-cap: "Stage 2 — top-5 donor weights from the horseshoe posterior. Only Nevada's 95% credible interval excludes zero."
alpha_summary <- tibble(
  state = donors,
  mean  = colMeans(alpha_draws_hs),
  lo95  = apply(alpha_draws_hs, 2, quantile, probs = 0.025),
  hi95  = apply(alpha_draws_hs, 2, quantile, probs = 0.975)
) |>
  arrange(desc(abs(mean)))

alpha_summary |>
  head(5) |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor", mean = "Posterior mean",
             lo95 = "2.5%", hi95 = "97.5%")

n_active <- sum(alpha_summary$mean > 0.01)
```

```{r}
#| label: fig-stage2-weights
#| fig-cap: "Stage 2 — posterior mean donor weights with 95% credible intervals, sorted by mean. Most donors hug zero; Nevada is the only top weight whose interval excludes zero."
#| fig-width: 8
#| fig-height: 6
alpha_summary |>
  mutate(state = fct_reorder(state, mean)) |>
  ggplot(aes(x = mean, y = state)) +
  geom_vline(xintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_errorbarh(aes(xmin = lo95, xmax = hi95), height = 0, color = "#6a9bcc") +
  geom_point(color = "#d97757", size = 1.6) +
  labs(x = "Donor weight (posterior mean, 95% CrI)", y = NULL)
```

```{r}
#| label: fig-stage2-paths
#| fig-cap: "Stage 2 — California (orange) vs the horseshoe-posterior-mean synthetic (steel blue), with a 95% credible band on the post-period gap."
#| fig-width: 8
#| fig-height: 5
synth_mean_pre  <- Yc_pre  %*% colMeans(alpha_draws_hs)
synth_mean_post <- Yc_post %*% colMeans(alpha_draws_hs)
post_gap_lo <- apply(gap_draws_hs, 1, quantile, probs = 0.025)
post_gap_hi <- apply(gap_draws_hs, 1, quantile, probs = 0.975)
post_gap_mid <- rowMeans(gap_draws_hs)

paths_hs <- tibble(
  year      = c(years_pre, years_post),
  observed  = c(Y0_pre, Y0_post),
  synthetic = c(as.numeric(synth_mean_pre), as.numeric(synth_mean_post)),
  synth_lo  = c(rep(NA_real_, length(years_pre)), as.numeric(synth_mean_post) - (post_gap_mid - post_gap_lo)),
  synth_hi  = c(rep(NA_real_, length(years_pre)), as.numeric(synth_mean_post) - (post_gap_mid - post_gap_hi))
)

ggplot(paths_hs, aes(x = year)) +
  geom_ribbon(aes(ymin = synth_lo, ymax = synth_hi), fill = "#6a9bcc", alpha = 0.25) +
  geom_line(aes(y = synthetic), color = "#6a9bcc", linewidth = 0.9) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

The donor pool **broadens dramatically**. About `r n_active` of 38 donors carry posterior mean weight above 0.01, versus four under the classical simplex. Crucially, even among the top-five posterior means, only Nevada's 95% credible interval excludes zero — every other top weight is statistically consistent with no contribution. The teaching point is that classical SCM's sparsity is partly a constraint artefact: when we admit posterior uncertainty over weights, the data do not strongly insist on a four-donor synthetic.

The ATT also moves — from $-18.5$ to about $-15.8$ packs/capita — and its 95% credible interval is wider than Stage 1's bootstrap CI by design. The horseshoe propagates donor-weight uncertainty into the gap series rather than treating $\alpha$ as fixed at the optimiser's best guess. The interval still never reaches zero.

## Stage 3 — Bayesian spatial synthetic control with SAR spillovers

Stage 2 relaxed the simplex but kept SUTVA. Stage 3 drops SUTVA by adding a SAR layer to the donor data-generating process:

$$Y_{c,t} \,=\, \rho \, W \, Y_{c,t} \,+\, X_{c,t}\beta \,+\, Y_c^{\text{lag}}\, \alpha \,+\, \varepsilon_t.$$

Each donor state's cigarette sales at time $t$ depend on a row-normalised average of its neighbours' sales (the spatial lag $W Y_{c,t}$ weighted by the autocorrelation parameter $\rho$), on covariates ($X = \text{retprice}$ here), on the donor-side outcomes via the horseshoe weights $\alpha$, and on idiosyncratic noise. When $\rho = 0$ the SAR layer collapses and Stage 3 reduces to Stage 2.

The package's `sc_spillover()` runs both MCMC samplers (horseshoe $\alpha$ then SAR $\rho$) and post-processes the per-state spillover effects in one call.

```{r}
#| label: stage3-fit
#| message: false
fit_sar <- sc_spillover(
  data            = panel_df,
  treated_unit    = "California",
  w = w, W = W,
  treatment_dummy = "treatment",
  y               = "cigsale",
  X               = c("retprice"),
  p_factors       = 1L,
  M               = MCMC_ITER, burn = MCMC_BURN,
  seed            = SEED, step_rho = 0.01,
  unit_col        = "state", time_col = "year",
  verbose         = FALSE
)

rho_hat    <- fit_sar$rho_hat
ess_rho    <- coda::effectiveSize(coda::as.mcmc(fit_sar$rho_draws))[[1]]
att_sar    <- fit_sar$effects$ate_point
att_sar_ci <- fit_sar$effects$ate_ci95
```

```{r}
#| label: stage3-print
glue(
  "Posterior mean rho: {format(round(rho_hat, 3), nsmall = 3)} | ",
  "ESS(rho) = {format(round(ess_rho, 1), nsmall = 1)}\n",
  "Stage 3 ATT (Bayesian Spatial SAR): {format(round(att_sar, 2), nsmall = 2)} ",
  "packs/capita, 95% CrI [{format(round(att_sar_ci[1], 2), nsmall = 2)}, ",
  "{format(round(att_sar_ci[2], 2), nsmall = 2)}]"
)
```

The posterior mean $\hat\rho$ is bounded clearly away from zero — moderate spatial autocorrelation, as the cross-border intuition predicts. The Stage 3 ATT sits between the classical and Bayesian-horseshoe estimates: adding the SAR layer *reattributes* a portion of the gap away from California's direct response toward neighbour spillovers.

:::{.callout-warning}
## ESS(ρ) at tutorial scale is too low for a trustworthy CrI

The chapter runs the sampler at 5,000 iterations to keep render times manageable. At that budget, the effective sample size for $\rho$ is in the single digits — far below the rule-of-thumb 200 — and the 95% credible interval for the Stage 3 ATT is therefore *artificially narrow*. The point estimate is recoverable (a posterior mean has low bias even at low ESS); the interval should be read as illustrative. To match the @sakaguchi2026spatial published numbers, raise `MCMC_ITER` to `100000L` and `MCMC_BURN` to `50000L` and rerender (expect ~30–90 min wall-clock).
:::

```{r}
#| label: fig-stage3-paths
#| fig-cap: "Stage 3 — California (orange) vs the SAR-spillover-corrected synthetic (teal). The treatment effect widens roughly linearly post-1988."
#| fig-width: 8
#| fig-height: 5
ycf_post <- Y0_post - fit_sar$effects$te_point
paths_sar <- tibble(
  year      = years_post,
  observed  = Y0_post,
  synthetic = ycf_post
)

paths_full <- bind_rows(
  tibble(year = years_pre, observed = Y0_pre, synthetic = NA_real_),
  paths_sar
)

ggplot(paths_full, aes(x = year)) +
  geom_line(aes(y = observed),  color = "#d97757", linewidth = 0.9) +
  geom_line(aes(y = synthetic), color = "#00d4c8", linewidth = 0.9) +
  geom_vline(xintercept = TREAT_YEAR, linetype = "dashed", color = "#94a3b8") +
  labs(x = NULL, y = "Per-capita cigarette sales (packs)")
```

### Spillover effects on donor states

The interesting output of the SAR layer is the per-donor spillover. The framework computes the average post-treatment effect on each donor by forward-simulating the SAR DGP with and without California's treatment, integrating over the posterior draws of $\rho$.

```{r}
#| label: tbl-spillover-top8
#| tbl-cap: "Top-8 donor states by absolute post-1988 mean spillover. Nevada absorbs roughly 16x the next state."
spill_mat   <- fit_sar$effects$spill
post_idx    <- which(as.numeric(rownames(spill_mat)) >= TREAT_YEAR)
spill_post  <- spill_mat[post_idx, , drop = FALSE]
spill_top8  <- tibble(state = colnames(spill_mat),
                      avg_spillover = colMeans(spill_post)) |>
  slice_max(abs(avg_spillover), n = 8) |>
  arrange(avg_spillover)

spill_top8 |>
  gt_pretty(decimals = 3) |>
  cols_label(state = "Donor", avg_spillover = "Mean post-1988 spillover")
```

```{r}
#| label: fig-spillover-bars
#| fig-cap: "Top-8 donor states by absolute spillover. Nevada dominates by an order of magnitude — the empirical signature of SUTVA failure for this case study."
#| fig-width: 8
#| fig-height: 4.5
spill_top8 |>
  mutate(state = fct_reorder(state, avg_spillover)) |>
  ggplot(aes(x = avg_spillover, y = state)) +
  geom_vline(xintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_col(fill = "#d97757") +
  labs(x = "Mean post-1988 spillover (packs/capita)", y = NULL)
```

**Nevada is the dominant spillover-receiver by an order of magnitude.** Idaho and Utah trail far behind; every other donor's spillover is small enough to be policy-noise. The story is consistent with cross-border tobacco flows reshaping consumption on both sides of the California–Nevada line — the empirical signature of SUTVA failure for this case study.

## Prior predictive diagnostic

Before reading Stage 3 as a posterior, we want to confirm that the prior specification is *compatible* with what the data actually look like. The helper `prior_predictive()` draws $R = 1{,}000$ joint prior samples, forward-simulates a synthetic donor panel under each draw, computes a battery of summary statistics, and compares them to the observed statistics from the real donor panel.

```{r}
#| label: prior-predictive
Xc_pre_arr <- panel_df |>
  filter(state != "California", year < TREAT_YEAR) |>
  pivot_wider(id_cols = year, names_from = state, values_from = retprice) |>
  select(-year) |> select(all_of(donors)) |> as.matrix()
dim(Xc_pre_arr) <- c(nrow(Xc_pre_arr), ncol(Xc_pre_arr), 1)

ppc <- prior_predictive(
  Y0_pre           = as.matrix(Y0_pre),
  Yc_obs           = Yc_pre,
  W_raw            = W,
  w_raw            = w,
  alpha_hat_scaled = colMeans(fit_sar$alpha_draws),
  Xc_pre           = Xc_pre_arr,
  p                = 0L,
  a0 = 3, b0 = 1, rho_support = c(-0.99, 0.99),
  R                = 1000L,
  seed             = SEED
)
```

```{r}
#| label: fig-prior-predictive
#| fig-cap: "Prior predictive check on four summary statistics. Histograms are R = 1,000 simulated prior draws; the orange line is the observed value. All four observed lines land inside the simulated prior cloud — the prior is compatible with the data, not overwhelming it."
#| fig-width: 9
#| fig-height: 6
stats_show <- c("yc\\_mean", "spatial\\_quadratic", "ac1", "pve\\_pc1")
pretty <- c("Donor mean", "Spatial y'Wy", "Lag-1 autocorrelation", "PC1 variance share")

ppc_long <- ppc$stat |>
  as_tibble() |>
  select(all_of(stats_show)) |>
  rename_with(~ pretty, all_of(stats_show)) |>
  pivot_longer(everything(), names_to = "stat", values_to = "value")
obs_df <- tibble(
  stat  = pretty,
  value = unname(ppc$observed[stats_show])
)

ggplot(ppc_long, aes(value)) +
  geom_histogram(bins = 30, fill = "#6a9bcc", color = "white") +
  geom_vline(data = obs_df, aes(xintercept = value),
             color = "#d97757", linewidth = 0.8) +
  facet_wrap(~ stat, scales = "free", ncol = 2) +
  labs(x = NULL, y = "Prior-draw count")
```

All four observed orange lines land *inside* the simulated prior cloud rather than in the tails: the prior is compatible with the data, not overwhelming it. Had any observed statistic landed deep in a prior tail, the corresponding posterior estimate would have been pulled by the prior rather than the likelihood.

## Cross-stage comparison

Stacking the three estimators makes the pedagogical arc visible.

```{r}
#| label: tbl-cross-stage
#| tbl-cap: "Cross-stage comparison of ATT estimates. The active-donor count rises monotonically as the prior structure relaxes; the Stage 3 CrI is artificially narrow because ESS(rho) is in single digits at tutorial scale."
tibble(
  Stage          = c("Classical SCM (tidysynth)",
                     "Bayesian HS (no spillovers)",
                     "Bayesian Spatial SAR (with spillovers)"),
  ATT            = c(att_classic, att_hs, att_sar),
  `2.5%`         = c(NA_real_,    att_hs_ci[1], att_sar_ci[1]),
  `97.5%`        = c(NA_real_,    att_hs_ci[2], att_sar_ci[2]),
  `Active donors`= c(sum(grab_unit_weights(sc_classic)$weight > 0.01),
                     sum(alpha_summary$mean > 0.01),
                     sum(colMeans(fit_sar$alpha_draws) > 0.01)),
  `ESS(rho)`     = c(NA_real_, NA_real_, ess_rho)
) |>
  gt_pretty(decimals = 2)
```

Three observations.

- **Sign and order of magnitude agree.** All three estimators place the ATT between roughly $-15$ and $-19$ packs/capita per year, and none of the intervals reaches zero. Whatever one believes about the simplex or SUTVA, Proposition 99 reduced California cigarette consumption.
- **Active-donor count rises monotonically.** Heavier-tailed priors admit more donors with non-trivial mass; the sparse four-donor synthetic of Stage 1 looks like one plausible counterfactual rather than the right one.
- **The Stage 3 credible interval is the narrowest but the least trustworthy.** ESS(ρ) is in single digits; downstream prose should treat that interval as illustrative.

## Recap

| Question | Answer |
|---|---|
| What does this chapter estimate? | The ATT of Proposition 99 on California cigarette sales, 1988–2000 |
| Which assumption does Stage 2 relax? | The simplex constraint on donor weights — replaced by a horseshoe prior |
| Which assumption does Stage 3 relax? | SUTVA — donor outcomes are now allowed to respond to the treatment via a SAR layer |
| Where does the spillover land? | Almost entirely on Nevada ($\approx -3.75$ packs/capita); other donors are an order of magnitude smaller |
| What is the chapter's caveat? | At 5,000 iterations the ESS for ρ is too small for the Stage 3 CrI to be trustworthy — raise iterations for paper-grade inference |
| Is the headline robust? | Yes — all three ATT estimates are negative, never cross zero, and agree on order of magnitude |

## Further reading

- @sakaguchi2026spatial — the original Bayesian spatial-SCM paper this chapter replicates.
- @carvalho2010horseshoe — the horseshoe prior used in Stage 2.
- @lesage2009introduction — textbook treatment of SAR and spatial econometrics.
- @abadie2010synthetic — the original synthetic-control treatment of Proposition 99, replicated in chapter 5.
- @dunford2024tidysynth — `tidysynth` package documentation.
