---
title: "Staggered Differences-in-Differences"
---

## When TWFE breaks under staggered adoption

Chapter 3 ran a textbook 2×2 DiD on Proposition 99, with California
treated in 1989 and Nevada as the single hand-picked control. The
estimate collapsed to roughly $-5.7$ packs per capita and could not
be distinguished from zero. The diagnosis there was that Nevada is
geographically and culturally adjacent to California and absorbs the
same secular forces — so its post-1988 decline soaks up most of what
we wanted to attribute to the policy.

There is a deeper structural problem too. Real policy data rarely
has just one treated unit and one clean control unit, both treated
at the same instant. States and counties **adopt at different times**
— some in 2004, others in 2006, still others never. The natural
extension of the chapter 3 model to that setting is *two-way fixed
effects (TWFE)* regression:

$$Y_{it} = \alpha_i + \gamma_t + \beta \cdot \text{post}_{it} + \varepsilon_{it}.$$

For two decades this was the default. We now know it is biased in the
presence of staggered adoption: already-treated units silently act as
*controls* for later-treated units, contaminating the contrast. The
@goodmanbacon2021difference decomposition shows that $\hat\beta$ is a
weighted average of 2×2 DiDs in which already-treated units serve as
controls for later-treated units — the *forbidden comparison*.
@dechaisemartin2020twoway further prove that under treatment-effect
heterogeneity some of those implicit weights can be **strictly
negative**, so $\hat\beta$ need not even be a convex average of the
underlying $ATT$s. Either way, when treatment effects grow over time —
the textbook policy story — the sign of $\hat\beta$ is no longer a
reliable summary.

This chapter walks through the modern toolkit for repairing that
damage. The methods build on @callaway2021difference: estimate
**group-time average treatment effects** $ATT(g, t)$ directly, then
aggregate them with weights you can defend. Along the way we look at
two companion ideas: the doubly-robust DiD estimator from
@callaway2021difference, and the Rambachan-Roth sensitivity analysis
that quantifies how much parallel-trends violation it would take to
overturn the conclusion [@rambachan2023more]. The Sun-Abraham
interaction-weighted event study [@sun2021estimating] is an equivalent
estimator built on a different regression-based decomposition; see
@roth2023whats for a side-by-side review.

The dataset is no longer Proposition 99. Staggered DiD requires
**variation in treatment timing**, which a single-treated-state panel
cannot provide. We switch to the Callaway-Sant'Anna minimum-wage
panel: 1,745 US counties × 2003–2007, with cohorts $G \in \{0, 2004,
2006\}$ (after dropping a small late-2007 cohort; see Setup)
indexing the year the county's state first raised its minimum
wage above the federal $5.15/h floor. The outcome is log teen
employment.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source helpers, and set the ggplot theme."
library(tidyverse)
library(did)
library(fixest)
library(HonestDiD)
source("R/table_helpers.R")
source("R/honest_did.R")

set.seed(42)  # all chunks below are analytically deterministic; this is just hygiene

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8"),
      strip.text       = element_text(color = "#94a3b8"),
      legend.text      = element_text(color = "#94a3b8")
    )
)
```

The dataset ships as `data/cs_minwage.rds` in the chapter bundle. We
follow the source-post convention of restricting to cohorts $G \in
\{0, 2004, 2006, 2007\}$, dropping the Northeast region (`region ==
"1"`) for comparability, and then carving out a clean working panel
without the late-2007 cohort and starting in 2003.

```{r}
#| label: data-load
#| code-summary: "Code: Load the minimum-wage panel and build the working sample."
mw_raw <- readRDS("data/cs_minwage.rds") |> as_tibble()

# Step 1: drop Northeast and keep cohorts of interest.
mw <- mw_raw |>
  filter(G %in% c(0, 2004, 2006, 2007), region != "1")

# Step 2: working sample for the main analysis.
data2 <- mw |>
  filter(G != 2007, year >= 2003)

dim(data2)
```

The working panel has `r nrow(data2)` rows =
`r length(unique(data2$id))` counties × 5 years, balanced across the
2003–2007 window.

```{r}
#| label: tbl-cohort-counts
#| tbl-cap: "Cohort sizes in the working sample. G = 0 is the never-treated control pool; cohorts 2004 and 2006 are the staggered treated groups."
#| code-summary: "Code: Tabulate county counts by treatment cohort."
data2 |>
  filter(year == 2003) |>
  count(G, name = "counties") |>
  rename(`Treatment cohort (G)` = G) |>
  gt_pretty()
```

## The TWFE baseline

The natural first move is the TWFE regression. The variable `post` in
the dataset is 1 in periods where the county's state has already
raised its minimum wage (i.e., `year >= G & G != 0`), and 0 otherwise.

```{r}
#| label: tbl-twfe
#| tbl-cap: "Two-way fixed-effects regression. The point estimate suggests minimum-wage increases cut log teen employment by roughly 3.8 percent — but see the discussion below."
#| code-summary: "Code: Estimate the TWFE baseline regression with clustered SEs."
twfe_res <- fixest::feols(lemp ~ post | id + year,
                          data = data2, cluster = "id")
ms_pretty(list("TWFE (county + year FE)" = twfe_res),
          coef_map = c("post" = "Post (any cohort)"),
          notes    = "SEs clustered at the county level.")
```

The TWFE coefficient is roughly $-0.038$. Read literally, the policy
reduced teen employment by 3.8 percent. The next section estimates
the same effect with an aggregator that is unbiased under staggered
adoption — and gets a substantially larger magnitude. The gap between
the two is the contamination problem made concrete.

## Group-time ATTs: the Callaway-Sant'Anna approach

The fix is to estimate the *primitive* objects directly. For each
cohort $g$ (a year a group of counties was first treated) and each
calendar year $t$, define

$$ATT(g, t) = \mathbb{E}\!\left[Y_{it}(g) - Y_{it}(\infty) \mid G_i = g\right],$$

the average effect on cohort $g$ in year $t$ relative to its own
never-treated potential outcome $Y_{it}(\infty)$ (equivalent to
$Y_{it}(0)$ used in Part I — Callaway-Sant'Anna write $\infty$ to
emphasise that the unit is *never* treated, not merely *not-yet*
treated).

Identification relies on **parallel trends**: conditional on
$G_i = g$, the never-treated potential outcome
$E[Y_{it}(\infty)]$ trends the same way for the treated cohort and
the never-treated pool. Under that assumption, `did::att_gt()`
estimates each $ATT(g, t)$ from a clean 2×2 DiD using only cohort
$g$ and an appropriate comparison group (here, the never-treated
$G = 0$), so no contamination from already-treated units sneaks in.

```{r}
#| label: tbl-attgt
#| tbl-cap: "Group-time average treatment effects $ATT(g, t)$ for the minimum-wage panel. Cohorts 2004 and 2006, each year 2003 through 2007. Pre-treatment cells should hover near zero if parallel trends holds; post-treatment cells are the effects we want."
#| code-summary: "Code: Estimate group-time ATTs with the did package."
attgt <- did::att_gt(yname = "lemp", idname = "id", gname = "G",
                     tname = "year", data = data2,
                     control_group = "nevertreated",
                     base_period = "universal")

attgt_df <- tibble(
  Cohort       = attgt$group,
  Year         = attgt$t,
  `ATT(g,t)`   = attgt$att,
  SE           = as.numeric(attgt$se)
) |>
  mutate(`Treated?` = ifelse(Year >= Cohort, "post", "pre")) |>
  arrange(Cohort, Year)

gt_pretty(attgt_df, decimals = 4)
```

Cohort 2006 carries a visible pre-trend: $ATT(2006, 2003) \approx
-0.034$ with SE $\approx 0.013$ (so $t \approx -2.7$). The 2006
counties were already on a downward log-employment path relative to
the never-treated pool three years *before* their state raised the
minimum wage. That pre-treatment violation is roughly the same
magnitude as the on-impact effect we will estimate below — which is
exactly what motivates the formal sensitivity analysis at the end of
the chapter. Naming the pre-trend now keeps the rest of the story
honest.

### Aggregation: overall ATT and event study

The 8 cells in @tbl-attgt are the primitives. They are not the
headline. Aggregating them produces summaries that are valid in the
presence of staggered adoption.

The **overall ATT** first averages each cohort's post-treatment
$ATT(g, t)$ values into a single cohort-specific summary, then
averages those summaries across cohorts weighted by cohort size. It
answers: *across treated cohorts, what is the average post-treatment
effect on a typical treated county?* This is the
@callaway2021difference recommended summary; it differs from a plain
mean of post-treatment cells (`type = "simple"`) when cohorts have
different post-treatment horizons — as they do here (cohort 2004 has
4 post cells, cohort 2006 has 2).

```{r}
#| label: tbl-cs-overall
#| tbl-cap: "Callaway-Sant'Anna overall ATT (sample-weighted aggregation across cohorts). This is the staggered-DiD analogue of chapter 3's single DiD coefficient."
#| code-summary: "Code: Aggregate group-time ATTs into the overall ATT summary."
attO <- did::aggte(attgt, type = "group")

tibble(
  Estimator             = "Callaway-Sant'Anna overall ATT",
  `Estimate`            = attO$overall.att,
  `SE`                  = attO$overall.se,
  `CI lower`            = attO$overall.att - 1.96 * attO$overall.se,
  `CI upper`            = attO$overall.att + 1.96 * attO$overall.se
) |> gt_pretty(decimals = 4)
```

The overall ATT is roughly $-0.057$ — almost 50 percent larger in
magnitude than the TWFE coefficient. The two estimands answer
different questions, but the size of the gap is exactly the
contamination problem in operation.

The **event-study aggregation** averages $ATT(g, t)$ within each
event time $e = t - g$, giving a curve of dynamic effects since
treatment.

```{r}
#| label: tbl-cs-event
#| tbl-cap: "Callaway-Sant'Anna event-study aggregation. Event time -1 is the omitted reference period under the universal base."
#| code-summary: "Code: Aggregate ATTs into the dynamic event-study table."
attes <- did::aggte(attgt, type = "dynamic")

tibble(
  `Event time` = attes$egt,
  `ATT(e)`     = attes$att.egt,
  SE           = attes$se.egt,
  `CI lower`   = attes$att.egt - 1.96 * attes$se.egt,
  `CI upper`   = attes$att.egt + 1.96 * attes$se.egt
) |> gt_pretty(decimals = 4)
```

```{r}
#| label: fig-cs-event
#| fig-cap: "Callaway-Sant'Anna event study. The on-impact effect is small; effects accumulate to roughly -0.13 log points by event-time +3 (three years post-treatment)."
#| fig-width: 8
#| fig-height: 4.5
#| code-summary: "Code: Plot the event-study estimates with 95 percent CIs."
cs_es_df <- tibble(
  egt   = attes$egt,
  est   = attes$att.egt,
  se    = attes$se.egt,
  phase = ifelse(attes$egt < 0, "Pre", "Post")
)

ggplot(cs_es_df, aes(x = egt, y = est, color = phase)) +
  geom_hline(yintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_vline(xintercept = -0.5, color = "#94a3b8", linetype = "dashed") +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = est - 1.96 * se, ymax = est + 1.96 * se),
                width = 0.15) +
  scale_color_manual(values = c("Pre"  = "#e2e8f0",
                                "Post" = "#22d3ee")) +
  labs(x = "Event time (years from first minimum-wage increase)",
       y = "Effect on log teen employment", color = NULL)
```

## Conditional parallel trends

Plain DiD assumes parallel trends *unconditionally*. Conditional DiD
allows trends to be parallel only after adjusting for observed
covariates that shape the trajectory. In this dataset, log county
population and log average county pay are obvious candidates: counties
with bigger populations or higher-paying jobs may move on different
underlying trajectories whether or not they raise the minimum wage.

`did::att_gt()` accepts `xformla = ~lpop + lavg_pay` plus a choice of
estimator: **regression adjustment** (impute the counterfactual mean
from the never-treated), **inverse probability weighting** (reweight
never-treated counties to look like treated ones), or **doubly
robust** (the @callaway2021difference default, which is consistent
if either the outcome model *or* the propensity score is correctly
specified).

```{r}
#| label: tbl-conditional
#| tbl-cap: "Overall ATT under four identification assumptions: unconditional parallel trends, regression adjustment, inverse probability weighting, and doubly robust. The four point estimates land within roughly one SE of each other, which is reassuring."
#| message: false
#| warning: false
#| code-summary: "Code: Compare overall ATT under regression, IPW, and doubly-robust adjustments."

cs_reg <- did::att_gt(yname="lemp", tname="year", idname="id", gname="G",
                      xformla = ~lpop + lavg_pay,
                      control_group = "nevertreated",
                      base_period = "universal",
                      est_method = "reg", data = data2)
attO_reg <- did::aggte(cs_reg, type = "group")

cs_ipw <- did::att_gt(yname="lemp", tname="year", idname="id", gname="G",
                      xformla = ~lpop + lavg_pay,
                      control_group = "nevertreated",
                      base_period = "universal",
                      est_method = "ipw", data = data2)
attO_ipw <- did::aggte(cs_ipw, type = "group")

cs_dr  <- did::att_gt(yname="lemp", tname="year", idname="id", gname="G",
                      xformla = ~lpop + lavg_pay,
                      control_group = "nevertreated",
                      base_period = "universal",
                      est_method = "dr", data = data2)
attO_dr  <- did::aggte(cs_dr, type = "group")

tibble(
  Identification = c("Unconditional parallel trends",
                     "Conditional · regression adjustment",
                     "Conditional · IPW",
                     "Conditional · doubly robust"),
  Estimate = c(attO$overall.att, attO_reg$overall.att,
               attO_ipw$overall.att, attO_dr$overall.att),
  SE       = c(attO$overall.se,  attO_reg$overall.se,
               attO_ipw$overall.se,  attO_dr$overall.se)
) |> gt_pretty(decimals = 4)
```

The doubly-robust estimate is the most reliable in practice — it
recovers a correct ATT if either the trend model or the selection
model is right, where the other two require their respective model
to be correctly specified.

## Sensitivity to parallel-trends violations

The robustness table only varies *design* choices; it does not
quantify *how much* parallel trends would have to be violated to
overturn the conclusion. @rambachan2023more provide exactly that
quantification. The `HonestDiD` package, paired with a small bridge
function in `R/honest_did.R`, lets us bound the post-treatment effect
under explicit assumptions about how an unobserved post-treatment
violation could relate to the visible pre-trend.

`HonestDiD` ships two restrictions on the post-treatment violation.
The **smoothness** bound $\Delta^{SD}(M)$ requires the post-treatment
violation to be a smooth continuation of the pre-trend (it caps the
*second differences* of the trend), which is the natural choice when
pre-trends look like near-linear drift. The **relative-magnitudes**
bound $\Delta^{RM}(\bar M)$ requires the post-treatment violation to
be at most $\bar M$ times the *largest* pre-treatment violation; this
is the natural choice when the pre-trend is visibly non-linear, as in
cohort 2006 above. We use the relative-magnitudes bound below.

```{r}
#| label: tbl-honest
#| tbl-cap: "Robust confidence intervals for the on-impact effect $ATT(e = 0)$ under the relative-magnitude restriction. As $\\bar M$ grows, the CI widens; we report the breakdown point — the $\\bar M$ at which the CI first contains zero."
#| message: false
#| warning: false
#| code-summary: "Code: Run HonestDiD relative-magnitude sensitivity on the on-impact ATT."

hd_rm <- honest_did(es      = attes,
                    e       = 0,
                    type    = "relative_magnitude",
                    method  = "C-LF",
                    Mbarvec = seq(0, 2, by = 0.5))

hd_rm |>
  transmute(`$\\bar M$` = Mbar,
            `CI lower`  = lb,
            `CI upper`  = ub,
            `Contains 0?` = ifelse(lb <= 0 & 0 <= ub, "yes", "no")) |>
  gt_pretty(decimals = 4)
```

```{r}
#| label: fig-honest
#| fig-cap: "HonestDiD sensitivity: as $\\bar M$ grows, the robust 95 percent CI on the on-impact effect widens until it crosses zero. The breakdown $\\bar M$ — the value at which the CI first contains zero — is below 0.5, reflecting the visible pre-trend in cohort 2006."
#| fig-width: 7.5
#| fig-height: 4.5
#| code-summary: "Code: Plot the robust CI band against the relative-magnitude bound."
ggplot(hd_rm, aes(x = Mbar)) +
  geom_hline(yintercept = 0, color = "#94a3b8", linetype = "dashed") +
  geom_ribbon(aes(ymin = lb, ymax = ub),
              fill = "#22d3ee", alpha = 0.2) +
  geom_line(aes(y = lb), color = "#22d3ee", linewidth = 0.9) +
  geom_line(aes(y = ub), color = "#22d3ee", linewidth = 0.9) +
  labs(x = expression(bar(M)),
       y = "Robust 95% CI on ATT(e = 0)")
```

The breakdown point lies somewhere in $(0, 0.5]$ — meaning the
robust CI first crosses zero at an $\bar M$ less than half the
*largest observed pre-trend in the data*. The on-impact effect of
$\approx -0.024$ does **not** survive even a fraction of the
cohort-2006 pre-trend we already flagged: that pre-trend, at
$-0.034$, is itself larger in magnitude than the on-impact effect,
so it would be surprising if the conclusion held up under a
relative-magnitudes assumption that allowed unobserved post-trend
violations of comparable size.

This is not a failure of the method — it is the method working as
intended. HonestDiD is the bridge from "the Callaway-Sant'Anna point
estimate looks negative" to "here is precisely the parallel-trends
violation under which that conclusion would flip." The right reading
is that the on-impact effect is **fragile to parallel-trends
violations**, even though the *cumulative* event-study trajectory
($-0.13$ by $e = +3$) is large enough that a similar exercise on the
later horizons would tolerate a much larger $\bar M$.

## Recap

::: {.callout-note appearance="simple"}
**The methods reconciled.** TWFE on this dataset returned
$\hat\beta = -0.038$. The Callaway-Sant'Anna overall ATT is
$-0.057$. The doubly-robust conditional ATT is $-0.065$. The
event-study trajectory is $\approx -0.024$ on impact, dropping to
$\approx -0.13$ by event-time $+3$. HonestDiD sensitivity is
*fragile*: the relative-magnitudes breakdown $\bar M$ for the
on-impact effect is **below 0.5**, reflecting the visible
cohort-2006 pre-trend. Four estimators agree on point estimate and
sign — *minimum-wage increases reduced teen employment in these
counties, and the effect grew over time* — but parallel trends does
not hold cleanly, and the on-impact effect would not survive even
half the observed pre-trend violation.

The gap between the TWFE estimate and the modern aggregators is the
contamination problem of @goodmanbacon2021difference made concrete:
TWFE silently uses already-treated units as controls for later-treated
units, and treatment-effect heterogeneity over time leaks into the
coefficient with unintended signs. The Callaway-Sant'Anna aggregator
avoids that by estimating the $ATT(g, t)$ primitives directly and
weighting them in a defensible way.

Every method in this chapter — TWFE, $ATT(g, t)$, conditional DiD,
even HonestDiD's bounds — leans on **parallel trends** as the
identifying assumption. The next chapter relaxes that assumption by
modelling $Y_{it}(0)$ with an *interactive fixed-effects* factor
structure:

$$Y_{it}(0) = \alpha_i + \xi_t + \lambda_i' f_t + \varepsilon_{it}.$$

Two estimators — matrix completion and IFEct — implement that idea on
the same minimum-wage panel and provide a second, assumption-distinct
read on the same substantive question.
:::

## Common pitfall

Running TWFE on staggered data and reporting the coefficient as if
it were a clean ATT. The bias is mechanical — already-treated units
get used as controls for later-treated units, and treatment-effect
heterogeneity over time then leaks into the coefficient with
unintended signs. **What to do instead.** Estimate the
$\{ATT(g, t)\}$ primitives directly with `did::att_gt()`, look at
the cells, and only then aggregate with `aggte()` using a target
that matches the question you actually want to answer. If you must
run TWFE for a referee, also report the Callaway-Sant'Anna overall
ATT for comparison so the magnitude of the contamination is visible.

## Further reading

The Callaway-Sant'Anna framework, the Goodman-Bacon decomposition,
and the Sun-Abraham interaction-weighted estimator are the three
modern reference points for staggered DiD
[@callaway2021difference; @goodmanbacon2021difference;
@sun2021estimating]. The `did` package vignettes
(<https://bcallaway11.github.io/did/>) are the canonical
implementation reference. For sensitivity analysis, the
@rambachan2023more paper plus the `HonestDiD` package documentation
cover both the smoothness and relative-magnitude bounds.
@dechaisemartin2020twoway is the parallel critique of TWFE from the
`DIDmultiplegt` perspective. @callaway2022handbook is a textbook-level
synthesis, and @roth2023whats is a recent review-style synthesis
covering staggered DiD, event studies, sensitivity analysis, and the
relationships between them.

For a longer R walkthrough that this chapter is adapted from, see
the companion post at
<https://cmg777.github.io/post/r_did/>.

## Exercises

1. Re-estimate the overall ATT using `control_group = "notyettreated"`
   instead of `"nevertreated"`. Does the answer change? Explain why
   the standard error usually *shrinks* under this alternative, and
   what assumption about not-yet-treated counties you are now relying
   on to make their comparison valid.
2. Use `did::aggte(attgt, type = "calendar")` to aggregate by
   *calendar year* rather than event time. Which calendar year shows
   the largest treatment effect, and what substantive story does that
   suggest?
3. Run `HonestDiD` with `type = "smoothness"` instead of
   `"relative_magnitude"`, supplying `Mvec = seq(0, 0.05, by = 0.01)`.
   What does the breakdown $M$ say about the credibility of the
   parallel-trends assumption in this dataset?
