---
title: "Interactive Fixed Effects and Matrix Completion"
---

## When parallel trends isn't enough

Chapter 8 squeezed every drop of usable signal out of the
Callaway-Sant'Anna minimum-wage panel under one identifying
assumption: **parallel trends**. Conditional or unconditional, with
or without never-treated counties, with or without anticipation
slack, with Rambachan-Roth bounds on top — the assumption is always
that, absent treatment, the trajectory of the treated counties would
have moved *in parallel* with the untreated.

What if it would not have? Counties select into raising the minimum
wage non-randomly. They are not a random subset of the country in
the time-varying dimension either: industry mix, demographics, and
local growth trends all shape both treatment timing and outcome
paths. Two counties can share an industry-mix-driven trajectory that
TWFE never absorbs — and so cannot net out.

This chapter relaxes parallel trends in a specific way. We model the
counterfactual outcome as a **factor structure**:

$$Y_{it}(0) \;=\; \alpha_i + \xi_t + \lambda_i' f_t + \varepsilon_{it}.$$

Each unit $i$ has a loading vector $\lambda_i$; each period a factor
vector $f_t$. Their interaction $\lambda_i' f_t$ encodes
time-varying unobserved heterogeneity that no county or year fixed
effect can absorb. We write the model without observable covariates
for simplicity, but `fect()` accepts a covariate matrix
$X_{it}$ — e.g. `lemp ~ D + lpop + lavg_pay` — and chapter 10
exercises exactly that extension on the same panel.

Two views of the same object give two estimators
[@liu2024practical]:

- **Interactive Fixed Effects (IFEct)** [@bai2009panel;
  @xu2017generalized; @liu2024practical] treats
  $(\alpha, \xi, \Lambda, F)$ as parameters to be estimated on
  control observations, then imputes $Y_{it}(0)$ on treated cells.
- **Matrix Completion (MC)** [@athey2021matrix] reframes the same
  problem as filling missing entries of the implicit $Y(0)$ matrix:
  mask the treated cells, then complete via nuclear-norm-regularised
  SVD.

Yiqing Xu's `fect` [@liu2024practical] package implements both with
a common API and a common counterfactual-plot diagnostic. By
chapter's end we compare IFEct and MC on the same panel — two
estimators, two identifying assumptions, one substantive question.
Chapter 10 then does the deep dive on one specific member of this
family — generalized synthetic control via the `gsynth` package —
using the same panel.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the ggplot theme."
library(tidyverse)
library(fect)
library(panelView)
library(patchwork)
source("R/table_helpers.R")

set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8"),
      strip.text       = element_text(color = "#94a3b8"),
      legend.text      = element_text(color = "#94a3b8")
    )
)
```

We work from the same `cs_minwage.rds` panel as chapter 8, but with
one small adjustment: factor-based estimators need **at least as
many pre-treatment periods per cohort as the number of factors they
try to fit** — concretely, identifying $r$ factors on a treated unit
requires at least $r + 1$ pre-treatment periods for that unit.
Chapter 8's 2003-2007 `data2` window (see
`08-staggered-did.qmd:99`) leaves the 2004 cohort with only one
pre-period (year 2003). To keep that cohort in the IFE sample we
would need to set `min.T0 = 1` — the user-controlled inclusion
threshold that drops treated units with fewer pre-periods — which
in turn caps the identifiable rank at zero for that cohort and
collapses IFEct back to TWFE. Widening the window to 2001-2007
gives 2004 three pre-periods (2001-2003), so we can set
`min.T0 = 2` and let CV search over $r \in \{0, 1, 2\}$. Everything
else (cohort filter, region drop, treatment indicator) is identical
to chapter 8.

```{r}
#| label: data-load
#| code-summary: "Code: Load the minimum-wage panel and restrict to the 2001-2007 working window."
mw_raw <- readRDS("data/cs_minwage.rds") |> as_tibble()

mw <- mw_raw |>
  filter(G %in% c(0, 2004, 2006, 2007), region != "1") |>
  filter(G != 2007, year >= 2001) |>
  mutate(D = as.integer(year >= G & G != 0))

dim(mw)
```

```{r}
#| label: tbl-cohorts
#| tbl-cap: "Cohorts in the 2001-2007 working panel. $G = 0$ is the never-treated control pool; the two treated cohorts have 3 and 5 pre-treatment years respectively."
#| code-summary: "Code: Summarize county counts and pre-treatment periods by cohort."
mw |>
  filter(year == 2001) |>
  count(G, name = "counties") |>
  mutate(`Pre-periods` = case_when(
    G == 0    ~ NA_integer_,
    G == 2004 ~ 3L,
    G == 2006 ~ 5L
  )) |>
  rename(`Cohort (G)` = G) |>
  gt_pretty()
```

The outcome `lemp` (log teen employment; see chapter 8 for full
data provenance) and the treatment indicator `D` (1 once a county's
state has raised its minimum wage above the federal floor, 0
otherwise) line up with chapter 8's definitions.

## Visualising the panel

`panelView` is the natural opener for any FECT workflow. It draws
units on the vertical axis and time on the horizontal, coloured by
treatment status, so the staggered-adoption structure of the panel
is immediate.

```{r}
#| label: fig-panelview
#| fig-cap: "Treatment status by county and year. The two horizontal bands of pink are the 2004 and 2006 cohorts; the wide grey band underneath is the never-treated $G = 0$ control pool. The visible step in the pink rows is the staggered adoption that breaks TWFE. All 1,745 counties are shown."
#| fig-width: 8
#| fig-height: 6
#| message: false
#| warning: false
#| code-summary: "Code: Visualize the staggered treatment status across counties with panelView."
panelview(data = as.data.frame(mw), formula = lemp ~ D,
          index = c("id", "year"),
          xlab = "Year", ylab = "County",
          main = "", legendOff = FALSE,
          display.all = TRUE,
          theme.bw = TRUE,
          background = "transparent")
```

## The factor model of counterfactuals

Both estimators target the same object — the counterfactual matrix
$Y(0)$ of what each county's log teen employment *would* have been
absent any minimum-wage increase. Where they differ is in how they
restrict that matrix.

**IFEct** imposes the explicit factor decomposition $Y_{it}(0) =
\alpha_i + \xi_t + \lambda_i' f_t + \varepsilon_{it}$, fits
$(\alpha, \xi, \Lambda, F)$ on the **control** observations only,
and uses those estimates to impute $Y_{it}(0)$ for treated
cells. The choice of $r$ — how many factors — is made by
cross-validation: hold out small blocks of control cells, refit,
score by MSPE on the held-out cells [@xu2017generalized;
@liu2024practical].

**MC** does not write the factor model down. It assumes only that
the matrix of $Y(0)$ outcomes is **approximately low rank**, then
completes it by minimising a Frobenius-norm fit penalty *plus* a
nuclear-norm penalty on the singular values. Nuclear-norm
regularisation is the convex relaxation of "low rank" the same way
$\ell_1$ is the convex relaxation of "sparse" [@athey2021matrix].
Concretely, MC solves
$$\min_{\widehat{Y(0)}} \; \|\,Y_{\mathrm{obs}} - \widehat{Y(0)}\,\|_F^2 \;+\; \eta \, \|\,\widehat{Y(0)}\,\|_*,$$
where $\|\cdot\|_*$ (the nuclear norm) is the sum of singular
values. As $\eta \to 0$ the fit interpolates the observed cells; as
$\eta \to \infty$ all singular values shrink to zero and
$\widehat{Y(0)}$ collapses to a constant. The penalty weight $\eta$
(written as `lambda` in the `fect` API and sometimes as
$\lambda_{\mathrm{MC}}$ in the matrix-completion literature; we use
$\eta$ in prose to avoid confusion with the unit-level loading
vector $\lambda_i$ from IFEct) plays the role of $r$ and is again
chosen by cross-validation.

The key practical implication: both methods bake in
**unit-specific time trends** automatically, without you having to
specify which units or which trend shapes. Parallel trends becomes
a special case (the $r = 0$ corner of IFEct, or the limit
$\eta \to \infty$ in MC) rather than an assumption.

## Estimating with FECT

::: {.callout-warning appearance="simple"}
**The short-panel caveat (read first).** With $T = 7$ and 3
pre-periods on the shorter cohort, the rank/penalty selected by
cross-validation is borderline-identifiable; we cap $r \le 2$
deliberately and the *effective* identification ceiling is
$r \le \min(\text{pre-periods}) - 1 = 1$. On a genuinely short
panel ($T \le 5$) IFE and MC become numerically delicate, and the
honest move is to *not* run them rather than to hand-pick a rank.
The `fect::simdata` panel ships a $T = 30$ example that shows both
methods working at full strength. Everything in the rest of this
chapter should be read with this caveat in mind.
:::

The two fits below cap the rank grid at $r \in \{0, 1, 2\}$. The
identification rule is that recovering $r$ factors on a treated
unit requires at least $r + 1$ pre-periods for that unit; with
`min.T0 = 2` the *effective* ceiling is $r = 1$. We let CV search
up to $r = 2$ only to verify it does not get fooled into climbing
past the ceiling. Any $r$ larger than that would be statistical
fantasy.

::: {.callout-note appearance="simple"}
**Learning-mode compute.** We set `nboots = 50` here so the chapter
renders quickly. The bootstrap dominates runtime and only affects
the *width* of confidence intervals — point estimates and the
CV-selected rank or penalty are unchanged. For research use, bump
`nboots` to 500-1000 so the bootstrap distribution of the ATT is
properly resolved.
:::

```{r}
#| label: fit-ife
#| message: false
#| warning: false
#| code-summary: "Code: Fit the interactive fixed effects estimator with fect and CV-selected rank."
out_ife <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "ife",
  force    = "two-way",
  CV       = TRUE,
  r        = 0:2,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)
```

```{r}
#| label: fit-mc
#| message: false
#| warning: false
#| code-summary: "Code: Fit the matrix-completion estimator with fect and CV-selected nuclear-norm penalty."
out_mc <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "mc",
  force    = "two-way",
  CV       = TRUE,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)
```

```{r}
#| label: tbl-cv
#| tbl-cap: "Hyperparameters selected by cross-validation. IFEct picks the number of latent factors $r$; MC picks the nuclear-norm penalty weight $\\eta$ (returned by `fect` as `lambda.cv`)."
#| code-summary: "Code: Report the CV-selected rank for IFEct and nuclear-norm penalty for MC."
tibble(
  Method            = c("IFEct", "MC"),
  `CV-selected`     = c(paste0("r = ", unname(out_ife$r.cv)),
                        sprintf("eta = %.4f", out_mc$lambda.cv))
) |>
  gt_pretty()
```

## Counterfactual paths and the ATT trajectory

The signature FECT figure overlays the observed average outcome on
treated units against the model-implied $Y(0)$, then plots the gap
— the ATT path — by event time. A credible factor model should
track the observed $Y$ closely before treatment and only diverge
after.

```{r}
#| label: fig-ife-mc
#| fig-cap: "Counterfactual trajectories and event-study ATT. Top row: IFEct. Bottom row: MC. Left panels overlay observed average outcome on treated units (solid) with model-implied $Y(0)$ (dashed). Right panels show the implied ATT by event time relative to a county's adoption year, with bootstrap 95% CIs."
#| fig-width: 9
#| fig-height: 7
#| code-summary: "Code: Plot counterfactual paths and event-study ATT for IFEct and MC side by side."

p_ife_ct <- plot(out_ife, type = "counterfactual",
                 main = "IFEct: observed vs counterfactual",
                 xlab = "Year", ylab = "Log teen employment")
p_ife_gap <- plot(out_ife, type = "gap",
                  main = "IFEct: event-study ATT",
                  xlab = "Event time", ylab = "ATT")
p_mc_ct <- plot(out_mc, type = "counterfactual",
                main = "MC: observed vs counterfactual",
                xlab = "Year", ylab = "Log teen employment")
p_mc_gap <- plot(out_mc, type = "gap",
                 main = "MC: event-study ATT",
                 xlab = "Event time", ylab = "ATT")

(p_ife_ct + p_ife_gap) / (p_mc_ct + p_mc_gap)
```

Both estimators reproduce the textbook story: a flat-ish pre-trend
that breaks downward at event time zero. The MC counterfactual is a
touch smoother than IFEct's, which is the regularisation doing its
job.

```{r}
#| label: tbl-att
#| tbl-cap: "Average post-treatment ATT under IFEct and MC, with bootstrap standard errors and 95% CIs. The two methods reach the same neighbourhood; CIs are wide because $T = 7$ and `nboots = 50`."
#| code-summary: "Code: Tabulate the average post-treatment ATT with bootstrap standard errors and CIs."
tibble(
  Method   = c("IFEct", "MC"),
  ATT      = c(out_ife$att.avg, out_mc$att.avg),
  `S.E.`   = c(out_ife$est.avg[1, "S.E."], out_mc$est.avg[1, "S.E."]),
  `CI lo`  = c(out_ife$est.avg[1, "CI.lower"], out_mc$est.avg[1, "CI.lower"]),
  `CI hi`  = c(out_ife$est.avg[1, "CI.upper"], out_mc$est.avg[1, "CI.upper"])
) |>
  gt_pretty(decimals = 4)
```

## IFEct and MC compared

A caveat first, then the comparison. On this 7-year window, CV
selects $r = 0$ for IFEct — meaning IFEct here operationally
collapses to two-way fixed effects: the factor term $\lambda_i' f_t$
drops out and the imputation reduces to $\widehat{Y_{it}(0)} =
\widehat{\alpha}_i + \widehat{\xi}_t$. This is not a bug, it is a
finding: with $T = 7$ and 3 pre-periods on the shorter cohort, CV
is conservative and refuses to credit a latent factor it cannot
validate out-of-sample. MC, by contrast, lands on a small but
non-zero penalty ($\eta = 0.0007$), so MC retains a regularised
low-rank deviation from TWFE.

So the empirical contrast on this panel is **not** between an
explicit factor model and a low-rank completion — it is between
TWFE-style imputation (IFEct at $r = 0$) and nuclear-norm
penalised imputation (MC at $\eta = 0.0007$). That two estimators
that disagree about the identifying assumption — *in principle* an
explicit linear factor model vs. an approximately low-rank $Y(0)$
matrix — still point to the same sign and the same order of
magnitude is meaningful: the qualitative conclusion is robust to
the relaxation of parallel trends that MC offers, exactly because
MC's relaxation is small here. The CV-selected hyperparameters are
not directly comparable (an integer rank vs. a continuous penalty),
but the counterfactual-path overlay, the ATT trajectory, and the
numeric ATT in @tbl-att are. When the two methods agree, both
relaxations of parallel trends are consistent with the same
conclusion, which is the strongest evidence a short panel design
can produce.

## Recap

::: {.callout-note appearance="simple"}
**The methods reconciled.** Two answers to the same question on the
2001-2007 minimum-wage panel:

- *IFEct*: factor model of $Y(0)$ fit on never-treated, imputed on
  treated, $r$ chosen by cross-validation. Here CV picks $r = 0$,
  so IFEct collapses to two-way fixed effects — a finding, not a
  bug.
- *Matrix Completion*: low-rank completion of the masked $Y(0)$
  matrix, nuclear-norm penalty $\eta$ chosen by cross-validation.

Both point downward; both sit in the same neighbourhood; neither is
uniquely correct. Factor-based estimators *relax* parallel trends
rather than replace it — the counterfactual-path plot is the visual
quality check, and a flat pre-treatment overlay is the signal you
want.
:::

## Common pitfall

Cranking $r$ up until the in-sample fit looks great. With a panel
this short, an IFEct with $r$ near $T/2$ will fit the pre-treatment
cells almost perfectly and produce an absurd counterfactual on the
treated cells, because it has fitted noise into the loadings.
Trust the CV-selected rank; the MC analogue is choosing $\eta$
too small. Both are over-fitting masquerading as identification.

## Further reading

The factor-model formulation traces to @bai2009panel and the IFEct /
generalised synthetic control reading is @xu2017generalized; the
matrix-completion view is @athey2021matrix. @liu2024practical pairs
with the `fect` package — its online companion at
<https://yiqingxu.org/packages/fect/04-ife-mc.html> is the
authoritative tutorial. Athey et al.'s own implementation lives in
the `MCPanel` package
(<https://github.com/susanathey/MCPanel>) for readers who want to
compare `fect::mc` against the canonical reference code. Chapter 10
zooms in on `gsynth`, the IFE estimator implemented in a
standalone package, on the same Callaway-Sant'Anna panel.

## Exercises

1. Re-fit `out_ife` with `r=0:3` (raising the search ceiling
   above the theoretical `min.T0` bound) and inspect `out_ife$MSPE`.
   Does CV ever pick a rank above 0 on this panel? What does that
   tell you about the data?
2. Add the two-covariate extension `lemp ~ D + lpop + lavg_pay` to
   both `fect()` calls, refit, and compare the ATT trajectory and
   the numeric headline ATT against @tbl-att. Does conditioning on
   log population and log average pay move the gap plot
   meaningfully?
3. Re-fit `out_mc` on a tighter window (`year >= 2002`) so cohort
   2004 has 2 pre-periods instead of 3. At what window does the MC
   counterfactual visibly break (sharp pre-trend in the gap plot)?
