---
title: "Interactive Fixed Effects and Matrix Completion"
---

## Learning objectives

1. Fit interactive-fixed-effects models and select the latent factor rank by cross-validation in `fect`. Letting CV pick the rank is what frees the reader from the unverifiable parallel-trends assumption while keeping the model from overfitting.
2. Overlay counterfactual and observed outcome paths to assess pre-treatment fit. Visual pre-treatment fit is the chapter's headline diagnostic and the only honest check that the factor model captured the relevant comovement.
3. Compare interactive-fixed-effects and matrix-completion estimators on the same panel. The two estimators rest on structurally different identifying assumptions, so agreement between them is one of the strongest robustness statements available for a staggered design.
4. Reason about the trade-off between pre-period depth and identifiable factor rank. Short panels cannot support many factors, and ignoring this ceiling fits noise into the loadings — producing a phantom counterfactual on the treated cells rather than a credible one.

## When parallel trends isn't enough

Chapter 8 squeezed every drop of usable signal out of the
Callaway-Sant'Anna minimum-wage panel under one identifying
assumption: **parallel trends**. Conditional or unconditional, with
or without never-treated counties, with or without anticipation
slack, with Rambachan-Roth bounds on top — the assumption is always
that, absent treatment, the trajectory of the treated counties would
have moved *in parallel* with the untreated.

What if it would not have? Counties select into raising the minimum
wage non-randomly. They are not a random subset of the country in
the time-varying dimension either: industry mix, demographics, and
local growth trends all shape both treatment timing and outcome
paths. Two counties can share an industry-mix-driven trajectory that
TWFE never absorbs — and so cannot net out.

This chapter relaxes parallel trends in a specific way. We model the
counterfactual outcome as a **factor structure**:

$$Y_{it}(0) \;=\; \alpha_i + \xi_t + \lambda_i^\top f_t + \varepsilon_{it}.$$

Each unit $i$ has a loading vector $\lambda_i$; each period a factor
vector $f_t$. Their interaction $\lambda_i^\top f_t$ encodes
time-varying unobserved heterogeneity that no county or year fixed
effect can absorb. We write the model without observable covariates
for simplicity, but `fect()` accepts a covariate matrix
$X_{it}$ — e.g. `lemp ~ D + lpop + lavg_pay` — and one of this
chapter's exercises walks through exactly that extension.

Two views of the same object give two estimators
[@liu2024practical]:

- **Interactive Fixed Effects (IFEct)** [@bai2009panel;
  @xu2017generalized; @liu2024practical] treats
  $(\alpha, \xi, \Lambda, F)$ as parameters to be estimated on
  control observations, then imputes $Y_{it}(0)$ on treated cells.
  We write the resulting estimator $\widehat{\tau}_{\text{IFE}}$.
- **Matrix Completion (MC)** [@athey2021matrix] reframes the same
  problem as filling missing entries of the implicit $Y(0)$ matrix:
  mask the treated cells, then complete via nuclear-norm-regularised
  SVD. We write the resulting estimator $\widehat{\tau}_{\text{MC}}$.

Yiqing Xu's `fect` [@liu2024practical] package implements both with
a common API and a common counterfactual-plot diagnostic. By
chapter's end we compare $\widehat{\tau}_{\text{IFE}}$ and
$\widehat{\tau}_{\text{MC}}$ on the same panel — two estimators, two
identifying assumptions, one substantive question. Chapter 10 then
does the deep dive on one specific member of this family —
generalized synthetic control via the `gsynth` package — using the
same panel.

## Setup and data

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set the ggplot theme."
library(tidyverse)
library(fect)
library(panelView)
library(patchwork)
source("R/table_helpers.R")

set.seed(42)

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8"),
      strip.text       = element_text(color = "#94a3b8"),
      legend.text      = element_text(color = "#94a3b8")
    )
)
```

We work from the same `cs_minwage.rds` panel as chapter 10, but with
one small adjustment: factor-based estimators need **at least as
many pre-treatment periods per cohort as the number of factors they
try to fit** — concretely, identifying $r$ factors on a treated unit
requires at least $r + 1$ pre-treatment periods for that unit.
Chapter 10's 2003-2007 working window leaves the 2004 cohort with
only one pre-period (year 2003). To keep that cohort in the IFE
sample we would need to set `min.T0 = 1` — the user-controlled
inclusion threshold that drops treated units with fewer
pre-periods — which in turn caps the identifiable rank at zero for
that cohort and collapses IFEct back to TWFE. Widening the window
to 2001-2007 gives 2004 three pre-periods (2001-2003), so we can
set `min.T0 = 2` and let CV search over $r \in \{0, 1, 2\}$.
Everything else (cohort filter, region drop, treatment indicator)
is identical to chapter 10.

```{r}
#| label: data-load
#| code-summary: "Code: Load the minimum-wage panel and restrict to the 2001-2007 working window."
mw_raw <- readRDS("data/cs_minwage.rds") |> as_tibble()

mw <- mw_raw |>
  filter(G %in% c(0, 2004, 2006, 2007), region != "1") |>
  filter(G != 2007, year >= 2001) |>
  mutate(D = as.integer(year >= G & G != 0))

dim(mw)
```

```{r}
#| label: tbl-cohorts
#| tbl-cap: "Cohorts in the 2001-2007 working panel. $G = 0$ is the never-treated control pool; the two treated cohorts have 3 and 5 pre-treatment years respectively."
#| code-summary: "Code: Summarize county counts and pre-treatment periods by cohort."
mw |>
  filter(year == 2001) |>
  count(G, name = "counties") |>
  mutate(`Pre-periods` = case_when(
    G == 0    ~ NA_integer_,
    G == 2004 ~ 3L,
    G == 2006 ~ 5L
  )) |>
  rename(`Cohort (G)` = G) |>
  gt_pretty()
```

The outcome `lemp` (log teen employment; see chapter 10 for full
data provenance) and the treatment indicator `D` (1 once a county's
state has raised its minimum wage above the federal floor, 0
otherwise) line up with chapter 10's definitions.

## Visualising the panel

`panelView` is the natural opener for any FECT workflow. It draws
units on the vertical axis and time on the horizontal, coloured by
treatment status, so the staggered-adoption structure of the panel
is immediate.

```{r}
#| label: fig-panelview
#| fig-cap: "Treatment status by county and year. The two horizontal bands of pink are the 2004 and 2006 cohorts; the wide grey band underneath is the never-treated $G = 0$ control pool. The visible step in the pink rows is the staggered adoption that breaks TWFE. All 1,745 counties are shown."
#| fig-width: 8
#| fig-height: 6
#| message: false
#| warning: false
#| code-summary: "Code: Visualize the staggered treatment status across counties with panelView."
panelview(data = as.data.frame(mw), formula = lemp ~ D,
          index = c("id", "year"),
          xlab = "Year", ylab = "County",
          main = "", legendOff = FALSE,
          display.all = TRUE,
          theme.bw = TRUE,
          background = "transparent")
```

## The factor model of counterfactuals

Both estimators target the same object — the counterfactual matrix
$Y(0)$ of what each county's log teen employment *would* have been
absent any minimum-wage increase. Where they differ is in how they
restrict that matrix.

**IFEct** imposes the explicit factor decomposition $Y_{it}(0) =
\alpha_i + \xi_t + \lambda_i^\top f_t + \varepsilon_{it}$, fits
$(\alpha, \xi, \Lambda, F)$ on the **control** observations only,
and uses those estimates to impute $Y_{it}(0)$ for treated
cells. The choice of $r$ — how many factors — is made by
cross-validation: hold out small blocks of control cells, refit,
score by MSPE on the held-out cells [@xu2017generalized;
@liu2024practical].

**MC** does not write the factor model down. It assumes only that
the matrix of $Y(0)$ outcomes is **approximately low rank**, then
completes it by minimising a Frobenius-norm fit penalty *plus* a
nuclear-norm penalty on the singular values. Nuclear-norm
regularisation is the convex relaxation of "low rank" the same way
$\ell_1$ is the convex relaxation of "sparse" [@athey2021matrix].
Concretely, MC solves
$$\min_{\widehat{Y(0)}} \; \|\,Y_{\mathrm{obs}} - \widehat{Y(0)}\,\|_F^2 \;+\; \eta \, \|\,\widehat{Y(0)}\,\|_*,$$
where $\|\cdot\|_*$ (the nuclear norm) is the sum of singular
values. As $\eta \to 0$ the fit interpolates the observed cells; as
$\eta \to \infty$ all singular values shrink to zero and
$\widehat{Y(0)}$ collapses to a constant. The penalty weight $\eta$
(written as `lambda` in the `fect` API and sometimes as
$\lambda_{\mathrm{MC}}$ in the matrix-completion literature; we use
$\eta$ in prose to avoid confusion with the unit-level loading
vector $\lambda_i$ from IFEct) plays the role of $r$ and is again
chosen by cross-validation.

The key practical implication: both methods bake in
**unit-specific time trends** automatically, without you having to
specify which units or which trend shapes. Parallel trends becomes
a special case (the $r = 0$ corner of IFEct, or the limit
$\eta \to \infty$ in MC) rather than an assumption.

## Estimating with FECT

::: {.callout-warning appearance="simple"}
**The short-panel caveat (read first).** With $T = 7$ and 3
pre-periods on the shorter cohort, the rank/penalty selected by
cross-validation is borderline-identifiable; we cap $r \le 2$
deliberately and the *effective* identification ceiling is
$r \le \min(\text{pre-periods}) - 1 = 1$. On a genuinely short
panel ($T \le 5$) IFE and MC become numerically delicate, and the
honest move is to *not* run them rather than to hand-pick a rank.
The `fect::simdata` panel ships a $T = 30$ example that shows both
methods working at full strength. Everything in the rest of this
chapter should be read with this caveat in mind.
:::

The two fits below cap the rank grid at $r \in \{0, 1, 2\}$. The
identification rule is that recovering $r$ factors on a treated
unit requires at least $r + 1$ pre-periods for that unit; with
`min.T0 = 2` the *effective* ceiling is $r = 1$. We let CV search
up to $r = 2$ only to verify it does not get fooled into climbing
past the ceiling. Any $r$ larger than that would be statistical
fantasy.

::: {.callout-note appearance="simple"}
**Learning-mode compute.** We set `nboots = 50` here so the chapter
renders quickly. The bootstrap dominates runtime and only affects
the *width* of confidence intervals — point estimates and the
CV-selected rank or penalty are unchanged. For research use, bump
`nboots` to 500-1000 so the bootstrap distribution of the ATT is
properly resolved.
:::

```{r}
#| label: fit-ife
#| message: false
#| warning: false
#| code-summary: "Code: Fit the interactive fixed effects estimator with fect and CV-selected rank."
out_ife <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "ife",
  force    = "two-way",
  CV       = TRUE,
  r        = 0:2,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)
```

```{r}
#| label: fit-mc
#| message: false
#| warning: false
#| code-summary: "Code: Fit the matrix-completion estimator with fect and CV-selected nuclear-norm penalty."
out_mc <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "mc",
  force    = "two-way",
  CV       = TRUE,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)
```

```{r}
#| label: tbl-cv
#| tbl-cap: "Hyperparameters selected by cross-validation. IFEct picks the number of latent factors $r$; MC picks the nuclear-norm penalty weight $\\eta$ (returned by `fect` as `lambda.cv`)."
#| code-summary: "Code: Report the CV-selected rank for IFEct and nuclear-norm penalty for MC."
tibble(
  Method            = c("IFEct", "MC"),
  `CV-selected`     = c(paste0("r = ", unname(out_ife$r.cv)),
                        sprintf("eta = %.4f", out_mc$lambda.cv))
) |>
  gt_pretty()
```

## Counterfactual paths and the ATT trajectory

The signature FECT figure overlays the observed average outcome on
treated units against the model-implied $Y(0)$, then plots the gap
— the ATT path — by event time. A credible factor model should
track the observed $Y$ closely before treatment and only diverge
after.

```{r}
#| label: fig-ife-mc
#| fig-cap: "Counterfactual trajectories and event-study ATT. Top row: IFEct. Bottom row: MC. Left panels overlay observed average outcome on treated units (solid) with model-implied $Y(0)$ (dashed). Right panels show the implied ATT by event time relative to a county's adoption year, with bootstrap 95% CIs."
#| fig-width: 9
#| fig-height: 7
#| code-summary: "Code: Plot counterfactual paths and event-study ATT for IFEct and MC side by side."

p_ife_ct <- plot(out_ife, type = "counterfactual",
                 main = "IFEct: observed vs counterfactual",
                 xlab = "Year", ylab = "Log teen employment")
p_ife_gap <- plot(out_ife, type = "gap",
                  main = "IFEct: event-study ATT",
                  xlab = "Event time", ylab = "ATT")
p_mc_ct <- plot(out_mc, type = "counterfactual",
                main = "MC: observed vs counterfactual",
                xlab = "Year", ylab = "Log teen employment")
p_mc_gap <- plot(out_mc, type = "gap",
                 main = "MC: event-study ATT",
                 xlab = "Event time", ylab = "ATT")

(p_ife_ct + p_ife_gap) / (p_mc_ct + p_mc_gap)
```

Both estimators reproduce the textbook story: a flat-ish pre-trend
that breaks downward at event time zero. The MC counterfactual is a
touch smoother than IFEct's, which is the regularisation doing its
job.

```{r}
#| label: tbl-att
#| tbl-cap: "Average post-treatment ATT under IFEct and MC, with bootstrap standard errors and 95% CIs. The two methods reach the same neighbourhood; CIs are wide because $T = 7$ and `nboots = 50`."
#| code-summary: "Code: Tabulate the average post-treatment ATT with bootstrap standard errors and CIs."
tibble(
  Method   = c("IFEct", "MC"),
  ATT      = c(out_ife$att.avg, out_mc$att.avg),
  `S.E.`   = c(out_ife$est.avg[1, "S.E."], out_mc$est.avg[1, "S.E."]),
  `CI lo`  = c(out_ife$est.avg[1, "CI.lower"], out_mc$est.avg[1, "CI.lower"]),
  `CI hi`  = c(out_ife$est.avg[1, "CI.upper"], out_mc$est.avg[1, "CI.upper"])
) |>
  gt_pretty(decimals = 4)
```

## IFEct and MC compared

A caveat first, then the comparison. On this 7-year window, CV
selects $r = `r unname(out_ife$r.cv)`$ for IFEct — meaning IFEct
here operationally collapses to two-way fixed effects: the factor
term $\lambda_i^\top f_t$ drops out and the imputation reduces to
$\widehat{Y_{it}(0)} = \widehat{\alpha}_i + \widehat{\xi}_t$. This
is not a bug, it is a finding: with $T = 7$ and 3 pre-periods on
the shorter cohort, CV is conservative and refuses to credit a
latent factor it cannot validate out-of-sample. MC, by contrast,
lands on a small but non-zero penalty
($\eta \approx `r sprintf("%.4f", out_mc$lambda.cv)`$), so MC
retains a regularised low-rank deviation from TWFE.

So the empirical contrast on this panel is **not** between an
explicit factor model and a low-rank completion — it is between
TWFE-style imputation ($\widehat{\tau}_{\text{IFE}}$ at $r =
`r unname(out_ife$r.cv)`$) and nuclear-norm penalised imputation
($\widehat{\tau}_{\text{MC}}$ at $\eta \approx
`r sprintf("%.4f", out_mc$lambda.cv)`$). That two estimators that
disagree about the identifying assumption — *in principle* an
explicit linear factor model vs. an approximately low-rank $Y(0)$
matrix — still point to the same sign and the same order of
magnitude is meaningful: the qualitative conclusion is robust to
the relaxation of parallel trends that MC offers, exactly because
MC's relaxation is small here. The CV-selected hyperparameters are
not directly comparable (an integer rank vs. a continuous penalty),
but the counterfactual-path overlay, the ATT trajectory, and the
numeric ATT in @tbl-att are. When the two methods agree, both
relaxations of parallel trends are consistent with the same
conclusion, which is the strongest evidence a short panel design
can produce.

## Recap

::: {.callout-note appearance="simple"}
**The methods reconciled.** Two answers to the same question on the
2001-2007 minimum-wage panel:

- $\widehat{\tau}_{\text{IFE}}$: factor model of $Y(0)$ fit on
  never-treated, imputed on treated, $r$ chosen by cross-validation.
  Here CV picks $r = `r unname(out_ife$r.cv)`$, so IFEct collapses
  to two-way fixed effects — a finding, not a bug.
- $\widehat{\tau}_{\text{MC}}$: low-rank completion of the masked
  $Y(0)$ matrix, nuclear-norm penalty $\eta$ chosen by
  cross-validation
  ($\eta \approx `r sprintf("%.4f", out_mc$lambda.cv)`$).

Both point downward; both sit in the same neighbourhood; neither is
uniquely correct. Factor-based estimators *relax* parallel trends
rather than replace it — the counterfactual-path plot is the visual
quality check, and a flat pre-treatment overlay is the signal you
want.
:::

## Common pitfall

Cranking $r$ up until the in-sample fit looks great. With a panel
this short, an IFEct with $r$ near $T/2$ will fit the pre-treatment
cells almost perfectly and produce an absurd counterfactual on the
treated cells, because it has fitted noise into the loadings.
Trust the CV-selected rank; the MC analogue is choosing $\eta$
too small. Both are over-fitting masquerading as identification.

## Key takeaways

**Methods:**

- *Interactive fixed effects* (IFEct) writes the untreated potential outcome as $Y_{it}(0) = \alpha_i + \xi_t + \lambda_i^\top f_t + \varepsilon_{it}$, where $\lambda_i^\top f_t$ is a low-rank product of $r$ latent factors $f_t$ and unit loadings $\lambda_i$ that absorbs time-varying unobserved heterogeneity no fixed effect can net out. Fitting $(\alpha, \xi, \Lambda, F)$ on never-treated cells and imputing $\widehat{Y_{it}(0)}$ on treated cells turns the ATT into an outcome-imputation problem and yields $\widehat{\tau}_{\text{IFE}}$; the rank $r$ is chosen by holding out blocks of control cells and minimising out-of-sample MSPE.
- *Matrix completion* (MC) drops the explicit factor decomposition and only assumes the implicit $Y(0)$ matrix is approximately low-rank, then fills its masked treated entries by minimising $\|Y_{\mathrm{obs}} - \widehat{Y(0)}\|_F^2 + \eta\,\|\widehat{Y(0)}\|_*$ to yield $\widehat{\tau}_{\text{MC}}$. The *nuclear norm* $\|\cdot\|_*$ — the sum of singular values — is the convex relaxation of "low rank" in the same way that the $\ell_1$ norm relaxes "sparse", and the penalty weight $\eta$ (returned by `fect` as `lambda.cv`) is again chosen by cross-validation.

**Lessons:**

- Both estimators *relax* parallel trends rather than replacing it: TWFE is the $r = 0$ corner of IFEct and the $\eta \to \infty$ limit of MC, so any non-zero CV pick is the data telling you parallel trends was insufficient on its own.
- The signature FECT diagnostic is the **counterfactual-path overlay**: plot the observed average outcome on treated units against the model-implied $Y(0)$, and read off the event-study ATT as the gap. A flat pre-treatment overlay is the visual quality check — the analogue of a pre-trend test for parallel-trends DiD.
- On the 2001-2007 minimum-wage panel CV picks $r = `r unname(out_ife$r.cv)`$ for IFEct (it collapses to two-way fixed effects) but a small non-zero $\eta \approx `r sprintf("%.4f", out_mc$lambda.cv)`$ for MC; the two estimators nevertheless land in the same neighbourhood, and that agreement is the strongest robustness check a short panel can produce — qualitatively the same answer under two different relaxations of parallel trends.

**Caveats:**

- Factor models are data-hungry: identifying $r$ factors on a treated unit requires at least $r + 1$ pre-treatment periods, so short panels ($T \le 5$) cap the recoverable rank near zero and IFEct degenerates to TWFE. The chapter widens the chapter-8 window to 2001-2007 precisely to give the 2004 cohort three pre-periods; on genuinely short panels the honest move is to not run IFEct or MC at all rather than hand-pick a rank.
- CV-selected hyperparameters are not directly comparable across the two methods — an integer rank $r$ and a continuous penalty $\eta$ live on different scales — so cross-method validation has to run through the *counterfactual paths*, the *event-study gap plot*, and the *numeric ATT*, not through the tuning parameters themselves.
- The reported confidence intervals come from a bootstrap whose width depends on `nboots`; the chapter uses `nboots = 50` for fast rendering, which is fine for point estimates and CV selection but under-resolves the tails — research use should bump it to 500-1000.

## Further reading

The factor-model formulation traces to @bai2009panel and the IFEct /
generalised synthetic control reading is @xu2017generalized; the
matrix-completion view is @athey2021matrix. @liu2024practical pairs
with the `fect` package — its online companion at
<https://yiqingxu.org/packages/fect/04-ife-mc.html> is the
authoritative tutorial. Athey et al.'s own implementation lives in
the `MCPanel` package
(<https://github.com/susanathey/MCPanel>) for readers who want to
compare `fect::mc` against the canonical reference code. Chapter 10
zooms in on `gsynth`, the IFE estimator implemented in a
standalone package, on the same Callaway-Sant'Anna panel.

## Exercises

These exercises probe the chapter's central tensions: CV's reluctance to credit a factor on a short panel, the MC penalty's role, the covariate extension, and the placebo check that should be passed before the headline ATT is trusted. All reuse `mw`, `out_ife`, and `out_mc` from the setup chunks above.

### Exercise 1: Force IFEct at $r = 1$

CV picked $r = `r unname(out_ife$r.cv)`$, collapsing IFEct to TWFE. Refit `fect()` with `CV = FALSE` and $r = 1$ (override the CV decision and add one explicit factor). Report the ATT and compare to the chapter's CV-selected baseline. Does the on-impact event-time ATT shift?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch11-01
#| message: false
#| warning: false
out_ife_r1 <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "ife",
  force    = "two-way",
  CV       = FALSE,
  r        = 1,
  min.T0   = 2,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)

tibble(spec = c(sprintf("IFEct r = %d (CV-selected, chapter)", out_ife$r.cv),
                "IFEct r = 1 (forced)"),
       att  = c(out_ife$att.avg, out_ife_r1$att.avg)) |>
  gt_pretty(decimals = 4)
```

Forcing $r = 1$ moves the ATT modestly: one explicit factor absorbs some of the within-county trend that the $r = 0$ specification had to push into the gap. The CV-selected $r = 0$ is the more conservative call on this short panel — it refuses to credit a factor it cannot validate out-of-sample. The fact that $r = 1$ does not radically change the headline is reassuring; if it did, the chapter's caveat about short-panel fragility would be a five-alarm warning rather than a flagged limitation.
:::

### Exercise 2: MC penalty hand-tune

Refit `fect()` for MC with `CV = FALSE` at two hand-chosen penalty values — $\eta = 0.05$ (looser, more shrinkage) and $\eta = 0.0001$ (tighter, near-interpolation). Report the ATTs and discuss which lands closer to the CV-selected $\eta \approx `r sprintf("%.4f", out_mc$lambda.cv)`$.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch11-02
#| message: false
#| warning: false
out_mc_loose <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "mc",
  force    = "two-way",
  CV       = FALSE,
  lambda   = 0.05,
  min.T0   = 2,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)

out_mc_tight <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "mc",
  force    = "two-way",
  CV       = FALSE,
  lambda   = 0.0001,
  min.T0   = 2,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)

tibble(spec = c("MC eta = 0.05 (loose)",
                sprintf("MC eta = %.4f (CV, chapter)", out_mc$lambda.cv),
                "MC eta = 0.0001 (tight)"),
       att  = c(out_mc_loose$att.avg, out_mc$att.avg, out_mc_tight$att.avg)) |>
  gt_pretty(decimals = 4)
```

The loose penalty pushes the MC counterfactual toward a constant — the ATT moves toward zero because the imputed $Y(0)$ stops tracking county-specific trends. The tight penalty lets MC interpolate the observed cells and produces an ATT close to the CV-selected value. This is the bias-variance trade-off that CV is trying to navigate: too much shrinkage erases real signal; too little fits pre-period noise.
:::

### Exercise 3: Two-covariate conditional ATT

Refit both IFEct and MC with `lemp ~ D + lpop + lavg_pay` (adding log county population and log average pay as covariates). Compare the conditional ATTs to the chapter's uncovariate-adjusted baselines. Does the conditional adjustment move the gap meaningfully?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch11-03
#| message: false
#| warning: false
out_ife_x <- fect(
  lemp ~ D + lpop + lavg_pay, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "ife",
  force    = "two-way",
  CV       = TRUE,
  r        = 0:2,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)

out_mc_x <- fect(
  lemp ~ D + lpop + lavg_pay, data = as.data.frame(mw),
  index    = c("id", "year"),
  method   = "mc",
  force    = "two-way",
  CV       = TRUE,
  min.T0   = 2,
  cv.nobs  = 2,
  cv.donut = 0,
  se       = TRUE,
  nboots   = 50,
  parallel = TRUE,
  seed     = 42
)

tibble(spec = c("IFEct (chapter, no covariates)",
                "IFEct + lpop + lavg_pay",
                "MC (chapter, no covariates)",
                "MC + lpop + lavg_pay"),
       att  = c(out_ife$att.avg, out_ife_x$att.avg,
                out_mc$att.avg,  out_mc_x$att.avg)) |>
  gt_pretty(decimals = 4)
```

The covariate-augmented ATTs sit close to the unadjusted ones — the factor and matrix-completion structures were already absorbing most of the time-varying heterogeneity that `lpop` and `lavg_pay` could supply. Covariates in a factor model behave as *partial* identification aids: they help when the latent factors do not span the relevant time-varying confounders, and they are redundant when the factors already do.
:::

### Exercise 4: Event-time ATT side-by-side

The chapter shows the per-event-time ATT as plots only. Build a numeric side-by-side table of $\widehat{ATT}(e)$ for $e \in \{-4, \dots, +3\}$ under IFEct and MC, with bootstrap 95% CIs. Where do the two estimators most disagree?

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch11-04
extract_eventtime <- function(out, label) {
  m <- out$est.att
  tibble(method = label,
         event_time = as.integer(rownames(m)),
         att        = m[, "ATT"],
         lower      = m[, "CI.lower"],
         upper      = m[, "CI.upper"])
}

bind_rows(
  extract_eventtime(out_ife, "IFEct"),
  extract_eventtime(out_mc,  "MC")
) |>
  filter(event_time >= -4, event_time <= 3) |>
  arrange(event_time, method) |>
  gt_pretty(decimals = 4)
```

The two estimators agree closely in event-time tails where the data is sparse and CIs are wide. Disagreement (if any) concentrates at the on-impact event time, where IFEct (here collapsed to TWFE at $r = `r unname(out_ife$r.cv)`$) lacks the MC penalty's ability to smooth the imputed $Y(0)$ across nearby cells. When the two methods agree across event time, both relaxations of parallel trends are consistent with the same trajectory — the strongest robustness signal a short panel can produce.
:::

### Exercise 5 (stretch): In-time placebo test via `fect`

`fect()` has a built-in placebo facility: set `placeboTest = TRUE` and `placebo.period = c(-2, -1)` to *mask* the two periods immediately before treatment and ask whether the model assigns them a non-zero "effect." A well-calibrated factor or matrix-completion model should place near-zero effects on these placebo periods. Run the placebo test for IFEct and report the placebo $p$-value.

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch11-05
#| message: false
#| warning: false
out_ife_placebo <- fect(
  lemp ~ D, data = as.data.frame(mw),
  index          = c("id", "year"),
  method         = "ife",
  force          = "two-way",
  CV             = FALSE,
  r              = out_ife$r.cv,
  min.T0         = 2,
  se             = TRUE,
  nboots         = 50,
  placeboTest    = TRUE,
  placebo.period = c(-2, -1),
  parallel       = TRUE,
  seed           = 42
)

list(
  placebo_att      = out_ife_placebo$att.placebo,
  placebo_p_value  = out_ife_placebo$est.placebo
)
```

The placebo ATT is small in absolute value and the placebo $p$-value sits well above 0.05 — IFEct is *not* assigning a sizeable "effect" to the two periods immediately before treatment. A *failed* placebo (a placebo ATT comparable in magnitude to the real on-impact effect, with $p < 0.05$) would suggest the model's pre-treatment fit is borrowing structure from the post-treatment cells through the factor decomposition, and would force a reconsideration of the headline. The clean placebo here is the strongest single-number reassurance you can get out of a short-panel IFE fit.
:::
