---
title: "{{CHAPTER TITLE}}"
---

<!--
  CHAPTER SKELETON for Comparative Causal Metrics.
  Copy this to NN-your-slug.qmd and fill every {{PLACEHOLDER}}.
  Read references/house-style.md first. Do NOT reorder sections.
  Replace ALL <!-- comments --> with real content or delete them.
  Every number in prose must come from inline R: `r sprintf("%.1f", obj)`.
-->

## Learning objectives

<!-- 3-4 numbered objectives. Each is TWO sentences: what the reader does,
     then why it matters / what it buys them. -->

1. {{Do X with `pkg::fn()`.}} {{Why this is the technical core / what it buys.}}
2. {{Do Y.}} {{Why.}}
3. {{Do Z.}} {{Why.}}

<!-- PART-SEAM ONLY: if this chapter opens Part II, paste the callout-note from
     house-style.md §9 here and delete this comment. Otherwise delete this block. -->

## {{The <method> idea}}

<!-- Opening CALLBACK to the previous chapter by name and number, then what THIS
     chapter changes. Keep sentences short. -->

{{Chapter K (Previous Method) did ... . This chapter ... .}}

This chapter fits {{N}} variants on the same {{dataset}} data:

1. {{Variant 1 — one line.}}
2. {{Variant 2 — one line.}}

<!-- State the through-line: each variant builds the counterfactual
     $\widehat{Y_{1t}(0)}$ a different way, then ATT = observed - imputed. -->

**Identification.** {{One paragraph: the single assumption under which this
method recovers the ATT. Name it. Say everything in the chapter is conditional
on it.}}

## Setup and data

**Packages.** {{One short paragraph naming each package and what it does in this
chapter — tidyverse for wrangling/plots, the method package, any SE/diagnostic
helper, and the in-repo `R/table_helpers.R`.}}

```{r}
#| label: setup
#| message: false
#| warning: false
#| code-summary: "Code: Load packages, source table helpers, and set transparent ggplot theme."
library(tidyverse)
# {{method-specific packages}}
source("R/table_helpers.R")

knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.background  = element_rect(fill = "transparent", color = NA),
      panel.background = element_rect(fill = "transparent", color = NA),
      panel.grid.major = element_line(color = "#94a3b8", linewidth = 0.25),
      panel.grid.minor = element_line(color = "#94a3b8", linewidth = 0.15),
      text             = element_text(color = "#94a3b8"),
      axis.text        = element_text(color = "#94a3b8")
    )
)
```

**Dataset.** {{Shape (rows × cols), treated unit(s), cutoff t*, outcome and its
units, and how this chapter subsets/transforms it.}}

```{r}
#| label: data-load
#| code-summary: "Code: {{Load the dataset and build the chapter-specific subset.}}"
# {{Part I:}} prop99 <- read_rds("data/proposition99.rds") |> as_tibble()
# {{Part II:}} panel <- read_rds("data/cs_minwage.rds")
# {{build the subset/transform used below}}
```

<!-- ============ ONE SECTION PER METHOD VARIANT (repeat this block) ============ -->

## {{Method variant 1}}

**The idea.** {{Plain-English intuition in 1-2 short sentences.}}

**The equation.** {{Lead-in sentence, then display math in book notation.}}

$$\widehat{Y_{1t}(0)} = {{...}}, \qquad \widehat{\tau}_{\text{{{method}}}} = \frac{1}{T_{\text{post}}} \sum_{t > t^*} \Big[ Y_{1t} - \widehat{Y_{1t}(0)} \Big].$$

In words: {{restate the equation in plain English; define every symbol.}}

```{r}
#| label: tbl-{{slug}}
#| tbl-cap: "{{Caption that fully describes the table.}}"
#| code-summary: "Code: {{imperative description}}."
# {{fit the model; render with ms_pretty() for models or gt_pretty() for frames}}
```

```{r}
#| label: {{slug}}-stats
#| echo: false
# {{precompute scalars the prose references inline, e.g.:}}
# est_1 <- mean(post$y - cf)
```

**Reading the output.** {{Interpret the numbers using inline R, e.g. the estimate
is $\approx `r sprintf("%.1f", est_1)`$. Compare to neighbouring methods. State
the window/estimand so it's comparable.}}

```{r}
#| label: fig-{{slug}}
#| fig-cap: "{{Caption.}}"
#| fig-width: 8
#| fig-height: 5
#| code-summary: "Code: {{imperative description}}."
# {{plot observed (#d97757) vs counterfactual (#6a9bcc dashed), dotted cutoff vline}}
```

{{One paragraph reading the figure.}}

**Common pitfall.** {{The single most likely way to misuse or misread this
variant.}}

<!-- ============ END method-variant block; repeat as needed ============ -->

## What the {{...}} estimates tell us

<!-- Cross-variant discussion. Reconcile the numbers (inline R). Name the lesson.
     Then CLOSE with a hand-off to the next chapter by name and number
     (or, for the last method chapter, to the planned comparison chapter). -->

**{{The disagreement / Where this leaves us}}.** {{...}}

**A note on standard errors.** {{If relevant: which SE/uncertainty story applies
and why.}}

**Where this leaves us.** {{Hand-off: name the next chapter and what it adds.}}

**Recap.** {{One-sentence recap with the headline numbers inline.}}

## Key takeaways

**Methods:**

- {{Estimator(s) + R functions used.}}

**Lessons:**

- {{Conceptual morals; headline ATT with inline numbers; cross-method comparison.}}

**Caveats:**

- {{Assumptions and failure modes; when not to trust the estimate.}}

## Further reading

- @{{key}} — *{{textbook|tutorial|R package|practical guide|review|paper}}* — {{one sentence.}}
- @{{key}} — *{{type}}* — {{one sentence.}}
- @{{key}} — *{{type}}* — {{one sentence.}}

## Exercises

<!-- Short intro: exercises reuse the setup objects; nothing to re-load. -->

The exercises below {{probe the chapter's lesson}}. All reuse the objects from the
setup chunks; nothing needs to be re-loaded.

### Exercise 1: {{Title}}

{{Prompt.}}

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch{{NN}}-01
# {{solution; tables via gt_pretty()}}
```

{{One short paragraph interpreting the result and tying it to the lesson.}}
:::

### Exercise 2: {{Title}}

{{Prompt.}}

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch{{NN}}-02
```

{{Interpretation.}}
:::

### Exercise 3: {{Title}}

{{Prompt.}}

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch{{NN}}-03
```

{{Interpretation.}}
:::

### Exercise 4: {{Title}}

{{Prompt.}}

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch{{NN}}-04
```

{{Interpretation.}}
:::

### Exercise 5 (stretch): {{Title}}

{{Prompt — a harder, open-ended extension, often a placebo on an untreated unit.}}

::: {.callout-tip collapse="true"}
## Solution

```{r}
#| label: ex-ch{{NN}}-05
```

{{Interpretation.}}
:::
